### 2019-01-23 thesis.tex
Start latex thesis file

\documentclass[titlepage]{article}
\usepackage[style=nature]{biblatex}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{hyperref}

\begin{document}

\title{Removing nonlinear batch effects using adversarial deep neural networks}
\author[1]{Jonathan B. Dayton}
\author[1,2]{Stephen R. Piccolo}
\affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
\affil[2]{Department of Biomedical Informatics, University of Utah, Salt Lake City, UT 84108 USA}

\maketitle

\begin{abstract}
	The abstract text goes here.
\end{abstract}

\section{Background}
Here is the text of your introduction.

\subsection{Subsection Heading Here}
Write your subsection text here.

\begin{figure}
	\centering
	\includegraphics[width=4.5in]{figures/network}
	\caption{Network architecture}
	\label{simulationfigure}
\end{figure}

\section{Implementation}

\emph{\href{https://bmcbioinformatics.biomedcentral.com/submission-guidelines/preparing-your-manuscript/software-article}{Oxford Bioinformatics' guidelines} say that you should have an "Implementation" section instead of Materials and Methods.}

\begin{equation}
	\label{simple_equation}
	\alpha = \sqrt{ \beta }
\end{equation}

\section{Results}

\section{Discussion}

\section{Conclusions}

\end{document}

### 2019-01-24 thesis.tex
Add outline to .tex file

@@
 \author[1]{Jonathan B. Dayton}
-\author[1,2]{Stephen R. Piccolo}
+\author[1]{Stephen R. Piccolo}
 \affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
-\affil[2]{Department of Biomedical Informatics, University of Utah, Salt Lake City, UT 84108 USA}
 
@@
 \section{Background}
-Here is the text of your introduction.
+
+\emph{[Explanation of what's been done in batch adjusting so far].}
+
+Many tools have been developed to account and adjust for batch effects in expression data.
+However, most of these tools only adjust for linear batch effects without accounting for nonlinear interactions between variables.
+Because of the large number of features in expression data and the nature of these features (i.e.
+transcripts and proteins that interact biologically), we anticipate that these nonlinear effects may still be present even after adjustment with popular tools.
+A few notable examples use deep neural networks to correct for batch effects, but these are optimized for use in large, balanced, dual-batched datasets and are therefore unusable for many real-world examples.
+
+\textbf{Hypothesis:}
+Using an adversarial neural network can correct for batch effects more completely than previous tools do. % clarify here what "more completely" means
 
@@
 
+\subsection{Network Structure}
+
+\subsubsection{Autoencoder}
+
+\subsubsection{Discriminator}
+
+\subsubsection{Loss functions}
+
+\subsubsection{Training}
+
+\subsection{Datasets}
+
+\subsubsection{Data Format}
+
+\subsubsection{MNIST}
+
+\paragraph{Synthetic Batch Effects}
+
+\subsubsection{Bladderbatch}
+
+\subsubsection{GSE37199}
+
+\subsubsection{TCGA Pan-cancer Data}
+
+\subsection{Statistics and Metrics}
+
+\subsubsection{Mean squared error}
+
+\begin{equation}
+	\label{mse}
+	MSE = \frac{1}{n}\sum_{i=1}^n{(\hat{x}_i - x_i)^2}
+\end{equation}
+
+\subsubsection{Maximum mean discrepancy}
+
+Sample MMD:
+
+\begin{equation}
+	\label{mmd}
+	MMD = \frac{1}{n^2}\sum_{i=1}^n{\sum_{j=1}^n{k(x_i, x_j)}} - \frac{2}{nm}\sum_{i=1}^n{\sum_{j=1}^m{k(x_i, y_j)}} + \frac{1}{m^2}\sum_{i=1}^m{\sum_{j=1}^m{k(y_i, y_j)}}
+\end{equation}
+
+Where $k(x, y)$ is the Gaussian kernel between $x$ and $y$ as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel}.
+
+\subsubsection{Classification accuracy}
+
 \begin{equation}

### 2019-01-24 thesis.tex
Get references working

@@
 \documentclass[titlepage]{article}
-\usepackage[style=nature]{biblatex}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
+\usepackage[square,numbers]{natbib}
+\bibliographystyle{unsrtnat}
 
@@
 \affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
+\date{}
 
@@
 transcripts and proteins that interact biologically), we anticipate that these nonlinear effects may still be present even after adjustment with popular tools.
-A few notable examples use deep neural networks to correct for batch effects, but these are optimized for use in large, balanced, dual-batched datasets and are therefore unusable for many real-world examples.
+A few notable examples \cite{shaham_removal_2017,shaham_batch_2018} use deep neural networks to correct for batch effects, but these are optimized for use in large, balanced, dual-batched datasets and are therefore unusable for many real-world examples.
 
@@
 
+\bibliography{references}
+
 \end{document}

### 2019-01-29 thesis.tex
Initial implementation section

@@
 
-\subsection{Subsection Heading Here}
-Write your subsection text here.
+\section{Implementation}
+
+\emph{\href{https://bmcbioinformatics.biomedcentral.com/submission-guidelines/preparing-your-manuscript/software-article}{Oxford Bioinformatics' guidelines} say that you should have an "Implementation" section instead of Materials and Methods.}
+
+\subsection{Network Structure}
+
+We used an adversarial autoencoder network to model and remove the batch effects.
+This network was structured in two parts: an autoencoder to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder.
+By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over training to output the expression data with all signs of batch effects removed.
+The neural network was implemented in TensorFlow under Python 3.6.
+All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final activations in the autoencoder and the discriminator, which used the sigmoid for their final activations.
+We also tried replacing the ReLU activations with SELUs \cite{klambauer_self-normalizing_2017} but saw little if any improvement and therefore returned to using ReLUs.
+
+We toyed with the idea of using convolutional layers \cite{krizhevsky_imagenet_2012-1} in the network because of their incredible success in the realm of image recognition.
+We thought at first that a 20,000-dimensional vector of expression values was quite similar to a 200x100 image and that similar techniques could be used for both, which was encouraging because convolutional filters are quite efficient for processing images.
+However, we were unable to devise a representation or arrangement of gene expression data points in which convolutional filters would make sense.
+Although one could easily wrap a gene expression vector into two dimensions and arrange the genes by location in the genome, each data point fundamentally represents a completely different gene product.
+Contrast this with image data, where each pixel represents the same data type (i.e.
+relative light intensities of some wavelength); if each pixel value was shifted to the left by one pixel, the image would still represent the same idea.
+However, if each expression value were to be shifted one position to the left, it would in essence be like mistaking each gene product for another unrelated gene product.
+Therefore, passing convolutional filters over a gene expression array, 1D or 2D, would not make sense since the relationship between a pair of consecutive genes is not necessarily analogous to the relationship between another pair of consecutive genes.
+Because we couldn't justify any arrangement where spatial relationships between genes were consistently meaningful, we chose to only use fully connected layers in Confounded rather than convolutional layers.
+We welcome any future intellectual advances that may allow convolutional networks and other image processing techniques to be applied to expression data.
 
@@
 
-\section{Implementation}
+\subsubsection{Autoencoder}
 
-\emph{\href{https://bmcbioinformatics.biomedcentral.com/submission-guidelines/preparing-your-manuscript/software-article}{Oxford Bioinformatics' guidelines} say that you should have an "Implementation" section instead of Materials and Methods.}
-
-\subsection{Network Structure}
-
-\subsubsection{Autoencoder}
+We initially implemented a basic fully connected autoencoder with two layers for encoding and two layers for decoding and with a smaller (about $\frac{1}{5}$ the size of the input) code layer.
+In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
+We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
+We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
+% TODO: implement and test the variational ae
 
@@
 
+The discriminator is given the output of the autoencoder and was trained to determine the original batch of that output.
+It consists of a variable number of fully connected layers of size 500 and a one-hot encoded output for determining batch.
+We noticed that the discriminator was overfitting on the data.
+Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
+In order to combat this overfitting, we also added 50\%-probability dropout to each layer of the discriminator.
+After adding dropout, the network appeared to no longer be overfitting.
+
 \subsubsection{Loss functions}
+
+Three loss functions were used to train the network.
+First, the autoencoder was trained on mean squared error \ref{mse} ($MSE_A$) between the input and output layers.
+Second, the discriminator was trained on mean squared error ($MSE_D$) between its output and a one-hot encoding of the samples' batch labels.
+Finally, the autoencoder was also trained on a combination of the two previous losses,
+
+\begin{equation}
+	\label{dual_loss}
+	Loss = MSE_A - \lambda{}(MSE_D)
+\end{equation}
+
+Where the $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
 
@@
 
+In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of either 0.001 or 0.0001.
+We typically trained for 10,000 iterations on mini-batches of size 100.
+This training typically took roughly 30 minutes to complete, including time spent to load the input and to save the output.
+
 \subsection{Datasets}
+
+In order to test both theoretical and practical differences between Confounded and previous methods, we compared them for a variety of different datasets.
+Working with each new dataset enlightened us of a new edge case that hadn't yet been addressed in our code.
+By applying to our methods to each of the following datasets, our code became much more robust than if we had only developed and tested on one or two toy datasets.
 
@@
 
+Confounded takes input data in the tidy data format \cite{wickham_tidy_2014-1}, namely:
+
+\begin{itemize}
+	\item Each row represents a different sample.
+	\item Each column represents a different variable (either a clinical variable or a gene's expression).
+\end{itemize}
+
+Confounded parses through files in this format to determine which columns are discrete (integer or string) and which are continuous (floating point).
+Each continuous column is interpreted to be a gene expression column and is adjusted for by the network according to the batch column.
+This is one limitation to our software.
+In a future iteration, it would be more robust to allow the user to specify continuous columns that should not be interpreted as gene expression data (e.g.
+ages in years with decimal precision representing fractions of years).
+However for the datasets we tested, this didn't prove to be a problem.
+
 \subsubsection{MNIST}
+
+The MNIST digits dataset \cite{lecun_mnist_nodate} is a database of images of handwritten digits that are size-normalized and centered.
+It contains 60,000 training images and 10,000 test images.
+In order to use this dataset, we flattened each 28 by 28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit information.
+We limited our dataset to only the 10,000 test images.
+
+By using MNIST, we were able to visually assess how much true signal (in this case, the shape and digit of each handwritten digit image) is preserved after batch adjustment.
+Although convolutional layers would typically be used when working with images such as the MNIST dataset, we chose to use only fully connected layers even for this image dataset.
+In this way, we show that the autoencoder is still able to find and represent spatial relationships without explicitly defining spatial relationships in the model while testing the same network we use on expression data, where no spatial relationships are inherent.
 
@@
 
+Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear batch effects.
+To do so, we wrote a Python script that would take the MNIST data in, apply some nonlinear effect, and output the adjusted data.
+We applied nonlinear effects by iteratively realizing vectors of normal values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
+We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
+The same normal vectors were applied to each image in a batch.
+Finally, random noise was added to each image in order to prevent images in a batch from being too similar to each other.
+
 \subsubsection{Bladderbatch}
+
+The bladderbatch dataset is a microarray transcriptome expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
+It has been made available in an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package to illustrate how to batch-adjust using ComBat \cite{leek_sva_2017}.
+It contains expression values for 57 samples of patients with and without bladder cancer across 5 unbalanced batches.
+Because it is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
 
@@
 
+The GSE37199 dataset contains microarray data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
+We accessed a version of this data from the Open Science Framework that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced.
+Additionally, this dataset has two levels of batches (``centre'' and ``plate'').
+We identified this as an item of potential interest: can Confounded adjust for multiple batch effects simultaneously?
+However, we left this question for future research and chose instead to focus on the ``plate'' effect for the time being.
+
 \subsubsection{TCGA Pan-cancer Data}
+
+The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
+In a previous study, we used this dataset and attempted to remove the effect of tissue type from this dataset using ComBat \cite{dayton_classifying_2017-1}.
+However, we found that a strong nonlinear signal could still be identified in the data after adjustment.
+We used the same dataset that we tidied in this study to see if we could remove this nonlinear batch signal more completely.
+This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
+We used this dataset as a way to test whether Confounded works on RNA-Seq data as well as to test whether we could remove batch effects that we know ComBat cannot.
+
+\subsection{Other adjusters}
+
+We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \cite{johnson_adjusting_2007}.
+We implemented the mean and scale adjuster in the R programming language \cite{r_core_team_r_2014} and used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
+We wanted to also compare with SVA \cite{leek_capturing_2007} and the Shaham lab's work \cite{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods to work.
+With SVA, the documentation showed how to identify surrogate variables and use them in modeling, but we were unable to find examples of removing the effects that those surrogate variables represent.
+We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code (see \ref{section:mmd} for more information).
 
@@
 
+In order to determine how well Confounded removes batch effects as compared to other batch correction software, we calculated several metrics.
+First, we calculated mean squared error to see how different the output is from the input.
+Second, we calculated sample maximum mean discrepancy to see how differently distributed the batches are from each other.
+Third, we determined classification accuracy for several machine learning classifiers unrelated to deep neural networks based on batch in order to determine whether batch can still be identified post-adjustement.
+Finally, we determined the same classification accuracies but based on ``true class'' labels to determine how well important signal is maintained after adjustment.
+
 \subsubsection{Mean squared error}
+
+Mean squared error (MSE) is a measure of how much two vectors or matrices deviate from one another.
+It is commonly used as a loss value in autoencoders to make the network minimize the difference between the input and output values.
+We used MSE as a constraint on the autoencoder in Confounded and wanted to see how well other batch correction software maintain patterns in the input data as measured by MSE.
+
+We calculate MSE by the following:
 
@@
 
-\subsubsection{Maximum mean discrepancy}
+Where $x$ represents the values in the autoencoder's input and $\hat{x}$ represents the approximated values in the output.
+If $x = \hat{x}$, MSE will be 0.
 
-Sample MMD:
+\subsubsection{Maximum mean discrepancy} \label{section:mmd}
+
+A recent paper also used neural networks to remove batch effects \cite{shaham_removal_2017}.
+Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
+We wanted to compare our network against this network, so we calculated MMD for each adjuster.
+Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).
+However, we still calculated this metric to determine whether batches looked like they came from the same distribution after adjustement.
+
+We calculated MMD using the following formula:
 
@@
 	\label{mmd}
-	MMD = \frac{1}{n^2}\sum_{i=1}^n{\sum_{j=1}^n{k(x_i, x_j)}} - \frac{2}{nm}\sum_{i=1}^n{\sum_{j=1}^m{k(x_i, y_j)}} + \frac{1}{m^2}\sum_{i=1}^m{\sum_{j=1}^m{k(y_i, y_j)}}
+	MMD(x, y) = \frac{1}{n^2}\sum_{i=1}^n{\sum_{j=1}^n{k(x_i, x_j)}} - \frac{2}{nm}\sum_{i=1}^n{\sum_{j=1}^m{k(x_i, y_j)}} + \frac{1}{m^2}\sum_{i=1}^m{\sum_{j=1}^m{k(y_i, y_j)}}
 \end{equation}
 
-Where $k(x, y)$ is the Gaussian kernel between $x$ and $y$ as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel}.
+Where $k(x, y)$ is the Gaussian kernel between $x$ and $y$ as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel}, and where $x$ represents the data from one batch and $y$ represents the data from another batch.
+In cases where there were more than two batches (e.g. $x$, $y$, and $z$), we averaged the pairwise MMD values to calculate an overall MMD:
+
+\begin{equation}
+	\label{mmd-mean}
+	MMD_{total}(x, y, z) = \frac{1}{3}[MMD(x, y) + MMD(x, z) + MMD(y, z)]
+\end{equation}
+% TODO: make this formula ^ apply to more than 3 batches
 
@@
 
-\begin{equation}
-	\label{simple_equation}
-	\alpha = \sqrt{ \beta }
-\end{equation}
+We used four classifiers from the scikit-learn Python library in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
+Our true classes were as follows:
 
-\section{Results}
+\begin{itemize}
+	\item MNIST: digit
+	\item bladderbatch: presence of cancer
+	\item GSE37199: stage of cancer
+	\item TCGA: presence of a TP53 mutation
+\end{itemize}
+
+We calculated the average of classification accuracies for four-fold cross-validation repeated 10 times.
+We interpret lower accuracy for batch classification as meaning that the batch is removed more effectively.
+We also interpret higher true class classification as meaning that the important signal is not lost during the process of adjustment.
+Therefore, given output data from the ideal batch adjuster, batch classification would be no better than random for any classification algorithm, and true class accuracy would be no lower than accuracy for the unadjusted data.
+We report these values in section \ref{results}.
+% TODO: ^ This
+
+\section{Results} \label{results}
 

### 2019-02-13 thesis.tex
Add figures files

@@
-\documentclass[titlepage]{article}
+\documentclass[notitlepage]{article}
 \usepackage{graphicx}
@@
 \usepackage[square,numbers]{natbib}
+\usepackage[nomarkers]{endfloat}
+\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
+\linespread{1.5}
 \bibliographystyle{unsrtnat}
+% TODO: increase spacing, decrease margins.
 
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/network}
+	\includegraphics[width=4.5in]{figures/rough/adjuster_workflow}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/batch_accuracy}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/network}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mmd}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mnist}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mse}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/network}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/pca}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/training_loss}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/true_class_accuracy}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/network}
+	\caption{Network architecture}
+	\label{simulationfigure}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/tsne}
 	\caption{Network architecture}
@@
 
+% TODO: put all the figures in here
+
+\subsection{Effects are still detectable after ComBat and scale adjusting}
+
+\subsection{Effects not removed by other adjusters are removed by Confounded}
+
+\subsection{Important signal is not removed by Confounded}
+
 \section{Discussion}

### 2019-02-13 thesis.tex
Fix tables
Put figures in the right sections

@@
 \usepackage[square,numbers]{natbib}
-\usepackage[nomarkers]{endfloat}
+\usepackage[nomarkers,nolists]{endfloat}
+\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
 \usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
+\usepackage{csvsimple}
 \linespread{1.5}
 \bibliographystyle{unsrtnat}
-% TODO: increase spacing, decrease margins.
 
@@
 \emph{[Explanation of what's been done in batch adjusting so far].}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/adjuster_workflow}
+	\caption{\textbf{Batch adjustment justification and steps} (a) When measurements are collected from a sample ($X$), systematic effects ($H$) also affect the measurements.
+(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
+(c) In orer to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
+(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
+	\label{fig:workflow}
+\end{figure}
 
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/adjuster_workflow}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/batch_accuracy}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
 	\includegraphics[width=4.5in]{figures/rough/network}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/mmd}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/mnist}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/mse}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/network}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/pca}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/training_loss}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/true_class_accuracy}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/network}
-	\caption{Network architecture}
-	\label{simulationfigure}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/tsne}
-	\caption{Network architecture}
-	\label{simulationfigure}
+	\caption{\textbf{Network architecture} of Confounded.
+Data with batch effects (represented by different colors) are input into an autoencoder.
+The output of the autoencoder is classified by a discriminator network based on batch.
+The autoencoder is then penalized based on the success of the discriminator.
+Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
+	\label{fig:network}
 \end{figure}
@@
 This training typically took roughly 30 minutes to complete, including time spent to load the input and to save the output.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/training_loss}
+	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
+Over the course of training, the autoencoder more faithfully replicates the input data.
+The autoencoder also seems to introduce noise (see around iteration 3100) in response to the discriminator's slight improvements.}
+	\label{fig:training_loss}
+\end{figure}
 
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mnist}
+	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment, (b), with artificial noise added, (c) adjusted for noise by ComBat, and (d) adjusted for noise by Confounded.
+Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
+	\label{fig:mnist}
+\end{figure}\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/pca}
+	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.}
+	\label{fig:pca}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/tsne}
+	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.}
+	\label{fig:tsne}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mse}
+	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.}
+	\label{fig:mse}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mmd}
+	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. In cases with more than two batches, MMD is computed pairwise between each batch and averaged.}
+	\label{fig:mmd}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/batch_accuracy}
+	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.}
+	\label{fig:batch}
+\end{figure}
+\begin{table}
+	\centering
+	\csvautotabular{figures/rough/batch.csv}
+	\caption{Batch classification accuracy}
+	\label{tab:batch}
+\end{table}
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/true_class_accuracy}
+	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.}
+	\label{fig:true_class}
+\end{figure}
+\begin{table}
+	\centering
+	\csvautotabular{figures/rough/true_class.csv}
+	\caption{True class classification accuracies}
+	\label{tab:true_class}
+\end{table}
+
 \section{Discussion}

### 2019-02-15 thesis.tex
Update captions

@@
 	\includegraphics[width=4.5in]{figures/rough/batch_accuracy}
-	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.}
+	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers. See also Table \ref{tab:batch}.}
 	\label{fig:batch}
@@
 	\csvautotabular{figures/rough/batch.csv}
-	\caption{Batch classification accuracy}
+	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
+The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
+See also Figure \ref{fig:batch}.}
 	\label{tab:batch}
@@
 	\includegraphics[width=4.5in]{figures/rough/true_class_accuracy}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.}
+	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times. See also Table \ref{tab:true_class}.}
 	\label{fig:true_class}
@@
 	\csvautotabular{figures/rough/true_class.csv}
-	\caption{True class classification accuracies}
+	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
+After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
+See also Figure \ref{fig:true_class}}
 	\label{tab:true_class}

### 2019-02-19 thesis.tex

@@
 
-\subsection{Effects are still detectable after ComBat and scale adjusting}
+\subsection{Batch effects are still detectable after ComBat and scale adjusting}
 
-\subsection{Effects not removed by other adjusters are removed by Confounded}
-
-\subsection{Important signal is not removed by Confounded}
+PCA and t-SNE plots seem to show a decrease in separability after adjustment with various adjusters; however, previous research has shown that these plots aren't completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
+Figure \ref{fig:pca}, Figure \ref{fig:tsne}
 
 \begin{figure}
-	\centering
-	\includegraphics[width=4.5in]{figures/rough/mnist}
-	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment, (b), with artificial noise added, (c) adjusted for noise by ComBat, and (d) adjusted for noise by Confounded.
-Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
-	\label{fig:mnist}
-\end{figure}\begin{figure}
 	\centering
@@
 \end{figure}
+
+The MSE and MMD metrics show mixed success.
+Figure \ref{mse}, Figure \ref{mmd}.
+% TODO: this might not be the case after running with updated code...
+
 \begin{figure}
@@
 \end{figure}
+
+We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable.
+
+ComBat performs very well on very small datasets and on traditional batch effects.
+
+Although ComBat removes batch well enough to greatly decrease the accuracy of Naive Bayes (a classifier that doesn't account for interactions between variables) in the larger datasets (MNIST and TCGA), Random Forests is still able to predict batch with a high degree of accuracy after ComBat adjustment.
+
+\subsection{Batch effects not removed by other adjusters are removed by Confounded}
+
+Confounded slightly decreases batch accuracy with the MNIST dataset.
+Figure \ref{fig:batch}, Table \ref{tab:batch}
+
 \begin{figure}
@@
 \end{table}
+
+Confounded greatly decreases batch accuracy with the TCGA dataset.
+
+Additional modifications to the network may increase Confounded's efficacy with the larger datasets and possibly with the smaller ones as well.
+
+\subsection{Important signal is not removed by Confounded}
+
+With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat.
+Figure \ref{fig:true_class}, Table \ref{tab:true_class}
+
 \begin{figure}
@@
 
+With the larger datasets, Confounded seems to be losing some accuracy.
+Figure \ref{fig:mnist}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=4.5in]{figures/rough/mnist}
+	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment, (b), with artificial noise added, (c) adjusted for noise by ComBat, and (d) adjusted for noise by Confounded.
+Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
+	\label{fig:mnist}
+\end{figure}
+
+Loss of accuracy may be ameliorated by longer training.
+Or it might be an acceptable amount of lost accuracy, depending on the application.
+
 \section{Discussion}

### 2019-02-26 thesis.tex
Copy intro from prospectus

@@
 
+When working with biological data, problems can arise from combining data from multiple sources.
+For example, if two labs quantify the same 300 RNA samples for lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements.
+If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
+These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
+In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014}.
+Referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as expected \citep{gilad_reanalysis_2015}.
+This type of bias can be even more pronounced when using data collected from different tissue types and can skew results, such as in pan-cancer analyses \citep{dayton_classifying_2017}.
+
+Several methods exist for removing batch effects from biological datasets.
+Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
+ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to identify and remove batch effects.
+Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
+However, recent advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
+
+Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
+This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
+Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
+Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g.
+which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
+which gene is consistently upregulated in a disease) \citep{louizos_causal_2017}.
+Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
+1. to learn as much as possible about the input data and
+2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014}.
+This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015}; the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
+Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
+
+Our approach will build on this previous research by creating an adversarial autoencoder that removes confounding effects from multi-origin data.
+This autoencoder will be packaged and made available to allow other researchers to remove these effects from their own data to enable further analysis.
+
 \begin{figure}
@@
 	\includegraphics[width=4.5in]{figures/rough/adjuster_workflow}
-	\caption{\textbf{Batch adjustment justification and steps} (a) When measurements are collected from a sample ($X$), systematic effects ($H$) also affect the measurements.
+	\caption{\textbf{Batch adjustment justification and steps}
+	(a) When measurements are collected from a sample ($X$), systematic effects ($H$) also affect the measurements.
 	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
-(c) In orer to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
+	(c) In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
 	(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
@@
 \end{figure}
+
+-------
 
@@
 
-\subsection{Batch effects are still detectable after ComBat and scale adjusting}
+\subsection{Some batch effects not removed by other adjusters are removed by Confounded}
 
-PCA and t-SNE plots seem to show a decrease in separability after adjustment with various adjusters; however, previous research has shown that these plots aren't completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
-Figure \ref{fig:pca}, Figure \ref{fig:tsne}
+Various previous studies use differing methods to show qualitatively and quantitatively how well batch effects have been removed.
+We used several plots and metrics to compare adjusters, including PCA, t-SNE, MSE, MMD, and classifier batch prediction accuracy (for various classifiers).
+Classifier accuracy may be a better indicator of adjuster success than other metrics because many classifiers are able to identify nonlinear interactions;
+therefore, a decrease in classifier accuracy represents that nonlinear batch effects have been removed.
+
+PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods (see Figures \ref{fig:pca} and \ref{fig:tsne})
+However, previous research has shown that these plots aren't completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
+Additionally, Confounded seems to change the overall shape of the distribution in these two plots, which may indicate an unwanted change to the underlying data.
 
@@
 
-The MSE and MMD metrics show mixed success.
-Figure \ref{mse}, Figure \ref{mmd}.
+The MSE and MMD metrics show mixed success (see Figures \ref{mse} and \ref{mmd}).
 % TODO: this might not be the case after running with updated code...
@@
 	\includegraphics[width=4.5in]{figures/rough/mmd}
-	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. In cases with more than two batches, MMD is computed pairwise between each batch and averaged.}
+	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches.
+	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.}
 	\label{fig:mmd}
@@
 
+Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat seems to perform better on smaller datasets (see Figure \ref{fig:batch}).
 We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable.
-
-ComBat performs very well on very small datasets and on traditional batch effects.
+With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
+Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
+This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
+Although Naive Bayes is no longer able to identify batch effects in the data after ComBat adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
+In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
+This indicates that while ComBat's performance may still be superior for smaller, more traditional expression datasets, Confounded may be better with newer larger datasets.
 
 Although ComBat removes batch well enough to greatly decrease the accuracy of Naive Bayes (a classifier that doesn't account for interactions between variables) in the larger datasets (MNIST and TCGA), Random Forests is still able to predict batch with a high degree of accuracy after ComBat adjustment.
-
-\subsection{Batch effects not removed by other adjusters are removed by Confounded}
-
-Confounded slightly decreases batch accuracy with the MNIST dataset.
-Figure \ref{fig:batch}, Table \ref{tab:batch}
 
@@
 	\includegraphics[width=4.5in]{figures/rough/batch_accuracy}
-	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers. See also Table \ref{tab:batch}.}
+	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.
+	See also Table \ref{tab:batch}.}
 	\label{fig:batch}
@@
 
-\subsection{Important signal is not removed by Confounded}
+\subsection{Important signal is still detectable after adjustment by Confounded}
 
@@
 	\includegraphics[width=4.5in]{figures/rough/true_class_accuracy}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times. See also Table \ref{tab:true_class}.}
+	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times.
+	See also Table \ref{tab:true_class}.}
 	\label{fig:true_class}
@@
 	\includegraphics[width=4.5in]{figures/rough/mnist}
-	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment, (b), with artificial noise added, (c) adjusted for noise by ComBat, and (d) adjusted for noise by Confounded.
+	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment,
+	(b) with artificial noise added,
+	(c) adjusted for noise by ComBat, and
+	(d) adjusted for noise by Confounded.
 	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}

### 2019-03-05 thesis.tex
Fix references, start organizing intro

@@
 
-\emph{[Explanation of what's been done in batch adjusting so far].}
+\subsection{Hook / funnel}
+
+\subsection{Awesome stuff that's been done, noble failure}
+
+\subsection{To address this question, we need...}
+
+\subsection{In this study, we address this question by addressing the following hypotheses}
 
 When working with biological data, problems can arise from combining data from multiple sources.
-For example, if two labs quantify the same 300 RNA samples for lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements.
+For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements.
 If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
@@
 In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014}.
-Referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as expected \citep{gilad_reanalysis_2015}.
-This type of bias can be even more pronounced when using data collected from different tissue types and can skew results, such as in pan-cancer analyses \citep{dayton_classifying_2017}.
+Referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
+Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.
 
@@
 Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
-ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to identify and remove batch effects.
+ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
 Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
@@
 which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
-which gene is consistently upregulated in a disease) \citep{louizos_causal_2017}.
+which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
 Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
 1. to learn as much as possible about the input data and
-2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014}.
+2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
 This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015}; the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
@@
 
--------
+\noindent\rule{\textwidth}{1pt}
 
@@
 
-\section{Implementation}
-
-\emph{\href{https://bmcbioinformatics.biomedcentral.com/submission-guidelines/preparing-your-manuscript/software-article}{Oxford Bioinformatics' guidelines} say that you should have an "Implementation" section instead of Materials and Methods.}
+\section{Methods} \label{sec:methods}
 
@@
 Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).
-However, we still calculated this metric to determine whether batches looked like they came from the same distribution after adjustement.
+However, we still calculated this metric to determine whether batches looked like they came from the same distribution after adjustment.
 
@@
 Therefore, given output data from the ideal batch adjuster, batch classification would be no better than random for any classification algorithm, and true class accuracy would be no lower than accuracy for the unadjusted data.
-We report these values in section \ref{results}.
-% TODO: ^ This
+We report these values in section \ref{sec:results}.
 
-\section{Results} \label{results}
+\section{Results} \label{sec:results}
 

### 2019-03-05 thesis.tex
Tidy up the background

@@
 
-\section{Background}
-
-\subsection{Hook / funnel}
-
-\subsection{Awesome stuff that's been done, noble failure}
-
-\subsection{To address this question, we need...}
-
-\subsection{In this study, we address this question by addressing the following hypotheses}
+\section{Background} \label{sec:background}
 
 When working with biological data, problems can arise from combining data from multiple sources.
-For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements.
+For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
 If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
 These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
-In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014}.
-Referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
+In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
+however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
 Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.
-
-Several methods exist for removing batch effects from biological datasets.
-Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
-ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
-Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
-However, recent advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
-
-Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
-This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
-Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
-Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g.
-which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
-which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
-Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
-1. to learn as much as possible about the input data and
-2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
-This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015}; the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
-Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
-
-Our approach will build on this previous research by creating an adversarial autoencoder that removes confounding effects from multi-origin data.
-This autoencoder will be packaged and made available to allow other researchers to remove these effects from their own data to enable further analysis.
 
@@
 
-\noindent\rule{\textwidth}{1pt}
+Several methods exist for removing batch effects from biological datasets.
+Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
+ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
+Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
+As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
+Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
+These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
+Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
+However these recent programs have several issues that make them unusable on many real-world datasets:
+they require that the input data only contains two batches, that the batches are perfectly balanced, and that the batches are sufficiently large.
+These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R SVA package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
+Additionally, the statistics these studies have used to validate results don't test whether complex interactions still remain in the data.
 
-Many tools have been developed to account and adjust for batch effects in expression data.
-However, most of these tools only adjust for linear batch effects without accounting for nonlinear interactions between variables.
-Because of the large number of features in expression data and the nature of these features (i.e.
-transcripts and proteins that interact biologically), we anticipate that these nonlinear effects may still be present even after adjustment with popular tools.
-A few notable examples \cite{shaham_removal_2017,shaham_batch_2018} use deep neural networks to correct for batch effects, but these are optimized for use in large, balanced, dual-batched datasets and are therefore unusable for many real-world examples.
+To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
+Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
+This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
+Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
+Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g.
+which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
+which gene is consistently upregulated in a disease) \citep{louizos_causal_2017}.
+Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
+1. to learn as much as possible about the input data and
+2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014}.
+This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
+Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
 
-\textbf{Hypothesis:}
-Using an adversarial neural network can correct for batch effects more completely than previous tools do. % clarify here what "more completely" means
+In this study, we present Confounded, an adversarial autoencoder that identifies and removes effects between different batches.
+We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
+We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess how much nonlinear effects between batches still remains after adjustment using various classification algorithms.
 
@@
 
-\section{Discussion}
+\section{Discussion} \label{sec:discussion}
 
-\section{Conclusions}
+Most exciting takeaways
+
+\begin{enumerate}
+	\item Nonlinear interactions still remain after ComBat for larger datasets
+	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
+	\item ComBat works really well for small, traditional datasets
+	\item Confounded removes nonlinear effects in larger datasets
+\end{enumerate}
+
+\section{Conclusions} \label{sec:conclusion}
 

### 2019-03-05 thesis.tex
Fix citations

@@
 which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
-which gene is consistently upregulated in a disease) \citep{louizos_causal_2017}.
+which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
 Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
 1. to learn as much as possible about the input data and
-2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014}.
+2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
 This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.

### 2019-04-11 thesis.tex
Add "final" figures

@@
 \usepackage{csvsimple}
+\usepackage{multirow}
+\usepackage[table]{xcolor}
 \linespread{1.5}
@@
 For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
+% Is this ^ the best way to reference figures?
 If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/adjuster_workflow}
+	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow}
 	\caption{\textbf{Batch adjustment justification and steps}
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/network}
+	\includegraphics[width=\columnwidth]{figures/final/network}
 	\caption{\textbf{Network architecture} of Confounded.
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/training_loss}
+	\includegraphics[width=\columnwidth]{figures/rough/training_loss}
 	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/pca}
+	\includegraphics[width=\columnwidth]{figures/final/pca}
 	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.}
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/tsne}
+	\includegraphics[width=\columnwidth]{figures/final/tsne}
 	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.}
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/mse}
+	\includegraphics[width=\columnwidth]{figures/final/mse}
 	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.}
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/mmd}
+	\includegraphics[width=\columnwidth]{figures/final/mmd}
 	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches.
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/batch_accuracy}
+	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
 	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.
@@
 	\centering
-	\csvautotabular{figures/rough/batch.csv}
+	\input{figures/final/batch.tex}
 	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/true_class_accuracy}
+	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
 	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times.
@@
 	\centering
-	\csvautotabular{figures/rough/true_class.csv}
+	\input{figures/final/true_class.tex}
 	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
 	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
-	See also Figure \ref{fig:true_class}}
+	See also Figure \ref{fig:true_class}.}
 	\label{tab:true_class}
@@
 	\centering
-	\includegraphics[width=4.5in]{figures/rough/mnist}
+	\includegraphics[width=\columnwidth]{figures/rough/mnist}
 	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment,
@@
 	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
-	\item ComBat works really well for small, traditional datasets
+	\item ComBat works really well for small, traditional datasets % include flip side for #s 3&4
 	\item Confounded removes nonlinear effects in larger datasets
+	\item Confounded shows promise for other confounders, not just batch effects (could work with combining rnaseq + microarray) and seems to work with multiple types of data
 \end{enumerate}
-
-\section{Conclusions} \label{sec:conclusion}
 

### 2019-04-11 thesis.tex
Add prospectus abstract

@@
 \begin{abstract}
-	The abstract text goes here.
+	Many biological studies, particularly studies involving the genetics of human disease, could discover items of greater statistical and biological significance if performed with larger data sets.
+	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
+	In order to do so, we would like to apply deep learning to remove these effects and combine molecular biological data sets.
+	We will create a neural network with a dual loss function in order to remove confounding effects from the data while minimizing the amount of change to the input data.
+	This model will output data the same shape and size as the input data but with confounding effects removed.
+	We plan to test the model on artificially constructed data and commonly used real biological data sets and compare the output with the output from other common batch adjustment algorithms.
+	We will also apply the model to combine microarray and RNA-Seq data sets and to remove the tissue signal from pan-cancer data.
+	We will publish our software to enable other scientists to remove confounding effects from their own data using our methodologies.
 \end{abstract}

### 2019-04-11 thesis.tex
Change abstract to present tense

--- 2019-04-11 thesis.tex
@@
 \begin{abstract}
-	The abstract text goes here.
+	Many biological studies, particularly studies involving the genetics of human disease, could discover items of greater statistical and biological significance if performed with larger data sets.
+	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
+	In order to do so, we applied deep learning to remove these effects and combine molecular biological data sets.
+	We created an adversarial variational autoencoder in order to remove confounding effects from the data while minimizing the amount of change to the input data.
+	We tested the model on artificially constructed data and commonly used real biological data sets and compared the output with the output from other common batch adjustment algorithms.
+	We also applied the model to remove the tissue signal from pan-cancer data.
+	Our software is available on GitHub to enable other scientists to remove confounding effects from their own data using our methodologies.
 \end{abstract}

### 2019-04-11 thesis.tex
Add a few points to discussion

@@
 	\item Nonlinear interactions still remain after ComBat for larger datasets
+	\begin{enumerate}
+		\item Random Forests often has high batch classification accuracy after ComBat-adjustment.
+	\end{enumerate}
 	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
+	\begin{enumerate}
+		\item Figs \ref{fig:mmd} and \ref{fig:mse} both show that ComBat performed better than Confounded for the MNIST dataset, but the effects left after ComBat finished are still identifiable by Random Forests with 99.9\% accuracy (Table \ref{tab:batch}).
+	\end{enumerate}
 	\item ComBat works really well for small, traditional datasets % include flip side for #s 3&4

### 2019-04-18 figures/final/datasets.tex
Update the whole paper, particularly the results

\begin{tabular}{|p{0.18\linewidth}|p{0.14\linewidth}|p{0.1\linewidth}|p{0.15\linewidth}|p{0.23\linewidth}|p{0.16\linewidth}|}
  \hline
  Dataset         & Dimensions   & Number of Batches & Batch Label      & True Class Label       & Data Type        \\
  \hline
  Bladder Batch   & $57\times22,283$  & 5                 & Batch            & Cancer status          & Microarray       \\
  \hline
  GSE37199        & $93\times20,024$  & 2                 & Plate            & Cancer stage           & Microarray       \\
  \hline
  MNIST           & $10,000\times784$ & 2                 & Artificial batch & Digit                  & Grayscale images \\
  \hline
  TCGA Pan-Cancer & $9,366\times325$  & 25                & Cancer Type      & TP53 mutation presence & RNA-Seq \\
  \hline
\end{tabular}

### 2019-04-18 supplement.tex
Update the whole paper, particularly the results

--- 2019-04-11 thesis.tex
@@
-\documentclass[notitlepage]{article}
-\usepackage{graphicx}
-\usepackage{authblk}
-\usepackage{hyperref}
-\usepackage[square,numbers]{natbib}
-\usepackage[nomarkers,nolists]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
-\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
-\usepackage{csvsimple}
-\usepackage{multirow}
-\usepackage[table]{xcolor}
-\linespread{1.5}
-\bibliographystyle{unsrtnat}
-
-\begin{document}
-
-\title{Removing nonlinear batch effects using adversarial deep neural networks}
-\author[1]{Jonathan B. Dayton}
-\author[1]{Stephen R. Piccolo}
-\affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
-\date{}
-
-\maketitle
-
-\begin{abstract}
-	Many biological studies, particularly studies involving the genetics of human disease, could discover items of greater statistical and biological significance if performed with larger data sets.
-	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
-	In order to do so, we applied deep learning to remove these effects and combine molecular biological data sets.
-	We created an adversarial variational autoencoder in order to remove confounding effects from the data while minimizing the amount of change to the input data.
-	We tested the model on artificially constructed data and commonly used real biological data sets and compared the output with the output from other common batch adjustment algorithms.
-	We also applied the model to remove the tissue signal from pan-cancer data.
-	Our software is available on GitHub to enable other scientists to remove confounding effects from their own data using our methodologies.
-\end{abstract}
-
-\section{Background} \label{sec:background}
-
-When working with biological data, problems can arise from combining data from multiple sources.
-For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
-% Is this ^ the best way to reference figures?
-If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
-These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
-In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
-however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
-Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow}
-	\caption{\textbf{Batch adjustment justification and steps}
-	(a) When measurements are collected from a sample ($X$), systematic effects ($H$) also affect the measurements.
-	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
-	(c) In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
-	(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
-	\label{fig:workflow}
-\end{figure}
-
-Several methods exist for removing batch effects from biological datasets.
-Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
-ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
-Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
-As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
-Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
-These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
-Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
-However these recent programs have several issues that make them unusable on many real-world datasets:
-they require that the input data only contains two batches, that the batches are perfectly balanced, and that the batches are sufficiently large.
-These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R SVA package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
-Additionally, the statistics these studies have used to validate results don't test whether complex interactions still remain in the data.
-
-To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
-Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
-This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
-Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
-Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g.
-which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
-which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
-Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
-1. to learn as much as possible about the input data and
-2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
-This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
-Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
-
-In this study, we present Confounded, an adversarial autoencoder that identifies and removes effects between different batches.
-We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
-We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess how much nonlinear effects between batches still remains after adjustment using various classification algorithms.
-
-\section{Methods} \label{sec:methods}
-
-\subsection{Network Structure}
-
-We used an adversarial autoencoder network to model and remove the batch effects.
-This network was structured in two parts: an autoencoder to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder.
-By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over training to output the expression data with all signs of batch effects removed.
-The neural network was implemented in TensorFlow under Python 3.6.
-All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final activations in the autoencoder and the discriminator, which used the sigmoid for their final activations.
 We also tried replacing the ReLU activations with SELUs \cite{klambauer_self-normalizing_2017} but saw little if any improvement and therefore returned to using ReLUs.
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/network}
-	\caption{\textbf{Network architecture} of Confounded.
-	Data with batch effects (represented by different colors) are input into an autoencoder.
-	The output of the autoencoder is classified by a discriminator network based on batch.
-	The autoencoder is then penalized based on the success of the discriminator.
-	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
-	\label{fig:network}
-\end{figure}
+Autoencoder
 
-\subsubsection{Autoencoder}
-
-We initially implemented a basic fully connected autoencoder with two layers for encoding and two layers for decoding and with a smaller (about $\frac{1}{5}$ the size of the input) code layer.
 In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
@@
 We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
-% TODO: implement and test the variational ae
 
-\subsubsection{Discriminator}
+Discriminator
 
-The discriminator is given the output of the autoencoder and was trained to determine the original batch of that output.
-It consists of a variable number of fully connected layers of size 500 and a one-hot encoded output for determining batch.
 We noticed that the discriminator was overfitting on the data.
 Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
-In order to combat this overfitting, we also added 50\%-probability dropout to each layer of the discriminator.
-After adding dropout, the network appeared to no longer be overfitting.
 
-\subsubsection{Loss functions}
+Datasets
 
-Three loss functions were used to train the network.
-First, the autoencoder was trained on mean squared error \ref{mse} ($MSE_A$) between the input and output layers.
-Second, the discriminator was trained on mean squared error ($MSE_D$) between its output and a one-hot encoding of the samples' batch labels.
-Finally, the autoencoder was also trained on a combination of the two previous losses,
-
-\begin{equation}
-	\label{dual_loss}
-	Loss = MSE_A - \lambda{}(MSE_D)
-\end{equation}
-
-Where the $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
-
-\subsubsection{Training}
-
-In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of either 0.001 or 0.0001.
-We typically trained for 10,000 iterations on mini-batches of size 100.
-This training typically took roughly 30 minutes to complete, including time spent to load the input and to save the output.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/training_loss}
-	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
-	Over the course of training, the autoencoder more faithfully replicates the input data.
-	The autoencoder also seems to introduce noise (see around iteration 3100) in response to the discriminator's slight improvements.}
-	\label{fig:training_loss}
-\end{figure}
-
-\subsection{Datasets}
-
-In order to test both theoretical and practical differences between Confounded and previous methods, we compared them for a variety of different datasets.
 Working with each new dataset enlightened us of a new edge case that hadn't yet been addressed in our code.
@@
 
-\subsubsection{MNIST}
+GSE37199
 
-The MNIST digits dataset \cite{lecun_mnist_nodate} is a database of images of handwritten digits that are size-normalized and centered.
-It contains 60,000 training images and 10,000 test images.
-In order to use this dataset, we flattened each 28 by 28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit information.
-We limited our dataset to only the 10,000 test images.
-
-By using MNIST, we were able to visually assess how much true signal (in this case, the shape and digit of each handwritten digit image) is preserved after batch adjustment.
-Although convolutional layers would typically be used when working with images such as the MNIST dataset, we chose to use only fully connected layers even for this image dataset.
-In this way, we show that the autoencoder is still able to find and represent spatial relationships without explicitly defining spatial relationships in the model while testing the same network we use on expression data, where no spatial relationships are inherent.
-
-\paragraph{Synthetic Batch Effects}
-
-Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear batch effects.
-To do so, we wrote a Python script that would take the MNIST data in, apply some nonlinear effect, and output the adjusted data.
-We applied nonlinear effects by iteratively realizing vectors of normal values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
-We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
-The same normal vectors were applied to each image in a batch.
-Finally, random noise was added to each image in order to prevent images in a batch from being too similar to each other.
-
-\subsubsection{Bladderbatch}
-
-The bladderbatch dataset is a microarray transcriptome expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
-It has been made available in an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package to illustrate how to batch-adjust using ComBat \cite{leek_sva_2017}.
-It contains expression values for 57 samples of patients with and without bladder cancer across 5 unbalanced batches.
-Because it is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
-
-\subsubsection{GSE37199}
-
-The GSE37199 dataset contains microarray data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this data from the Open Science Framework that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
-This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced.
 Additionally, this dataset has two levels of batches (``centre'' and ``plate'').
@@
 
-\subsubsection{TCGA Pan-cancer Data}
+Other methods
 
-The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
-In a previous study, we used this dataset and attempted to remove the effect of tissue type from this dataset using ComBat \cite{dayton_classifying_2017-1}.
-However, we found that a strong nonlinear signal could still be identified in the data after adjustment.
-We used the same dataset that we tidied in this study to see if we could remove this nonlinear batch signal more completely.
-This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
-We used this dataset as a way to test whether Confounded works on RNA-Seq data as well as to test whether we could remove batch effects that we know ComBat cannot.
-
-\subsection{Other adjusters}
-
-We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \cite{johnson_adjusting_2007}.
-We implemented the mean and scale adjuster in the R programming language \cite{r_core_team_r_2014} and used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
 We wanted to also compare with SVA \cite{leek_capturing_2007} and the Shaham lab's work \cite{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods to work.
@@
 
-\subsection{Statistics and Metrics}
-
-In order to determine how well Confounded removes batch effects as compared to other batch correction software, we calculated several metrics.
-First, we calculated mean squared error to see how different the output is from the input.
-Second, we calculated sample maximum mean discrepancy to see how differently distributed the batches are from each other.
-Third, we determined classification accuracy for several machine learning classifiers unrelated to deep neural networks based on batch in order to determine whether batch can still be identified post-adjustement.
-Finally, we determined the same classification accuracies but based on ``true class'' labels to determine how well important signal is maintained after adjustment.
-
-\subsubsection{Mean squared error}
-
-Mean squared error (MSE) is a measure of how much two vectors or matrices deviate from one another.
-It is commonly used as a loss value in autoencoders to make the network minimize the difference between the input and output values.
-We used MSE as a constraint on the autoencoder in Confounded and wanted to see how well other batch correction software maintain patterns in the input data as measured by MSE.
+MSE
 
@@
 
-\subsubsection{Maximum mean discrepancy} \label{section:mmd}
+MMD
 
-A recent paper also used neural networks to remove batch effects \cite{shaham_removal_2017}.
-Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
 We wanted to compare our network against this network, so we calculated MMD for each adjuster.
 Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).
-However, we still calculated this metric to determine whether batches looked like they came from the same distribution after adjustment.
 
-We calculated MMD using the following formula:
+following formula:
 
@@
 % TODO: make this formula ^ apply to more than 3 batches
-
-\subsubsection{Classification accuracy}
-
-We used four classifiers from the scikit-learn Python library in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
-Our true classes were as follows:
-
-\begin{itemize}
-	\item MNIST: digit
-	\item bladderbatch: presence of cancer
-	\item GSE37199: stage of cancer
-	\item TCGA: presence of a TP53 mutation
-\end{itemize}
-
-We calculated the average of classification accuracies for four-fold cross-validation repeated 10 times.
-We interpret lower accuracy for batch classification as meaning that the batch is removed more effectively.
-We also interpret higher true class classification as meaning that the important signal is not lost during the process of adjustment.
-Therefore, given output data from the ideal batch adjuster, batch classification would be no better than random for any classification algorithm, and true class accuracy would be no lower than accuracy for the unadjusted data.
-We report these values in section \ref{sec:results}.
-
-\section{Results} \label{sec:results}
-
-% TODO: put all the figures in here
-
-\subsection{Some batch effects not removed by other adjusters are removed by Confounded}
-
-Various previous studies use differing methods to show qualitatively and quantitatively how well batch effects have been removed.
-We used several plots and metrics to compare adjusters, including PCA, t-SNE, MSE, MMD, and classifier batch prediction accuracy (for various classifiers).
-Classifier accuracy may be a better indicator of adjuster success than other metrics because many classifiers are able to identify nonlinear interactions;
-therefore, a decrease in classifier accuracy represents that nonlinear batch effects have been removed.
-
-PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods (see Figures \ref{fig:pca} and \ref{fig:tsne})
-However, previous research has shown that these plots aren't completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
-Additionally, Confounded seems to change the overall shape of the distribution in these two plots, which may indicate an unwanted change to the underlying data.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/pca}
-	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.}
-	\label{fig:pca}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/tsne}
-	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.}
-	\label{fig:tsne}
-\end{figure}
-
-The MSE and MMD metrics show mixed success (see Figures \ref{mse} and \ref{mmd}).
-% TODO: this might not be the case after running with updated code...
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/mse}
-	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.}
-	\label{fig:mse}
-\end{figure}
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/mmd}
-	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches.
-	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.}
-	\label{fig:mmd}
-\end{figure}
-
-Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat seems to perform better on smaller datasets (see Figure \ref{fig:batch}).
-We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable.
-With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
-Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
-This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
-Although Naive Bayes is no longer able to identify batch effects in the data after ComBat adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
-In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
-This indicates that while ComBat's performance may still be superior for smaller, more traditional expression datasets, Confounded may be better with newer larger datasets.
-
-Although ComBat removes batch well enough to greatly decrease the accuracy of Naive Bayes (a classifier that doesn't account for interactions between variables) in the larger datasets (MNIST and TCGA), Random Forests is still able to predict batch with a high degree of accuracy after ComBat adjustment.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
-	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.
-	See also Table \ref{tab:batch}.}
-	\label{fig:batch}
-\end{figure}
-\begin{table}
-	\centering
-	\input{figures/final/batch.tex}
-	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
-	The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
-	See also Figure \ref{fig:batch}.}
-	\label{tab:batch}
-\end{table}
-
-Confounded greatly decreases batch accuracy with the TCGA dataset.
-
-Additional modifications to the network may increase Confounded's efficacy with the larger datasets and possibly with the smaller ones as well.
-
-\subsection{Important signal is still detectable after adjustment by Confounded}
-
-With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat.
-Figure \ref{fig:true_class}, Table \ref{tab:true_class}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times.
-	See also Table \ref{tab:true_class}.}
-	\label{fig:true_class}
-\end{figure}
-\begin{table}
-	\centering
-	\input{figures/final/true_class.tex}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
-	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
-	See also Figure \ref{fig:true_class}.}
-	\label{tab:true_class}
-\end{table}
-
-With the larger datasets, Confounded seems to be losing some accuracy.
-Figure \ref{fig:mnist}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/mnist}
-	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment,
-	(b) with artificial noise added,
-	(c) adjusted for noise by ComBat, and
-	(d) adjusted for noise by Confounded.
-	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
-	\label{fig:mnist}
-\end{figure}
-
-Loss of accuracy may be ameliorated by longer training.
-Or it might be an acceptable amount of lost accuracy, depending on the application.
-
-\section{Discussion} \label{sec:discussion}
-
-Most exciting takeaways
-
-\begin{enumerate}
-	\item Nonlinear interactions still remain after ComBat for larger datasets
-	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
-	\item ComBat works really well for small, traditional datasets % include flip side for #s 3&4
-	\item Confounded removes nonlinear effects in larger datasets
-	\item Confounded shows promise for other confounders, not just batch effects (could work with combining rnaseq + microarray) and seems to work with multiple types of data
-\end{enumerate}
-
-\bibliography{references}
-
-\end{document}

### 2019-04-18 thesis.tex
Update the whole paper, particularly the results

--- 2019-04-11 thesis.tex
@@
 	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
-	In order to do so, we applied deep learning to remove these effects and combine molecular biological data sets.
+	In order to combine date produced under different conditions, we applied deep learning to remove these effects and combine molecular biological data sets.
 	We created an adversarial variational autoencoder in order to remove confounding effects from the data while minimizing the amount of change to the input data.
@@
 When working with biological data, problems can arise from combining data from multiple sources.
-For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
+For example, if two labs sequence and quantify the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
 % Is this ^ the best way to reference figures?
@@
 however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
-Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.
+Systemic bias can be even more pronounced when using data collected from different tissue types and can skew results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.
 
@@
 	\caption{\textbf{Batch adjustment justification and steps}
-	(a) When measurements are collected from a sample ($X$), systematic effects ($H$) also affect the measurements.
+	(a) When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
 	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
@@
 ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
-Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
+Since both of these methods use linear adjustments to remove batch effects, they cannot account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
 As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
@@
 Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
-However these recent programs have several issues that make them unusable on many real-world datasets:
-they require that the input data only contains two batches, that the batches are perfectly balanced, and that the batches are sufficiently large.
+However, these recent programs have several items that complicate their usability on many real-world datasets:
+they require that the input data only contains two batches, that the batches are sufficiently large, and that the batches are perfectly balanced.
 These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R SVA package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
@@
 This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
-Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
-Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g.
-which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g.
+Research has shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
+Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output to be as similar to the input as possible \citep{hinton_reducing_2006}.
+Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on dataset-specific confounding effects (e.g.
+which researcher collected the data) instead of learning based on practically meaningful causal effects (e.g.
 which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
-Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
+Recently, researchers have experimented with discouraging neural networks from ,learning based on domain-specific information by giving them two objective functions:
 1. to learn as much as possible about the input data and
-2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
-This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
-Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
+2. to forget any patterns that help distinguish between domains \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
+This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``penalized'' for retaining class-specific patterns in the output.
 
@@
 We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
-We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess how much nonlinear effects between batches still remains after adjustment using various classification algorithms.
+We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess the extent to which batch effects remain after adjustment using various classification algorithms.
 
@@
 We used an adversarial autoencoder network to model and remove the batch effects.
-This network was structured in two parts: an autoencoder to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder.
-By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over training to output the expression data with all signs of batch effects removed.
-The neural network was implemented in TensorFlow under Python 3.6.
-All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final activations in the autoencoder and the discriminator, which used the sigmoid for their final activations.
-We also tried replacing the ReLU activations with SELUs \cite{klambauer_self-normalizing_2017} but saw little if any improvement and therefore returned to using ReLUs.
-
-We toyed with the idea of using convolutional layers \cite{krizhevsky_imagenet_2012-1} in the network because of their incredible success in the realm of image recognition.
-We thought at first that a 20,000-dimensional vector of expression values was quite similar to a 200x100 image and that similar techniques could be used for both, which was encouraging because convolutional filters are quite efficient for processing images.
-However, we were unable to devise a representation or arrangement of gene expression data points in which convolutional filters would make sense.
-Although one could easily wrap a gene expression vector into two dimensions and arrange the genes by location in the genome, each data point fundamentally represents a completely different gene product.
-Contrast this with image data, where each pixel represents the same data type (i.e.
-relative light intensities of some wavelength); if each pixel value was shifted to the left by one pixel, the image would still represent the same idea.
-However, if each expression value were to be shifted one position to the left, it would in essence be like mistaking each gene product for another unrelated gene product.
-Therefore, passing convolutional filters over a gene expression array, 1D or 2D, would not make sense since the relationship between a pair of consecutive genes is not necessarily analogous to the relationship between another pair of consecutive genes.
-Because we couldn't justify any arrangement where spatial relationships between genes were consistently meaningful, we chose to only use fully connected layers in Confounded rather than convolutional layers.
-We welcome any future intellectual advances that may allow convolutional networks and other image processing techniques to be applied to expression data.
+This network was structured in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder's output.
+By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data free of batch effects.
+The neural network was implemented in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
+All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.
 
@@
 
-We initially implemented a basic fully connected autoencoder with two layers for encoding and two layers for decoding and with a smaller (about $\frac{1}{5}$ the size of the input) code layer.
-In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
-We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
-We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
-% TODO: implement and test the variational ae
+We implemented the variational autoencoder \cite{louizos_variational_2015} described by Géron \cite{geron_hands-machine_2017}.
+Additional information can be found in the supplementary material.
 
@@
 
-The discriminator is given the output of the autoencoder and was trained to determine the original batch of that output.
-It consists of a variable number of fully connected layers of size 500 and a one-hot encoded output for determining batch.
-We noticed that the discriminator was overfitting on the data.
-Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
-In order to combat this overfitting, we also added 50\%-probability dropout to each layer of the discriminator.
-After adding dropout, the network appeared to no longer be overfitting.
+The discriminator was trained to determine the original batch of the autoencoder's output.
+The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches.
+In order to combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} and batch normalization \cite{ioffe_batch_2015} to each layer of the discriminator.
+These additions seemed to reduce overfitting in the discriminator.
 
@@
 Three loss functions were used to train the network.
-First, the autoencoder was trained on mean squared error \ref{mse} ($MSE_A$) between the input and output layers.
-Second, the discriminator was trained on mean squared error ($MSE_D$) between its output and a one-hot encoding of the samples' batch labels.
-Finally, the autoencoder was also trained on a combination of the two previous losses,
+First, the autoencoder's loss ($L_A$) was calculated by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler divergence of the code layer).
+Second, the discriminator's loss ($L_D$) was trained on sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
+Finally, the autoencoder layers were also trained on a combination of the two previous losses,
 
@@
 	\label{dual_loss}
-	Loss = MSE_A - \lambda{}(MSE_D)
+	L_dual = L_A - \lambda{}(L_D)
 \end{equation}
 
-Where the $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
+The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
+A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
+$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_dual$.
 
@@
 
-In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of either 0.001 or 0.0001.
-We typically trained for 10,000 iterations on mini-batches of size 100.
-This training typically took roughly 30 minutes to complete, including time spent to load the input and to save the output.
+In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
+In each iteration, both $L_D$ and $L_dual$ were optimized.
+When optimizing $L_D$ (i.e. training the discriminator), the autoencoder's weights were frozen, and vice versa.
+For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.
 
@@
 
-In order to test both theoretical and practical differences between Confounded and previous methods, we compared them for a variety of different datasets.
-Working with each new dataset enlightened us of a new edge case that hadn't yet been addressed in our code.
-By applying to our methods to each of the following datasets, our code became much more robust than if we had only developed and tested on one or two toy datasets.
+In order to test both theoretical and practical differences between Confounded and previous methods, we compared them for a variety of datasets of varying sizes and type of data measured.
 
-\subsubsection{Data Format}
-
-Confounded takes input data in the tidy data format \cite{wickham_tidy_2014-1}, namely:
-
-\begin{itemize}
-	\item Each row represents a different sample.
-	\item Each column represents a different variable (either a clinical variable or a gene's expression).
-\end{itemize}
-
-Confounded parses through files in this format to determine which columns are discrete (integer or string) and which are continuous (floating point).
-Each continuous column is interpreted to be a gene expression column and is adjusted for by the network according to the batch column.
-This is one limitation to our software.
-In a future iteration, it would be more robust to allow the user to specify continuous columns that should not be interpreted as gene expression data (e.g.
-ages in years with decimal precision representing fractions of years).
-However for the datasets we tested, this didn't prove to be a problem.
+\begin{table}
+	\centering
+	\input{figures/final/datasets.tex}
+	\caption{\textbf{Dataset information} for each dataset used.}
+	\label{tab:datasets}
+\end{table}
 
@@
 
-By using MNIST, we were able to visually assess how much true signal (in this case, the shape and digit of each handwritten digit image) is preserved after batch adjustment.
-Although convolutional layers would typically be used when working with images such as the MNIST dataset, we chose to use only fully connected layers even for this image dataset.
+Using MNIST allowed us to visually assess how much true signal (in this case, the shape and digit of each handwritten digit image) was preserved after batch adjustment.
+Although convolutional layers are typically used when working with image data, we only used fully connected layers even for this image dataset.
 In this way, we show that the autoencoder is still able to find and represent spatial relationships without explicitly defining spatial relationships in the model while testing the same network we use on expression data, where no spatial relationships are inherent.
@@
 Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear batch effects.
-To do so, we wrote a Python script that would take the MNIST data in, apply some nonlinear effect, and output the adjusted data.
-We applied nonlinear effects by iteratively realizing vectors of normal values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
+To do so, we wrote a Python script to take the MNIST data in, apply a nonlinear effect, and output the adjusted data.
+We applied nonlinear effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
 We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
-The same normal vectors were applied to each image in a batch.
-Finally, random noise was added to each image in order to prevent images in a batch from being too similar to each other.
+The same random vectors were applied to each image in a batch.
+Finally, random noise was added to each image in order to prevent images in a batch from being overly similar to each other.
 
@@
 The bladderbatch dataset is a microarray transcriptome expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
-It has been made available in an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package to illustrate how to batch-adjust using ComBat \cite{leek_sva_2017}.
-It contains expression values for 57 samples of patients with and without bladder cancer across 5 unbalanced batches.
+It has been made available as an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package \cite{leek_sva_2017} to illustrate how to batch-adjust using ComBat.
+It contains expression values for 57 tissue samples from patients with and without bladder cancer across 5 unbalanced batches.
 Because it is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
@@
 This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced.
-Additionally, this dataset has two levels of batches (``centre'' and ``plate'').
-We identified this as an item of potential interest: can Confounded adjust for multiple batch effects simultaneously?
-However, we left this question for future research and chose instead to focus on the ``plate'' effect for the time being.
 
@@
 The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
-In a previous study, we used this dataset and attempted to remove the effect of tissue type from this dataset using ComBat \cite{dayton_classifying_2017-1}.
+In a previous study, we used this dataset and attempted to adjust for the effect of tissue type using ComBat \cite{dayton_classifying_2017-1}.
 However, we found that a strong nonlinear signal could still be identified in the data after adjustment.
-We used the same dataset that we tidied in this study to see if we could remove this nonlinear batch signal more completely.
+We used the same dataset that we tidied in this study to see if we could remove the nonlinear batch effect more completely.
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
@@
 
-\subsection{Other adjusters}
+\subsection{Comparison to other methods}
 
 We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \cite{johnson_adjusting_2007}.
-We implemented the mean and scale adjuster in the R programming language \cite{r_core_team_r_2014} and used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
-We wanted to also compare with SVA \cite{leek_capturing_2007} and the Shaham lab's work \cite{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods to work.
-With SVA, the documentation showed how to identify surrogate variables and use them in modeling, but we were unable to find examples of removing the effects that those surrogate variables represent.
-We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code (see \ref{section:mmd} for more information).
+We implemented the mean and scale adjuster in the R programming language \cite{r_core_team_r_2014}.
+We used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
 
@@
 
-In order to determine how well Confounded removes batch effects as compared to other batch correction software, we calculated several metrics.
+In order to determine how well Confounded removes batch effects compared to other methods, we calculated several metrics.
 First, we calculated mean squared error to see how different the output is from the input.
 Second, we calculated sample maximum mean discrepancy to see how differently distributed the batches are from each other.
-Third, we determined classification accuracy for several machine learning classifiers unrelated to deep neural networks based on batch in order to determine whether batch can still be identified post-adjustement.
-Finally, we determined the same classification accuracies but based on ``true class'' labels to determine how well important signal is maintained after adjustment.
+Third, we determined classification accuracy based on batch for several machine learning classifiers in order to determine whether batch can still be identified post-adjustment.
+Finally, we used the same classifiers to predict ``true class'' labels in order to determine how well important signal is maintained after adjustment.
 
@@
 It is commonly used as a loss value in autoencoders to make the network minimize the difference between the input and output values.
-We used MSE as a constraint on the autoencoder in Confounded and wanted to see how well other batch correction software maintain patterns in the input data as measured by MSE.
-
-We calculate MSE by the following:
-
-\begin{equation}
-	\label{mse}
-	MSE = \frac{1}{n}\sum_{i=1}^n{(\hat{x}_i - x_i)^2}
-\end{equation}
-
-Where $x$ represents the values in the autoencoder's input and $\hat{x}$ represents the approximated values in the output.
-If $x = \hat{x}$, MSE will be 0.
+We wanted to see how well Confounded and other batch correction software maintain patterns in the input data as measured by MSE.
 
@@
 Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
-We wanted to compare our network against this network, so we calculated MMD for each adjuster.
-Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).
-However, we still calculated this metric to determine whether batches looked like they came from the same distribution after adjustment.
-
-We calculated MMD using the following formula:
-
-\begin{equation}
-	\label{mmd}
-	MMD(x, y) = \frac{1}{n^2}\sum_{i=1}^n{\sum_{j=1}^n{k(x_i, x_j)}} - \frac{2}{nm}\sum_{i=1}^n{\sum_{j=1}^m{k(x_i, y_j)}} + \frac{1}{m^2}\sum_{i=1}^m{\sum_{j=1}^m{k(y_i, y_j)}}
-\end{equation}
-
-Where $k(x, y)$ is the Gaussian kernel between $x$ and $y$ as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel}, and where $x$ represents the data from one batch and $y$ represents the data from another batch.
-In cases where there were more than two batches (e.g. $x$, $y$, and $z$), we averaged the pairwise MMD values to calculate an overall MMD:
-
-\begin{equation}
-	\label{mmd-mean}
-	MMD_{total}(x, y, z) = \frac{1}{3}[MMD(x, y) + MMD(x, z) + MMD(y, z)]
-\end{equation}
-% TODO: make this formula ^ apply to more than 3 batches
+We calculated MMD to determine whether batches looked like they came from the same distribution after adjustment.
+We calculated MMD using the formula in this recent paper.
+For the kernel, we used the Gaussian kernel between two batches as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel} \cite{pedregosa_scikit-learn_2011}.
+In cases where there were more than two batches, we averaged all pairwise MMD values to calculate an overall MMD.
 
@@
 
-We used four classifiers from the scikit-learn Python library in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
-Our true classes were as follows:
+We used four classifiers from the scikit-learn 0.19.1 Python library \cite{pedregosa_scikit-learn_2011} in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
+See Table \ref{tab:datasets} for details on which columns were used for training.
 
-\begin{itemize}
-	\item MNIST: digit
-	\item bladderbatch: presence of cancer
-	\item GSE37199: stage of cancer
-	\item TCGA: presence of a TP53 mutation
-\end{itemize}
-
-We calculated the average of classification accuracies for four-fold cross-validation repeated 10 times.
+We calculated the average of classification accuracies for four-fold cross-validation repeated 3 times.
 We interpret lower accuracy for batch classification as meaning that the batch is removed more effectively.
@@
 
-% TODO: put all the figures in here
-
 \subsection{Some batch effects not removed by other adjusters are removed by Confounded}
@@
 We used several plots and metrics to compare adjusters, including PCA, t-SNE, MSE, MMD, and classifier batch prediction accuracy (for various classifiers).
-Classifier accuracy may be a better indicator of adjuster success than other metrics because many classifiers are able to identify nonlinear interactions;
-therefore, a decrease in classifier accuracy represents that nonlinear batch effects have been removed.
 
-PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods (see Figures \ref{fig:pca} and \ref{fig:tsne})
-However, previous research has shown that these plots aren't completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
-Additionally, Confounded seems to change the overall shape of the distribution in these two plots, which may indicate an unwanted change to the underlying data.
+PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods for the GSE37199 dataset (see Figures \ref{fig:pca} and \ref{fig:tsne}).
+However, previous research has shown that these plots are not completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
+In the PCA plot, Confounded appears to maintain a similar distribution to the unadjusted data, indicating that the underlying distribution has been faithfully reproduced by the networks.
+The t-SNE plot shows that the data post-adjustment by Confounded and ComBat appear to cluster less tightly by batch than the unadjusted and scale-adjusted data.
+This may indicate an effective removal of nonlinear effects in both cases.
 
@@
 	\includegraphics[width=\columnwidth]{figures/final/pca}
-	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.}
+	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
+	None of the datasets appear to be linearly separable.
+	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
 	\label{fig:pca}
@@
 	\includegraphics[width=\columnwidth]{figures/final/tsne}
-	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.}
+	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
+	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
 	\label{fig:tsne}
@@
 
-The MSE and MMD metrics show mixed success (see Figures \ref{mse} and \ref{mmd}).
-% TODO: this might not be the case after running with updated code...
+Confounded shows mixed success with the MSE and MMD metrics, but both values were typically lower for Confounded than for the scale adjuster and lower for ComBat than for any other adjuster we tested (See Figures \ref{fig:mse} and \ref{fig:mmd}).
 
@@
 	\includegraphics[width=\columnwidth]{figures/final/mse}
-	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.}
+	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
+	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
+	MSE for unadjusted data will always be 0 because the input data is identical to the output data.
+	Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
 	\label{fig:mse}
@@
 	\includegraphics[width=\columnwidth]{figures/final/mmd}
-	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches.
-	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.}
+	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
+	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
+	In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
 	\label{fig:mmd}
@@
 
-Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat seems to perform better on smaller datasets (see Figure \ref{fig:batch}).
-We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable.
+We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable by any machine learning classifier.
+Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see Figure \ref{fig:batch}).
 With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
@@
 This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
-Although Naive Bayes is no longer able to identify batch effects in the data after ComBat adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
+Although Naive Bayes is no longer able to identify batch effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
 In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
-This indicates that while ComBat's performance may still be superior for smaller, more traditional expression datasets, Confounded may be better with newer larger datasets.
-
-Although ComBat removes batch well enough to greatly decrease the accuracy of Naive Bayes (a classifier that doesn't account for interactions between variables) in the larger datasets (MNIST and TCGA), Random Forests is still able to predict batch with a high degree of accuracy after ComBat adjustment.
+This indicates that while ComBat's performance may work at least as well as Confounded for smaller expression datasets, Confounded may work better with larger datasets.
 
@@
 	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
-	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.
+	\caption{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
+	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
+	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
 	See also Table \ref{tab:batch}.}
@@
 
-Confounded greatly decreases batch accuracy with the TCGA dataset.
-
-Additional modifications to the network may increase Confounded's efficacy with the larger datasets and possibly with the smaller ones as well.
+With the larger datasets in particular, Confounded outperforms the other adjusters.
+On the MNIST dataset, Random Forests is able to detect batch with perfect or near-perfect accuracy after adjustment with the scale adjuster and ComBat, but the highest batch classification accuracy after adjustment by Confounded is Naive Bayes, with an accuracy of 68.8\%.
+With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 87.6\% to 96.3\% and 97.1\% respectively, whereas Confounded decreases the accuracy to 8.8\%.
 
@@
 
-With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat.
-Figure \ref{fig:true_class}, Table \ref{tab:true_class}
+With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see Figure \ref{fig:true_class} and Table \ref{tab:true_class}).
+For the Bladderbatch dataset, true class accuracy is much lower after adjusting with any algorithm, indicating that cancer status and batch may not be independent.
+
+With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
+A look at the MNIST digits before and after adjustment (see Figure \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
+With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs.
+11.4\%), but with TCGA, the accuracy decreases below baseline (66.5\% vs.
+69.8\%) while the other adjusters' accuracies remain above baseline.
+However, preliminary testing indicates that fine-tuning Confounded's parameters to run on these datasets rather than using the default settings may greatly improve Confounded's post-adjustment true class accuracy.
+% TODO: include this^
 
@@
 	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times.
+	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
 	See also Table \ref{tab:true_class}.}
@@
 
-With the larger datasets, Confounded seems to be losing some accuracy.
-Figure \ref{fig:mnist}
-
 \begin{figure}
@@
 
-Loss of accuracy may be ameliorated by longer training.
-Or it might be an acceptable amount of lost accuracy, depending on the application.
-
 \section{Discussion} \label{sec:discussion}
 
-Most exciting takeaways
+\subsection{Classification accuracy quantifies batch removal more accurately than MSE and MMD}
 
-\begin{enumerate}
-	\item Nonlinear interactions still remain after ComBat for larger datasets
-	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
-	\item ComBat works really well for small, traditional datasets % include flip side for #s 3&4
-	\item Confounded removes nonlinear effects in larger datasets
-	\item Confounded shows promise for other confounders, not just batch effects (could work with combining rnaseq + microarray) and seems to work with multiple types of data
-\end{enumerate}
+Classifier accuracy may be a better indicator of adjuster success than other metrics because many classifiers are able to identify nonlinear interactions;
+therefore, a decrease in classifier accuracy represents that nonlinear batch effects have been removed.
+
+Figs \ref{fig:mmd} and \ref{fig:mse} both show that ComBat performed better than Confounded for the MNIST dataset, but the effects left after ComBat finished are still identifiable by Random Forests with 99.9\% accuracy (Table \ref{tab:batch}).
+
+\subsection{Dataset size should be considered when applying batch adjustment}
+
+Nonlinear batch effects remain after traditional batch adjustment for larger datasets
+
+Random Forests often has high batch classification accuracy after ComBat-adjustment.
+This is especially important to take into account when your analysis of your dataset will include some type of machine learning that can detect nonlinear effects.
+
+\subsection{Confounded may have other uses}
+
+Data integration
+
+Other types of data - already shown to work on images, microarray, \& rna seq
 

### 2019-04-18 thesis.tex
Add first draft of discussion section

@@
 	\includegraphics[width=\columnwidth]{figures/rough/mnist}
-	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment,
+	\caption{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits} (a) before any adjustment,
 	(b) with artificial noise added,
@@
 
-Classifier accuracy may be a better indicator of adjuster success than other metrics because many classifiers are able to identify nonlinear interactions;
-therefore, a decrease in classifier accuracy represents that nonlinear batch effects have been removed.
+Our MSE \ref{fig:mse} and MMD \ref{fig:mmd} metrics both show that ComBat performed better than Confounded for the MNIST dataset, but the effects left after ComBat finished are still identifiable by Random Forests with 99.9\% accuracy (Table \ref{tab:batch}).
+Additionally, previous work \cite{dayton_classifying_2017-1} has shown that PCA can be misleading when determining whether batch effects have been sufficiently removed from a dataset.
+Each of these methods seems to focus on linear separability.
 
-Figs \ref{fig:mmd} and \ref{fig:mse} both show that ComBat performed better than Confounded for the MNIST dataset, but the effects left after ComBat finished are still identifiable by Random Forests with 99.9\% accuracy (Table \ref{tab:batch}).
+Many machine learning classification algorithms are able to identify both linear and nonlinear interactions between variables.
+A decrease in classifier accuracy represents that nonlinear batch effects have been decreased.
+Random Forests often has high batch classification accuracy after ComBat-adjustment even when other classifiers are unable to identify differences between batches, whereas SVC often performs no better than baseline (see Figure \ref{fig:batch} and Table \ref{tab:batch}).
+In addition, ComBat often excels at hiding batch effects from Naive Bayes.
+We suggest that it may be important to use several classifiers to identify whether batch effects remain after correction.
 
-\subsection{Dataset size should be considered when applying batch adjustment}
+We encourage future researchers who study batch correction or who use it in their pipelines to use classifier accuracy to determine the degree to which batch effects still remain in their data.
+This is especially important to take into account when analyses will include some type of machine learning that can detect nonlinear effects.
 
-Nonlinear batch effects remain after traditional batch adjustment for larger datasets
+\subsection{Dataset should be considered when applying batch adjustment}
 
-Random Forests often has high batch classification accuracy after ComBat-adjustment.
-This is especially important to take into account when your analysis of your dataset will include some type of machine learning that can detect nonlinear effects.
+Nonlinear batch effects often remain after traditional batch adjustment for larger datasets.
+For the larger datasets in our study, Confounded outperformed the other batch adjustment algorithms according to the batch classification metrics.
+However, Confounded also reduced the amount of true signal in those same datasets more than other algorithms.
+This may be able to be ameliorated by hyperparameter optimization,
+but in cases where batch effects and true signal are not independent, researchers may need to decide the tradeoff between how much true signal loss and remaining batch effects are acceptable.
 
@@
 
-Data integration
-
-Other types of data - already shown to work on images, microarray, \& rna seq
+At its root, batch correction is a data integration problem;
+data from multiple batches must have batch-specific confounding effects removed in order to be treated as one dataset.
+Confounded shows promise in removing traditional batch effects from microarray data in the Bladderbatch and GSE37199 datasets.
+It also effectively decreased artificial batch effects in image data and tissue-specific effects in RNA-Seq data.
+Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq dataset, or merging several large datasets measured under different conditions.
+We invite future researchers to utilize Confounded in any pipelines where data must be integrated from different sources, and to assess the effectiveness of doing so by measuring batch classifier accuracy.
 

### 2019-04-30 thesis.tex
Figure references, instructions for pdf to docx

@@
 \usepackage[square,numbers]{natbib}
+% \usepackage{natbib}
 \usepackage[nomarkers,nolists]{endfloat}
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow}
+	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
 	\caption{\textbf{Batch adjustment justification and steps}
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/network}
+	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
 	\caption{\textbf{Network architecture} of Confounded.
@@
 	\label{dual_loss}
-	L_dual = L_A - \lambda{}(L_D)
+	L_{dual} = L_A - \lambda{}(L_D)
 \end{equation}
@@
 A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
-$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_dual$.
+$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_{dual}$.
 
@@
 In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
-In each iteration, both $L_D$ and $L_dual$ were optimized.
+In each iteration, both $L_D$ and $L_{dual}$ were optimized.
 When optimizing $L_D$ (i.e. training the discriminator), the autoencoder's weights were frozen, and vice versa.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/training_loss}
+	\includegraphics[width=\columnwidth]{figures/rough/training_loss.png}
 	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/pca}
+	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
 	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/tsne}
+	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
 	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/mse}
+	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
 	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/mmd}
+	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
 	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
+	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
 	\caption{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
+	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
 	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/mnist}
+	\includegraphics[width=\columnwidth]{figures/rough/mnist.png}
 	\caption{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits} (a) before any adjustment,

### 2019-05-04 thesis.tex
Some changes in response to Dr. Piccolo's comments

@@
 
-\title{Removing nonlinear batch effects using adversarial deep neural networks}
+\title{Removing nonlinear batch effects from gene-expression data using adversarial deep neural networks}
 \author[1]{Jonathan B. Dayton}
@@
 	We also applied the model to remove the tissue signal from pan-cancer data.
-	Our software is available on GitHub to enable other scientists to remove confounding effects from their own data using our methodologies.
+	Our software is available at \href{https://github.com/jdayton3/Confounded}{https://github.com/jdayton3/Confounded} to enable other scientists to remove confounding effects from their own data using our methodologies.
 \end{abstract}
@@
 We used an adversarial autoencoder network to model and remove the batch effects.
-This network was structured in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder's output.
+We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder's output.
 By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data free of batch effects.
-The neural network was implemented in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
+We implemented the neural network in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
 All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.
@@
 We implemented the variational autoencoder \cite{louizos_variational_2015} described by Géron \cite{geron_hands-machine_2017}.
-Additional information can be found in the supplementary material.
+Additional information can be found in our supplementary material.
 
@@
 
-The discriminator was trained to determine the original batch of the autoencoder's output.
+We trained the discriminator to determine the original batch of the autoencoder's output.
 The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches.
@@
 
-Three loss functions were used to train the network.
-First, the autoencoder's loss ($L_A$) was calculated by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler divergence of the code layer).
-Second, the discriminator's loss ($L_D$) was trained on sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
-Finally, the autoencoder layers were also trained on a combination of the two previous losses,
+We trained the network using three loss functions.
+First, we calculated the autoencoder's loss ($L_A$) by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler divergence \cite{kullback_information_1951} of the code layer).
+Second, we calculated the discriminator's loss ($L_D$) as sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
+Finally, we also trained the autoencoder layers on a combination of the two previous losses,
 
@@
 A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
-$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_{dual}$.
+We did not optimize $L_A$ directly; instead we trained the autoencoder by optimizing $L_{dual}$.
 
@@
 
-In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
-In each iteration, both $L_D$ and $L_{dual}$ were optimized.
-When optimizing $L_D$ (i.e. training the discriminator), the autoencoder's weights were frozen, and vice versa.
+In all cases, we trained the network using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
+In each iteration, we optimized on both $L_D$ and $L_{dual}$.
+When optimizing $L_D$ (i.e. training the discriminator), we froze the autoencoder's weights, and vice versa.
+We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ CPU and 16 GB of RAM.
 For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.
@@
 It contains 60,000 training images and 10,000 test images.
+We used MNIST so we could visually assess how well the true signal (in this case, the shape and digit of each handwritten digit image) was preserved after batch adjustment.
 In order to use this dataset, we flattened each 28 by 28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit information.
 We limited our dataset to only the 10,000 test images.
-
-Using MNIST allowed us to visually assess how much true signal (in this case, the shape and digit of each handwritten digit image) was preserved after batch adjustment.
 Although convolutional layers are typically used when working with image data, we only used fully connected layers even for this image dataset.
@@
 We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
-The same random vectors were applied to each image in a batch.
-Finally, random noise was added to each image in order to prevent images in a batch from being overly similar to each other.
+We applied the same random vectors to each image in a batch.
+Finally, we added random noise to each image in order to prevent images in a batch from being overly similar to each other.
 
@@
 It contains expression values for 57 tissue samples from patients with and without bladder cancer across 5 unbalanced batches.
-Because it is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
+The dataset also has a cancer status (cancerous vs. normal tissue) column, which we used for ``true class'' classification, and a batch column.
+Because bladderbatch is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
 
@@
 The GSE37199 dataset contains microarray data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this data from the Open Science Framework that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
-This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced.
+We accessed a version of this data from \href{http://doi.org/10.17605/OSF.IO/SSK3T}{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).
 
@@
 However, we found that a strong nonlinear signal could still be identified in the data after adjustment.
-We used the same dataset that we tidied in this study to see if we could remove the nonlinear batch effect more completely.
+We used the same dataset that we tidied in this study (available at \href{https://osf.io/7xjdn/}{https://osf.io/7xjdn/}) to see if we could remove the nonlinear batch effect more completely.
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
@@
 We used four classifiers from the scikit-learn 0.19.1 Python library \cite{pedregosa_scikit-learn_2011} in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
-See Table \ref{tab:datasets} for details on which columns were used for training.
+Table \ref{tab:datasets} details which columns were used for training.
 
-We calculated the average of classification accuracies for four-fold cross-validation repeated 3 times.
+We calculated the average of classification accuracies for four-fold cross-validation repeated three times.
 We interpret lower accuracy for batch classification as meaning that the batch is removed more effectively.

### 2019-05-07 thesis.tex
More updates based on Dr. Piccolo's comments

@@
 \usepackage{hyperref}
+\usepackage[inline,shortlabels]{enumitem}
 \usepackage[square,numbers]{natbib}
@@
 	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
-	\caption{\textbf{Batch adjustment justification and steps}
-	(a) When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
-	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
-	(c) In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
-	(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
+	\caption[]{\textbf{Batch adjustment justification and steps}
+	\begin{enumerate*}[(a)]
+		\item When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
+		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
+		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
+		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
+	\end{enumerate*}
+	}
 	\label{fig:workflow}
@@
 they require that the input data only contains two batches, that the batches are sufficiently large, and that the batches are perfectly balanced.
-These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R SVA package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
+These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
 Additionally, the statistics these studies have used to validate results don't test whether complex interactions still remain in the data.
@@
 
-We implemented the variational autoencoder \cite{louizos_variational_2015} described by Géron \cite{geron_hands-machine_2017}.
+We implemented the variational autoencoder \cite{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
 Additional information can be found in our supplementary material.
@@
 It has been made available as an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package \cite{leek_sva_2017} to illustrate how to batch-adjust using ComBat.
-It contains expression values for 57 tissue samples from patients with and without bladder cancer across 5 unbalanced batches.
-The dataset also has a cancer status (cancerous vs. normal tissue) column, which we used for ``true class'' classification, and a batch column.
+It contains expression values for 57 tissue samples with and without bladder cancer across 5 unbalanced batches.
+The dataset has a cancer status (cancerous vs. normal tissue) column, which we used for ``true class'' classification, and a batch column.
 Because bladderbatch is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
@@
 
-The GSE37199 dataset contains microarray data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
+The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
 We accessed a version of this data from \href{http://doi.org/10.17605/OSF.IO/SSK3T}{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
-This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).
+It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
+We used this cancer status variable as the ``true class'' for classification.
+It has two types of batch variables: ``plate'' and ``centre.''
+We adjusted against the ``plate'' variable because it was more balanced than ``centre'' (with counts of $\{43, 50\}$ compared to $\{27, 66\}$).
+The GSE37199 dataset represents a slightly larger dataset than bladderbatch, with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).
 
@@
 The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
-In a previous study, we used this dataset and attempted to adjust for the effect of tissue type using ComBat \cite{dayton_classifying_2017-1}.
-However, we found that a strong nonlinear signal could still be identified in the data after adjustment.
-We used the same dataset that we tidied in this study (available at \href{https://osf.io/7xjdn/}{https://osf.io/7xjdn/}) to see if we could remove the nonlinear batch effect more completely.
+In a previous study \cite{dayton_classifying_2017-1}, we classified this dataset based on the presence or absence of mutations in several known cancer genes.
+We attempted to adjust for the confounding effect of cancer type using ComBat prior to classification.
+However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
+Here, we used the same version of the dataset that we tidied in this previous study (available at \href{https://osf.io/7xjdn/}{https://osf.io/7xjdn/}).
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
-We used this dataset as a way to test whether Confounded works on RNA-Seq data as well as to test whether we could remove batch effects that we know ComBat cannot.
+We used this dataset as a way to test whether Confounded works on RNA-Seq data and to test whether we could remove batch effects that ComBat cannot remove.
 
@@
 We used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
+There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \cite{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
+% TODO: add like 8 more methods
+Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
+For these reasons, we compared Confounded only to the two methods listed above.
+Future batch adjustment research may benefit from standardization of input formats, user interfaces, and validation datasets.
 
 \subsection{Statistics and Metrics}
-
-In order to determine how well Confounded removes batch effects compared to other methods, we calculated several metrics.
-First, we calculated mean squared error to see how different the output is from the input.
-Second, we calculated sample maximum mean discrepancy to see how differently distributed the batches are from each other.
-Third, we determined classification accuracy based on batch for several machine learning classifiers in order to determine whether batch can still be identified post-adjustment.
-Finally, we used the same classifiers to predict ``true class'' labels in order to determine how well important signal is maintained after adjustment.
 
@@
 
-A recent paper also used neural networks to remove batch effects \cite{shaham_removal_2017}.
+In a recent paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
 Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
-We calculated MMD to determine whether batches looked like they came from the same distribution after adjustment.
-We calculated MMD using the formula in this recent paper.
+We calculated MMD using the same formula as \citeauthor{shaham_removal_2017} to determine whether batches looked like they came from the same distribution after adjustment.
 For the kernel, we used the Gaussian kernel between two batches as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel} \cite{pedregosa_scikit-learn_2011}.
@@
 
-We used four classifiers from the scikit-learn 0.19.1 Python library \cite{pedregosa_scikit-learn_2011} in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
+In order to determine \begin{enumerate*}[(a)]
+	\item whether batch can still be identified post-adjustment and
+	\item how well important signal is maintained after adjustment,
+\end{enumerate*}
+we determined classification accuracy based on batch and ``true class'' labels using several machine learning classifiers.
+
+We used four classifiers from the scikit-learn 0.19.1 Python library \cite{pedregosa_scikit-learn_2011} in order to classify on batch and true class before and after adjustment: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
 Table \ref{tab:datasets} details which columns were used for training.
@@
 Therefore, given output data from the ideal batch adjuster, batch classification would be no better than random for any classification algorithm, and true class accuracy would be no lower than accuracy for the unadjusted data.
-We report these values in section \ref{sec:results}.
 
@@
 
-\subsection{Some batch effects not removed by other adjusters are removed by Confounded}
+\subsection{Confounded removes nonlinear batch effects that other adjusters miss}
 
 Various previous studies use differing methods to show qualitatively and quantitatively how well batch effects have been removed.
-We used several plots and metrics to compare adjusters, including PCA, t-SNE, MSE, MMD, and classifier batch prediction accuracy (for various classifiers).
+To compare Confounded to other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).
 
@@
 
-Confounded shows mixed success with the MSE and MMD metrics, but both values were typically lower for Confounded than for the scale adjuster and lower for ComBat than for any other adjuster we tested (See Figures \ref{fig:mse} and \ref{fig:mmd}).
+Confounded shows mixed success with the MSE and MMD metrics.
+With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets by a factor of $\frac{1}{2}$ to $\frac{1}{6}$ but scored drastically higher on the fourth dataset.
+With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets, this time by a factor of $\frac{1}{2}$ to $\frac{1}{10}$ and tied the scale adjuster on the fourth dataset.
+With both metrics, Confounded consistently performed somewhat more poorly than ComBat.
+(See Figures \ref{fig:mse} and \ref{fig:mmd}.)
 
@@
 	\includegraphics[width=\columnwidth]{figures/rough/mnist.png}
-	\caption{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits} (a) before any adjustment,
-	(b) with artificial noise added,
-	(c) adjusted for noise by ComBat, and
-	(d) adjusted for noise by Confounded.
+	\caption[]{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits}
+	\begin{enumerate*}[(a)]
+		\item before any adjustment,
+		\item with artificial noise added,
+		\item adjusted for noise by ComBat, and
+		\item adjusted for noise by Confounded.
+	\end{enumerate*}
 	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}

### 2019-05-08 thesis.tex
Updates from Overleaf

--- 2019-04-30 thesis.tex
@@
 \usepackage[square,numbers]{natbib}
-% \usepackage{natbib}
 \usepackage[nomarkers,nolists]{endfloat}
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow}
 	\caption{\textbf{Batch adjustment justification and steps}
@@
 which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
-Recently, researchers have experimented with discouraging neural networks from ,learning based on domain-specific information by giving them two objective functions:
+Recently, researchers have experimented with discouraging neural networks from learning based on domain-specific information by giving them two objective functions:
 1. to learn as much as possible about the input data and
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/network}
 	\caption{\textbf{Network architecture} of Confounded.
@@
 	\label{dual_loss}
-	L_{dual} = L_A - \lambda{}(L_D)
+	L_dual = L_A - \lambda{}(L_D)
 \end{equation}
@@
 A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
-$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_{dual}$.
+$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_dual$.
 
@@
 In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
-In each iteration, both $L_D$ and $L_{dual}$ were optimized.
+In each iteration, both $L_D$ and $L_dual$ were optimized.
 When optimizing $L_D$ (i.e. training the discriminator), the autoencoder's weights were frozen, and vice versa.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/training_loss.png}
+	\includegraphics[width=\columnwidth]{figures/rough/training_loss}
 	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/pca}
 	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/tsne}
 	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/mse}
 	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/mmd}
 	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
 	\caption{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
 	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/mnist.png}
+	\includegraphics[width=\columnwidth]{figures/rough/mnist}
 	\caption{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits} (a) before any adjustment,

### 2019-05-09 thesis.tex
Some updates by Dr. Piccolo

@@
 Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output to be as similar to the input as possible \citep{hinton_reducing_2006}.
-Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on dataset-specific confounding effects (e.g.
+Neural networks have historically decreased in effectiveness when working with data from multiple research domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on dataset-specific confounding effects (e.g.
 which researcher collected the data) instead of learning based on practically meaningful causal effects (e.g.

### 2019-05-09 thesis.tex
Update "biological data" -> "expression data"

@@
 \usepackage{authblk}
-\usepackage{hyperref}
+\usepackage{url}
 \usepackage[inline,shortlabels]{enumitem}
@@
 	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
-	In order to combine date produced under different conditions, we applied deep learning to remove these effects and combine molecular biological data sets.
+	In order to combine date produced under different conditions, we applied deep learning to remove these effects and combine gene expression datasets.
 	We created an adversarial variational autoencoder in order to remove confounding effects from the data while minimizing the amount of change to the input data.
-	We tested the model on artificially constructed data and commonly used real biological data sets and compared the output with the output from other common batch adjustment algorithms.
-	We also applied the model to remove the tissue signal from pan-cancer data.
-	Our software is available at \href{https://github.com/jdayton3/Confounded}{https://github.com/jdayton3/Confounded} to enable other scientists to remove confounding effects from their own data using our methodologies.
+	We tested the model on artificially constructed data and commonly used real gene expression datasets and compared the output with the output from other common batch adjustment algorithms.
+	We also applied the model to remove the cancer-type-specific signal from pan-cancer expression data.
+	Our software is available at \url{https://github.com/jdayton3/Confounded} to enable other scientists to remove confounding effects from their own data using our methodologies.
 \end{abstract}
@@
 
-When working with biological data, problems can arise from combining data from multiple sources.
+When working with gene expression data, problems can arise from combining data from multiple sources.
 For example, if two labs sequence and quantify the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
@@
 If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
-These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
-In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
+These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
+In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
 however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
@@
 
-Several methods exist for removing batch effects from biological datasets.
+Several methods exist for removing batch effects from gene expression datasets.
 Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
@@
 These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
-Additionally, the statistics these studies have used to validate results don't test whether complex interactions still remain in the data.
+Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.
 
@@
 This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
-Research has shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
+Research has shown that neural networks are effective in working with gene expression data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
 Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output to be as similar to the input as possible \citep{hinton_reducing_2006}.
@@
 The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
-A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
+A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the input data.
 We did not optimize $L_A$ directly; instead we trained the autoencoder by optimizing $L_{dual}$.
@@
 
-The bladderbatch dataset is a microarray transcriptome expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
+The bladderbatch dataset is a microarray transcriptomic expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
 It has been made available as an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package \cite{leek_sva_2017} to illustrate how to batch-adjust using ComBat.
@@
 The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this data from \href{http://doi.org/10.17605/OSF.IO/SSK3T}{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+We accessed a version of this data from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
 It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
@@
 However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
-Here, we used the same version of the dataset that we tidied in this previous study (available at \href{https://osf.io/7xjdn/}{https://osf.io/7xjdn/}).
+Here, we used the same version of the dataset that we tidied in this previous study (available at \url{https://osf.io/7xjdn/}).
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
@@
 data from multiple batches must have batch-specific confounding effects removed in order to be treated as one dataset.
-Confounded shows promise in removing traditional batch effects from microarray data in the Bladderbatch and GSE37199 datasets.
+Confounded shows promise in removing traditional batch effects from microarray expression data in the Bladderbatch and GSE37199 datasets.
 It also effectively decreased artificial batch effects in image data and tissue-specific effects in RNA-Seq data.
-Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq dataset, or merging several large datasets measured under different conditions.
+Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq datasets, or merging several large datasets measured under different conditions.
 We invite future researchers to utilize Confounded in any pipelines where data must be integrated from different sources, and to assess the effectiveness of doing so by measuring batch classifier accuracy.

### 2019-05-10 thesis.tex
Change title
Update abstract

@@
 \usepackage{authblk}
+\usepackage{hyperref}
 \usepackage{url}
@@
 
-\title{Removing nonlinear batch effects from gene-expression data using adversarial deep neural networks}
+\title{Adversarial deep neural networks effectively remove nonlinear batch effects from gene-expression data}
 \author[1]{Jonathan B. Dayton}
@@
 \begin{abstract}
-	Many biological studies, particularly studies involving the genetics of human disease, could discover items of greater statistical and biological significance if performed with larger data sets.
-	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
-	In order to combine date produced under different conditions, we applied deep learning to remove these effects and combine gene expression datasets.
-	We created an adversarial variational autoencoder in order to remove confounding effects from the data while minimizing the amount of change to the input data.
+	Gene expression data has many uses.
+	However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
+	For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
+	known covariates that are not of interest to the study may have strong effects,
+	and there may be large systemic effects when integrating multiple expression datasets.
+	Additionally, many of these confounding effects represent higher-order interactions and cannot be removed with linear methodologies.
+	We created Confounded in order to remove these effects from expression data
+	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
 	We tested the model on artificially constructed data and commonly used real gene expression datasets and compared the output with the output from other common batch adjustment algorithms.

### 2019-05-10 thesis.tex
A bunch of updates, but not all of them.

@@
 \begin{abstract}
-	Gene expression data has many uses.
+	Gene expression data is really cool.
 	However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
@@
 	Additionally, many of these confounding effects represent higher-order interactions and cannot be removed with linear methodologies.
-	We created Confounded in order to remove these effects from expression data
+	We created Confounded in order to remove these effects from expression data.
 	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
@@
 
-When working with gene expression data, problems can arise from combining data from multiple sources.
-For example, if two labs sequence and quantify the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
-% Is this ^ the best way to reference figures?
-If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
-These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
+Gene expression data can be applied in many ways to advance our understanding of medicine and biology.
+For example, expression data has been applied to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict the clinical outcome in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
+Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
+Because of the sensitive nature of this data (i.e.
+gene product levels in organisms may respond drastically to small environmental changes) and of the measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
+Three examples of this include
+\begin{enumerate*}[(a)]
+	\item batch effects, \label{it:batch}
+	\item known covariates, and \label{it:cov}
+	\item dataset-level effects. \label{it:int}
+\end{enumerate*}
+\ref{it:batch} Batch effects occur when expression data are generated in multiple runs or multiple batches, and slight systemic differences occur during the different runs, such as different technicians operating the machine or slight temperature differences in the room.
+Batch effects are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
 In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
 however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
-Systemic bias can be even more pronounced when using data collected from different tissue types and can skew results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.
+\ref{it:cov} Systemic bias can be even more pronounced when there are known covariates---for example, using data collected from different cancer types when cancer type is not of interest to the study \citep{dayton_classifying_2017-1}.
+\ref{it:int} Finally, though batch effects within a dataset collected all by one lab at one time may be somewhat decreased experimentally by careful replication of experimental conditions, this is no longer possible when integrating different datasets;
+however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
+Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see Figure \ref{fig:workflow}).
 
@@
 This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
-Research has shown that neural networks are effective in working with gene expression data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
+Research has shown that neural networks are effective in working with gene expression data; for example, neural networks have been applied
+to detect cancer and identify critical cancer genes \citep{danaee_deep_2016},
+to infer gene-expression values from just the values of a few ``landmark genes'' \citep{chen_gene_2016},
+to extract biologically relevant latent spaces in RNA-Seq data \citep{way_extracting_2017},
+to reduce the dimensions of single-cell RNA-Seq data \citep{lin_using_2017},
+to identify drug-repurposing targets using transcriptomic data \citep{aliper_deep_2016},
+and to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
 Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output to be as similar to the input as possible \citep{hinton_reducing_2006}.
@@
 2. to forget any patterns that help distinguish between domains \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
-This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``penalized'' for retaining class-specific patterns in the output.
+\citet{louizos_variational_2015} used this type of dual objective function with a variational autoencoder and successfully removed domain-based variability in credit score, financial savings, and hospital admittance datasets.
 
@@
 
+All our code has been made publicly available at \url{https://github.com/jdayton3/Confounded}, and all our data are available as described below.
+
 \subsection{Network Structure}
@@
 We implemented the neural network in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
-All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.
+All layers in the network were fully connected and all activation functions were Rectified Linear Units (ReLUs) \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.
 
@@
 The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches.
-In order to combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} and batch normalization \cite{ioffe_batch_2015} to each layer of the discriminator.
+In order to combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
 These additions seemed to reduce overfitting in the discriminator.
@@
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
-We used this dataset as a way to test whether Confounded works on RNA-Seq data and to test whether we could remove batch effects that ComBat cannot remove.
+We used this dataset as a way to test whether Confounded works on RNA-Seq data and to test whether we could remove confounding effects that ComBat cannot remove.
 
@@
 
-We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \cite{johnson_adjusting_2007}.
-We implemented the mean and scale adjuster in the R programming language \cite{r_core_team_r_2014}.
-We used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
-There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \cite{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
+We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \citep{johnson_adjusting_2007}.
+We implemented the mean and scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
+We used the ComBat implementation from the sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
+There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
 % TODO: add like 8 more methods
@@
 	\item whether batch can still be identified post-adjustment and
-	\item how well important signal is maintained after adjustment,
+	\item how well class-related signal is maintained after adjustment,
 \end{enumerate*}
@@
 Confounded shows mixed success with the MSE and MMD metrics.
-With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets by a factor of $\frac{1}{2}$ to $\frac{1}{6}$ but scored drastically higher on the fourth dataset.
-With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets, this time by a factor of $\frac{1}{2}$ to $\frac{1}{10}$ and tied the scale adjuster on the fourth dataset.
-With both metrics, Confounded consistently performed somewhat more poorly than ComBat.
+With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the fourth dataset, with score pairs of $\{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}$.
+With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the fourth dataset, with scores of $\{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}$.
+With both metrics, Confounded consistently performed somewhat worse than ComBat.
 (See Figures \ref{fig:mse} and \ref{fig:mmd}.)
@@
 
-\subsection{Important signal is still detectable after adjustment by Confounded}
+\subsection{Class-related signal is still detectable after adjustment by Confounded}
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/mnist.png}
-	\caption[]{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits}
+	\includegraphics[width=\columnwidth]{figures/final/mnist.png}
+	\caption[]{\textbf{MNIST handwritten digits}
 	\begin{enumerate*}[(a)]
@@
 		\item with artificial noise added,
+		\item adjusted for noise by the scale adjuster,
 		\item adjusted for noise by ComBat, and
@@
 	\end{enumerate*}
-	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
+
+	Although Confounded seems to remove more noise from the background, it struggles in some cases to accurately replicate the input data.}
 	\label{fig:mnist}
@@
 \section{Discussion} \label{sec:discussion}
+
+\subsection{Why use neural networks for batch adjustment?}
+
+\subsection{How can I tell how well batch adjustment worked?}
+
+\subsection{Which batch adjuster should I use?}
+
+\subsection{What limitations does Confounded have?}
+
+\subsection{What else might Confounded be used for?}
 

### 2019-05-14 thesis.tex
Another draft of discussion, start to format for thesis

@@
-\documentclass[notitlepage]{article}
+\documentclass{article}
+\usepackage{titling}
+
+\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-expression Data}
+\author{Jonathan B. Dayton}
+% \author[1]{Jonathan B. Dayton}
+% \author[1]{Stephen R. Piccolo}
+% \affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
+\date{2019}
+
 \usepackage{graphicx}
@@
 \usepackage[table]{xcolor}
-\linespread{1.5}
 \bibliographystyle{unsrtnat}
@@
 
-\title{Adversarial deep neural networks effectively remove nonlinear batch effects from gene-expression data}
-\author[1]{Jonathan B. Dayton}
-\author[1]{Stephen R. Piccolo}
-\affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
-\date{}
+\newgeometry{top=1in, bottom=1in, left=1.25in, right=1.25in}
+% Cover page
+\begin{titlingpage}
+	\begin{center}
+		\thetitle\\
+		\vfill
+		\theauthor\\
+		\vfill
+		A thesis submitted to the faculty of\\
+		Brigham Young University\\
+		in partial fulfillment of the requirements for the degree of\\
+		\vspace{1em}
+		Master of Science\\
+		\vspace{3em}
+		Stephen R. Piccolo, Chair\\
+		Mark Clement\\
+		David Wingate\\
+		\vspace{4em}
+		Department of Biology\\
+		\vspace{1em}
+		Brigham Young University
+		\vfill
+		Copyright \textcopyright \ \thedate \ \theauthor\\
+		\vspace{1em}
+		All Rights Reserved
+	\end{center}
 
+\end{titlingpage}
+
+% Title page
+\begin{titlingpage}
+	\begin{center}
+		ABSTRACT
+	\end{center}
 	\maketitle
-
 	\begin{abstract}
-	Gene expression data is really cool.
+		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
 		However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
@@
 	\end{abstract}
+\end{titlingpage}
+
+\restoregeometry
+\linespread{2}
 
@@
 
-\subsection{Why use neural networks for batch adjustment?}
+\textbf{Why should I use a neural network for batch adjustment?}
+The process of measuring data typically leaves confounding effects, especially when the data or measurement process involves living things.
+Imagine an experiment where two researchers measure the length of caterpillars.
+If one researcher is left-eye dominant and the other is right-eye dominant, they may each see lengths as being a millimeter different.
+If both researchers use the same wooden ruler at different relative humidities, the measurements may come out differently due to the wood swelling or due to the caterpillars reactions to humidity.
+These same types of problems exist when measuring expression levels, but to more of an extreme---in addition to the target values and measuring equipment changing slightly due to systemic factors, the target values also change in response to each other.
+Each of the 20,000 transcript levels may influence or be influenced by other transcript levels.
+These cascading network-like effects are extremely likely to have nonlinear components, and our research shows that these nonlinear effects do indeed exist in gene expression data, thus rendering linear batch correction methods insufficient.
+Therefore, some form of nonlinear adjustment must be used in order to correct for real-world confounding effects.
+Rather than individually model each of infinitely many possible nonlinear interactions to see which represent confounders, we can use a neural network to both approximate and remove the confounders, since neural networks are proven to be universal function approximators \citep{csaji_approximation_2001}.
 
-\subsection{How can I tell how well batch adjustment worked?}
+\textbf{How can I tell how well batch adjustment worked?}
+Although the metrics and figures that have been used in the past to validate batch adjustment (such as PCA, MSE, and MMD) represent how well linear effects have been removed, they cannot completely display whether two batches are distinguishable from one another.
+Machine learning algorithms are designed specifically to tease out patterns in data that may distinguish one group from another.
+We suggest to users of batch correction software that they use machine learning classification accuracy before and after correction in order to determine the degree of batch removal.
+We also suggest to researchers in the field of batch correction that classification accuracy be used as a metric in validating their software.
+Specifically, the Random Forests algorithm \citep{tin_kam_ho_random_1995} seems to work very well and runs relatively quickly on gene expression data.
 
-\subsection{Which batch adjuster should I use?}
+\textbf{Which batch adjuster should I use?}
+In our testing, ComBat did very well with small ($n < 100$) datasets, even with removing any identifiable nonlinear effects.
+However, Confounded outperformed ComBat on the larger datasets according to the batch classification accuracy metric.
+In addition to dataset size, researchers selecting a batch adjustment algorithm should consider how important it is for them to accurately replicate their input data.
+Such researchers can adjust Confounded's $\lambda$ parameter in order to balance the tradeoff of removing batch and matching the inputs.
 
-\subsection{What limitations does Confounded have?}
+\textbf{What limitations does Confounded have?}
+\begin{enumerate*}[(a)]
+	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in Figure \ref{fig:mnist}).
+	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
+	Similar research with variational autoencoders and gene expression data may yield improved reconstruction losses and decrease the blurring effect.
+	\item Confounded takes a long time to run in comparison with ComBat and other linear adjusters.
+	Although we acknowledge this as a downfall of many types of machine learning and of neural network in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
+	\item It can be difficult to identify the optimal network structure and parameter set for a neural network.
+	Though this is the case for many applications of neural networks, we feel that Confounded's default structure worked well in our testing and that it will suffice for most batch correction applications.
+	\item Neural networks usually perform better when given large amounts of data, and traditional batch datasets typically have very few samples.
+	We did find that ComBat may outperform Confounded on many smaller, more traditional datasets, but that Confounded performs better on larger datasets.
+	However, Confounded did perform reasonably well with the smaller datasets and appeared to avoid overfitting.
+	In cases where ComBat is unable to completely remove batch effects in a small dataset, Confounded may be a viable replacement method.
+\end{enumerate*}
 
-\subsection{What else might Confounded be used for?}
-
-\subsection{Classification accuracy quantifies batch removal more accurately than MSE and MMD}
-
-Our MSE \ref{fig:mse} and MMD \ref{fig:mmd} metrics both show that ComBat performed better than Confounded for the MNIST dataset, but the effects left after ComBat finished are still identifiable by Random Forests with 99.9\% accuracy (Table \ref{tab:batch}).
-Additionally, previous work \cite{dayton_classifying_2017-1} has shown that PCA can be misleading when determining whether batch effects have been sufficiently removed from a dataset.
-Each of these methods seems to focus on linear separability.
-
-Many machine learning classification algorithms are able to identify both linear and nonlinear interactions between variables.
-A decrease in classifier accuracy represents that nonlinear batch effects have been decreased.
-Random Forests often has high batch classification accuracy after ComBat-adjustment even when other classifiers are unable to identify differences between batches, whereas SVC often performs no better than baseline (see Figure \ref{fig:batch} and Table \ref{tab:batch}).
-In addition, ComBat often excels at hiding batch effects from Naive Bayes.
-We suggest that it may be important to use several classifiers to identify whether batch effects remain after correction.
-
-We encourage future researchers who study batch correction or who use it in their pipelines to use classifier accuracy to determine the degree to which batch effects still remain in their data.
-This is especially important to take into account when analyses will include some type of machine learning that can detect nonlinear effects.
-
-\subsection{Dataset should be considered when applying batch adjustment}
-
-Nonlinear batch effects often remain after traditional batch adjustment for larger datasets.
-For the larger datasets in our study, Confounded outperformed the other batch adjustment algorithms according to the batch classification metrics.
-However, Confounded also reduced the amount of true signal in those same datasets more than other algorithms.
-This may be able to be ameliorated by hyperparameter optimization,
-but in cases where batch effects and true signal are not independent, researchers may need to decide the tradeoff between how much true signal loss and remaining batch effects are acceptable.
-
-\subsection{Confounded may have other uses}
-
-At its root, batch correction is a data integration problem;
+\textbf{What else might Confounded be used for?}
+At its root, batch correction is a data integration problem:
 data from multiple batches must have batch-specific confounding effects removed in order to be treated as one dataset.
@@
 Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq datasets, or merging several large datasets measured under different conditions.
-We invite future researchers to utilize Confounded in any pipelines where data must be integrated from different sources, and to assess the effectiveness of doing so by measuring batch classifier accuracy.
+
+Confounded, and adversarial autoencoders in general, show promise as a valuable way to remove confounding biases from expression datasets.
+Such methods will enable researchers access to larger datasets, therefore increasing the scope of analyses and furthering science as a whole.
 

### 2019-05-14 thesis.tex
Finish updating the thesis styling

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 \usepackage{titling}
 
-\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-expression Data}
+\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
 \author{Jonathan B. Dayton}
@@
 
+% Times New Roman for the thesis submission
+% \usepackage{mathptmx}
 \usepackage{graphicx}
+\usepackage{setspace}
 \usepackage{authblk}
+\usepackage{url}
 \usepackage{hyperref}
-\usepackage{url}
 \usepackage[inline,shortlabels]{enumitem}
@@
 % \usepackage{natbib}
-\usepackage[nomarkers,nolists]{endfloat}
+\usepackage[nomarkers,nolists,tabhead,fighead]{endfloat}
 \renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
-\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
+\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
 \usepackage{csvsimple}
@@
 \usepackage[table]{xcolor}
+\usepackage{changepage}
 \bibliographystyle{unsrtnat}
@@
 
-\newgeometry{top=1in, bottom=1in, left=1.25in, right=1.25in}
+\pagenumbering{roman}
 % Cover page
@@
 	\begin{center}
-		\thetitle\\
+		\begin{adjustwidth}{+0.5in}{+0.5in}
+			\begin{center}
+				\doublespacing \thetitle\\ \singlespacing
+			\end{center}
+		\end{adjustwidth}
 		\vfill
@@
 		Master of Science\\
-		\vspace{3em}
+		\vfill
 		Stephen R. Piccolo, Chair\\
@@
 		David Wingate\\
-		\vspace{4em}
+		\vfill
 		Department of Biology\\
@@
 
-% Title page
+% Title / abstract page
 \begin{titlingpage}
 	\begin{center}
-		ABSTRACT
+		ABSTRACT\\
+		\vspace{1em}
+		\begin{adjustwidth}{+0.5in}{+0.5in}
+			\begin{center}
+				\thetitle\\
 			\end{center}
-	\maketitle
-	\begin{abstract}
+		\end{adjustwidth}
+		\vspace{1em}
+		\theauthor\\
+		Department of Biology, BYU\\
+		Master of Science
+	\end{center}
+	% \begin{abstract}
 		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
@@
 		Our software is available at \url{https://github.com/jdayton3/Confounded} to enable other scientists to remove confounding effects from their own data using our methodologies.
-	\end{abstract}
+	% \end{abstract}
+	\vfill
+	\noindent
+	Keywords: batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder
 \end{titlingpage}
 
-\restoregeometry
-\linespread{2}
+\doublespacing
 
+\begin{titlingpage}
+	\begin{center}
+		ACKNOWLEDGMENTS\\
+		This page is optional.
+	\end{center}
+\end{titlingpage}
+
+\setcounter{page}{4}
+\tableofcontents
+\listoftables
+\listoffigures
+\pagebreak
+
+\pagenumbering{arabic}
 \section{Background} \label{sec:background}
@@
 	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
-	\caption[]{\textbf{Batch adjustment justification and steps}
+	\caption[Batch adjustment steps]{\textbf{Batch adjustment justification and steps}
 	\begin{enumerate*}[(a)]
@@
 	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
-	\caption{\textbf{Network architecture} of Confounded.
+	\caption[Network architecture of Confounded]{\textbf{Network architecture} of Confounded.
 	Data with batch effects (represented by different colors) are input into an autoencoder.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/rough/training_loss.png}
-	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
+	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
+	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded.
 	Over the course of training, the autoencoder more faithfully replicates the input data.
-	The autoencoder also seems to introduce noise (see around iteration 3100) in response to the discriminator's slight improvements.}
+	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
 	\label{fig:training_loss}
@@
 	\input{figures/final/datasets.tex}
-	\caption{\textbf{Dataset information} for each dataset used.}
+	\caption[Dataset information]{\textbf{Dataset information} for each dataset used.}
 	\label{tab:datasets}
@@
 	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
-	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
+	\caption[Principal components analysis (PCA)]{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
 	None of the datasets appear to be linearly separable.
@@
 	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
-	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
+	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
 	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
@@
 Confounded shows mixed success with the MSE and MMD metrics.
-With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the fourth dataset, with score pairs of $\{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}$.
-With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the fourth dataset, with scores of $\{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}$.
+With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the fourth dataset, with score pairs of \{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}.
+With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the fourth dataset, with scores of \{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}.
 With both metrics, Confounded consistently performed somewhat worse than ComBat.
@@
 	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
-	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
+	\caption[Mean squared error]{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
 	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
@@
 	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
-	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
+	\caption[Maximum mean discrepancy]{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
 	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
@@
 	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
-	\caption{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
+	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
 	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
@@
 	\input{figures/final/batch.tex}
-	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
+	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} for several datasets and adjusters.
 	The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
@@
 	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
+	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
 	See also Table \ref{tab:true_class}.}
@@
 	\input{figures/final/true_class.tex}
-	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
+	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters.
 	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
@@
 	\includegraphics[width=\columnwidth]{figures/final/mnist.png}
-	\caption[]{\textbf{MNIST handwritten digits}
+	\caption[MNIST handwritten digits]{\textbf{MNIST handwritten digits}
 	\begin{enumerate*}[(a)]
@@
 	\end{enumerate*}
-
 	Although Confounded seems to remove more noise from the background, it struggles in some cases to accurately replicate the input data.}

### 2019-05-15 thesis.tex
Get styling and bookmarks in line with department requirements

@@
 \usepackage{url}
-\usepackage{hyperref}
 \usepackage[inline,shortlabels]{enumitem}
@@
 % \usepackage{natbib}
-\usepackage[nomarkers,nolists,tabhead,fighead]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
+\usepackage{floatrow}
+\floatsetup[table]{capposition=top}
+\usepackage[nomarkers,nolists,tabhead,fighead,tablesfirst]{endfloat}
+% \renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
 \usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
@@
 \usepackage{changepage}
+\usepackage{tocloft}
+\usepackage[bookmarksnumbered,bookmarksopen=true,bookmarks=true]{hyperref}
+\usepackage{bookmark}
+
+% stuff for restyling the table of contents
+\renewcommand\cftsecfont{\mdseries}
+\renewcommand\cftsecpagefont{\mdseries}
+\renewcommand{\contentsname}{\begin{center}\normalsize{\textnormal{TABLE OF CONTENTS}}\end{center}}
+\renewcommand{\listtablename}{\begin{center}\normalsize{\textnormal{LIST OF TABLES}}\end{center}}
+\renewcommand{\listfigurename}{\begin{center}\normalsize{\textnormal{LIST OF FIGURES}}\end{center}}
+
 \bibliographystyle{unsrtnat}
@@
 \pagenumbering{roman}
-% Cover page
 \begin{titlingpage}
+	\setcounter{page}{1}
+	\addcontentsline{toc}{section}{Title Page}
 	\begin{center}
@@
 \begin{titlingpage}
+	\setcounter{page}{2}
+	\phantomsection
+	\addcontentsline{toc}{section}{Abstract}
 	\begin{center}
@@
 		and there may be large systemic effects when integrating multiple expression datasets.
-		Additionally, many of these confounding effects represent higher-order interactions and cannot be removed with linear methodologies.
-		We created Confounded in order to remove these effects from expression data.
+		Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
+		We created Confounded to remove these effects from expression data.
 		Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
-		We tested the model on artificially constructed data and commonly used real gene expression datasets and compared the output with the output from other common batch adjustment algorithms.
-		We also applied the model to remove the cancer-type-specific signal from pan-cancer expression data.
-		Our software is available at \url{https://github.com/jdayton3/Confounded} to enable other scientists to remove confounding effects from their own data using our methodologies.
+		We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
+		We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
+		Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
 	% \end{abstract}
@@
 \begin{titlingpage}
+	\setcounter{page}{3}
+	\phantomsection
+	\addcontentsline{toc}{section}{Acknowledgments}
 	\begin{center}
 		ACKNOWLEDGMENTS\\
-		This page is optional.
+		TODO: Fill this out.
 	\end{center}
@@
 \setcounter{page}{4}
+\phantomsection
+\addcontentsline{toc}{section}{Table of Contents}
 \tableofcontents
+\newpage
+\phantomsection
+\addcontentsline{toc}{section}{List of Tables}
 \listoftables
+\newpage
+\phantomsection
+\addcontentsline{toc}{section}{List of Figures}
 \listoffigures
-\pagebreak
+\newpage
+
+% make bookmarks for Figure and Tables sections
+\renewcommand{\figuresection}{\pdfbookmark[0]{Figures}{figure}Figures}
+\renewcommand{\tablesection}{\pdfbookmark[0]{Tables}{table}Tables}
 
@@
 Because of the sensitive nature of this data (i.e.
-gene product levels in organisms may respond drastically to small environmental changes) and of the measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
+gene product levels in organisms may respond drastically to small environmental changes) and of the equipment used to quantify expression levels, expression data often contains unwanted confounding effects that can skew analysis.
 Three examples of this include
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:workflow}}{workflow}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
-	\caption[Batch adjustment steps]{\textbf{Batch adjustment justification and steps}
+	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
 	\begin{enumerate*}[(a)]
@@
 ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
-Since both of these methods use linear adjustments to remove batch effects, they cannot account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
+Since both of these methods use linear methods to model batch effects, they are not designed to account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
 As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:network}}{network}
 	\centering
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:training_loss}}{training_loss}
 	\centering
@@
 	\input{figures/final/datasets.tex}
-	\caption[Dataset information]{\textbf{Dataset information} for each dataset used.}
+	\caption[Dataset information]{
+		\pdfbookmark[1]{Table \ref{tab:datasets}}{datasets}
+		\textbf{Dataset information} for each dataset used.
+	}
 	\label{tab:datasets}
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:pca}}{pca}
 	\centering
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:tsne}}{tsne}
 	\centering
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:mse}}{mse}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
-	\caption[Mean squared error]{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
+	\caption[Mean squared error (MSE)]{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
 	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:mmd}}{mmd}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
-	\caption[Maximum mean discrepancy]{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
+	\caption[Maximum mean discrepancy (MMD)]{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
 	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:batch}}{batch}
 	\centering
@@
 	\input{figures/final/batch.tex}
-	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} for several datasets and adjusters.
+	\caption[Batch classification accuracy]{
+		\pdfbookmark[1]{Table \ref{tab:batch}}{batch}
+		\textbf{Batch classification accuracy} for several datasets and adjusters.
 		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
-	See also Figure \ref{fig:batch}.}
+		See also Figure \ref{fig:batch}.
+	}
 	\label{tab:batch}
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:true_class}}{true_class}
 	\centering
@@
 	\input{figures/final/true_class.tex}
-	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters.
+	\caption[True class classification accuracy]{
+		\pdfbookmark[1]{Table \ref{tab:true_class}}{true_class}
+		\textbf{True class classification accuracy} for several datasets and adjusters.
 		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
-	See also Figure \ref{fig:true_class}.}
+		See also Figure \ref{fig:true_class}.
+	}
 	\label{tab:true_class}
@@
 \begin{figure}
+	\pdfbookmark[1]{Figure \ref{fig:mnist}}{asdf}
 	\centering
@@
 
+\newpage
+\phantomsection
+\addcontentsline{toc}{section}{References}
 \bibliography{references}

### 2019-05-16 thesis.tex
Address some of Dr. Piccolo's comments

@@
 		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
-		However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
+		However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
 		For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
@@
 Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
-Because of the sensitive nature of this data (i.e.
+Because of the sensitive nature of these data (i.e.
 gene product levels in organisms may respond drastically to small environmental changes) and of the equipment used to quantify expression levels, expression data often contains unwanted confounding effects that can skew analysis.
@@
 ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
-Since both of these methods use linear methods to model batch effects, they are not designed to account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
-As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
-Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
+Since both of these methods use linear methods to model confounding effects, they are not designed to account for nonlinear effects, such as the cascading changes in gene pathways in response to environmental stressors (e.g.
+gene X transcripts increase and gene Y transcripts decrease in response to rising temperature, and since both are now past a certain threshold, gene Z transcripts also increase, and so on).
+As machine learning becomes more common in biological research, these nonlinear confounding effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
+For example, advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
 These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
 Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
-However, these recent programs have several items that complicate their usability on many real-world datasets:
-they require that the input data only contains two batches, that the batches are sufficiently large, and that the batches are perfectly balanced.
-These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
+However, several limitations complicate their usability on many real-world datasets:
+they require that the input data only contains two batches, that the batches are sufficiently large (we received errors when testing with a subset of MNIST where n=100), and that the batches are perfectly balanced.
+These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
 Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.
@@
 
-In this study, we present Confounded, an adversarial autoencoder that identifies and removes effects between different batches.
-We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
-We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess the extent to which batch effects remain after adjustment using various classification algorithms.
+In this study, we present Confounded, an adversarial autoencoder that identifies and removes confounding effects.
+We test the hypothesis that using an adversarial neural network can correct for confounding effects more completely than previous tools do.
+We also explore the extent to which confounders still remain in different datasets after adjustment with various algorithms, and we present a framework to assess the extent to which confounding effects remain after adjustment using various classification algorithms.
 
@@
 
-We used an adversarial autoencoder network to model and remove the batch effects.
-We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder's output.
-By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data free of batch effects.
+We used an adversarial autoencoder network to model and remove the confounding effects.
+We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining confounding effects in the autoencoder's output.
+By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data with confounding effects minimized.
 We implemented the neural network in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
@@
 
-We implemented the variational autoencoder \cite{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
-Additional information can be found in our supplementary material.
+We implemented the variational autoencoder \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
+This network has 2 hidden encoding layers and 2 decoding layers, each of size 500.
+The code size is 20.
+Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015}.
+It is trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).
 
@@
 We trained the network using three loss functions.
-First, we calculated the autoencoder's loss ($L_A$) by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler divergence \cite{kullback_information_1951} of the code layer).
+First, we calculated the autoencoder's loss ($L_A$) by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler (KL) divergence \cite{kullback_information_1951} of the code layer).
 Second, we calculated the discriminator's loss ($L_D$) as sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
@@
 
-The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
-A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the input data.
+The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing confounding effects.
+A higher $\lambda$ value indicates that the network should remove confounding effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the input data.
 We did not optimize $L_A$ directly; instead we trained the autoencoder by optimizing $L_{dual}$.
@@
 
-In all cases, we trained the network using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
+In all cases, we trained the network using the Adam Optimizer \citep{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
 In each iteration, we optimized on both $L_D$ and $L_{dual}$.
@@
 	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
-	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded.
+	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded on the MNIST dataset.
 	Over the course of training, the autoencoder more faithfully replicates the input data.
@@
 
-Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear batch effects.
+Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear confonding effects.
 To do so, we wrote a Python script to take the MNIST data in, apply a nonlinear effect, and output the adjusted data.
@@
 The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this data from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+We accessed a version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
 It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
@@
 
-We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \citep{johnson_adjusting_2007}.
-We implemented the mean and scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
-We used the ComBat implementation from the sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
+We compared our method to two other batch adjusters: a scale adjuster and ComBat \citep{johnson_adjusting_2007}.
+We implemented the scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
+It adjusts the data by linearly expanding or contracting each batch so all batches have the same range.
+We used the ComBat implementation from the R sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
+We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
 There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
@@
 
-\subsection{Confounded removes nonlinear batch effects that other adjusters miss}
+The scaling method performed consistently worse across the qualitative and quantitative evaluations that we performed;
+ComBat and Confounded performed relatively well overall, but each of these algorithms excelled in different types of scenarios, which we illustrate below.
 
-Various previous studies use differing methods to show qualitatively and quantitatively how well batch effects have been removed.
+\subsection{Confounded removes nonlinear confounding effects that other adjusters miss}
+
+Various previous studies use differing methods to show qualitatively and quantitatively how well confounding effects have been removed.
 To compare Confounded to other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).
@@
 Confounded shows mixed success with the MSE and MMD metrics.
-With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the fourth dataset, with score pairs of \{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}.
-With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the fourth dataset, with scores of \{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}.
+With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the MNIST dataset, with scores (listed as: scale, Confounded) of \{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}.
+With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the TCGA dataset, with scores of \{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}.
 With both metrics, Confounded consistently performed somewhat worse than ComBat.
@@
 This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
-Although Naive Bayes is no longer able to identify batch effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
+Although Naive Bayes is no longer able to identify confounding effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
 In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
@@
 69.8\%) while the other adjusters' accuracies remain above baseline.
-However, preliminary testing indicates that fine-tuning Confounded's parameters to run on these datasets rather than using the default settings may greatly improve Confounded's post-adjustment true class accuracy.
-% TODO: include this^
+However, the particular set of parameters that we used in Confounded are likely not optimal for every dataset.
+Additional tuning may improve the performance metrics.
 
@@
 	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
+A higher accuracy after adjustment is desired because it represents that the adjuster has not destroyed the true class signal.
 	See also Table \ref{tab:true_class}.}
@@
 	However, Confounded did perform reasonably well with the smaller datasets and appeared to avoid overfitting.
-	In cases where ComBat is unable to completely remove batch effects in a small dataset, Confounded may be a viable replacement method.
+	In cases where ComBat is unable to completely remove confounding effects in a small dataset, Confounded may be a viable replacement method.
 \end{enumerate*}
@@
 Confounded shows promise in removing traditional batch effects from microarray expression data in the Bladderbatch and GSE37199 datasets.
-It also effectively decreased artificial batch effects in image data and tissue-specific effects in RNA-Seq data.
+It also effectively decreased artificial batch effects in image data and cancer-type-specific confounding effects in RNA-Seq data.
 Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq datasets, or merging several large datasets measured under different conditions.

### 2019-05-29 thesis.tex
Update thesis styling, figures, and notes from practicing

@@
 \title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
-\author{Jonathan B. Dayton}
+\author{Jonathan Bryan Dayton}
 % \author[1]{Jonathan B. Dayton}
@@
 \usepackage{tocloft}
-\usepackage[bookmarksnumbered,bookmarksopen=true,bookmarks=true]{hyperref}
+\usepackage[bookmarksnumbered,bookmarksopen=true,bookmarks=true,hidelinks,pdfpagelabels=true]{hyperref}
 \usepackage{bookmark}
+
+% Make all figures and tables hang towards the top
+\makeatletter
+  \providecommand*\setfloatlocations[2]{\@namedef{fps@#1}{#2}}
+\makeatother
+\setfloatlocations{figure}{htbp}
+\setfloatlocations{table}{htbp}
+
+% Prevent captions from setting stuff 2x and therefore making 2 pdfbookmarks
+\usepackage{caption}
+\captionsetup{singlelinecheck=off}
+
+% Some manual \pdfbookmarks are too low. This should raise them.
+\makeatletter
+ \newcommand{\pdfbookmarkraised}[3]{\Hy@raisedlink{\pdfbookmark[#1]{#2}{#3}}}
+\makeatother
+
+% TODO: change `\nolinkurl`s to `\url`s
+% Add dots to sections in TOC
+\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}
 
@@
 	\addcontentsline{toc}{section}{Title Page}
-	\begin{center}
 	\begin{adjustwidth}{+0.5in}{+0.5in}
@@
 	\end{adjustwidth}
+	\centering{
 		\vfill
@@
 		All Rights Reserved
-	\end{center}
+	\par}
 
@@
 	% \begin{abstract}
+		\paragraph{} % adds indentation
 		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
@@
 		We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
-		Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
+		Our software is publicly available at \nolinkurl{https://github.com/jdayton3/Confounded}.
 	% \end{abstract}
@@
 	\noindent
-	Keywords: batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder
+	Keywords:\@ batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder % the \@ removes the double "\nonfrenchspacing" spaces after the colon, which is required for thesis submission
 \end{titlingpage}
@@
 \phantomsection
-\addcontentsline{toc}{section}{List of Tables}
-\listoftables
+\addcontentsline{toc}{section}{List of \tablename{s}}
+{% Change table refs "1", "2" -> "Table 1", "Table 2" in LOT
+\let\oldnumberline\numberline%
+\renewcommand{\numberline}{\tablename~\oldnumberline}%
+\listoftables%
+}
 \newpage
 \phantomsection
-\addcontentsline{toc}{section}{List of Figures}
-\listoffigures
+\addcontentsline{toc}{section}{List of \figurename{s}}
+{% See comment on list of tables
+\let\oldnumberline\numberline%
+\renewcommand{\numberline}{\figurename~\oldnumberline}%
+\listoffigures%
+}
 \newpage
 
-% make bookmarks for Figure and Tables sections
-\renewcommand{\figuresection}{\pdfbookmark[0]{Figures}{figure}Figures}
-\renewcommand{\tablesection}{\pdfbookmark[0]{Tables}{table}Tables}
+% make bookmarks for figure and tables sections
+\renewcommand{\figuresection}{\pdfbookmarkraised{0}{\figurename{s}}{figure}\figurename{s}}
+\renewcommand{\tablesection}{\pdfbookmarkraised{0}{\tablename{s}}{table}\tablename{s}}
 
@@
 however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
-Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see Figure \ref{fig:workflow}).
+Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see \figurename{} \ref{fig:workflow}).
 
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:workflow}}{workflow}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:workflow}: Batch adjustment justification and steps}{fig:workflow}
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
+	% TODO I changed the width so "Figures" would be on the same page as the 1st fig per Gentri's request. Change this back eventually
+	\includegraphics[width=0.85\columnwidth]{figures/final/adjuster_workflow.pdf}
 	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
@@
 
-All our code has been made publicly available at \url{https://github.com/jdayton3/Confounded}, and all our data are available as described below.
+All our code has been made publicly available at \nolinkurl{https://github.com/jdayton3/Confounded}, and all our data are available as described below.
 
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:network}}{network}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:network}: Network architecture of Confounded}{fig:network}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
-	\caption[Network architecture of Confounded]{\textbf{Network architecture} of Confounded.
+	\caption[Network architecture of Confounded]{\textbf{Network architecture of Confounded.}
 	Data with batch effects (represented by different colors) are input into an autoencoder.
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:training_loss}}{training_loss}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:training_loss}: Autoencoder and discriminator loss}{fig:training_loss}
 	\centering
@@
 	\caption[Dataset information]{
-		\pdfbookmark[1]{Table \ref{tab:datasets}}{datasets}
+		\pdfbookmarkraised{1}{Table \ref{tab:datasets}: Dataset information}{tab:datasets}
 		\textbf{Dataset information} for each dataset used.
@@
 
-\paragraph{Synthetic Batch Effects}
+% \paragraph{Synthetic Batch Effects} % TODO: uncomment this after submitting thesis
 
@@
 The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+We accessed a version of this dataset from \nolinkurl{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
 It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
@@
 However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
-Here, we used the same version of the dataset that we tidied in this previous study (available at \url{https://osf.io/7xjdn/}).
+Here, we used the same version of the dataset that we tidied in this previous study (available at \nolinkurl{https://osf.io/7xjdn/}).
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
@@
 
+In this study, we created Confounded, an adversarial variational autoencoder neural network, to remove nonlinear batch and confounding effects from expression data that may not be accounted for by traditional linear methods.
+We compared Confounded to a scaling method and to ComBat \citep{johnson_adjusting_2007} using various metrics.
 The scaling method performed consistently worse across the qualitative and quantitative evaluations that we performed;
@@
 
-Various previous studies use differing methods to show qualitatively and quantitatively how well confounding effects have been removed.
 To compare Confounded to other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).
 
-PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods for the GSE37199 dataset (see Figures \ref{fig:pca} and \ref{fig:tsne}).
+PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods for the GSE37199 dataset (see \figurename{s} \ref{fig:pca} and \ref{fig:tsne}).
 However, previous research has shown that these plots are not completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:pca}}{pca}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:pca}: Principal components analysis (PCA)}{fig:pca}
 	\centering
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:tsne}}{tsne}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:tsne}: T-distributed Stochastic Neighbor Embedding (t-SNE)}{fig:tsne}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
-	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
+	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding (t-SNE)} plot for the GSE37199 dataset before and after adjustment with several algorithms.
 	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
@@
 Confounded shows mixed success with the MSE and MMD metrics.
-With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the MNIST dataset, with scores (listed as: scale, Confounded) of \{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}.
-With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the TCGA dataset, with scores of \{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}.
+With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the MNIST dataset, with scores listed in Table \ref{tab:mse} (see also \figurename{} \ref{fig:mse}).
+With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the TCGA dataset, with scores listed in Table \ref{tab:mmd} (see also \figurename{} \ref{fig:mmd}).
 With both metrics, Confounded consistently performed somewhat worse than ComBat.
-(See Figures \ref{fig:mse} and \ref{fig:mmd}.)
 
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:mse}}{mse}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mse}: Mean squared error (MSE)}{fig:mse}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
-	\caption[Mean squared error (MSE)]{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
+	\caption[Mean squared error (MSE)]{
+		\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
 		Lower MSE represents that the adjuster has more faithfully reproduced the input data.
 		MSE for unadjusted data will always be 0 because the input data is identical to the output data.
-	Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
+		Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
+		See also Table \ref{tab:mse}.
+	}
 	\label{fig:mse}
 \end{figure}
+\begin{table}
+	\input{figures/final/mse.tex}
+	\caption[Mean squared error (MSE)]{
+		\pdfbookmarkraised{1}{Table \ref{tab:mse}: Mean squared error (MSE)}{tab:mse}
+		\textbf{Mean squared error (MSE)} of the unadjusted input data compared to the data output by the given adjusters.
+		Lower MSE indicates that the output has changed less from the input.
+		See also \figurename{} \ref{fig:mse}.
+	}
+	\label{tab:mse}
+\end{table}
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:mmd}}{mmd}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mmd}: Maximum mean discrepancy (MMD)}{fig:mmd}
 	\centering
 	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
-	\caption[Maximum mean discrepancy (MMD)]{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
+	\caption[Maximum mean discrepancy (MMD)]{
+		\textbf{Maximum mean discrepancy} (MMD) between different batches.
+		Lower MMD indicates that the distributions of the different batches are more similar.
 		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
-	In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
+		In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
+		See also Table \ref{tab:mmd}.
+	}
 	\label{fig:mmd}
 \end{figure}
+\begin{table}
+	\centering
+	\input{figures/final/mmd.tex}
+	\caption[Maximum mean discrepancy (MMD)]{
+		\pdfbookmarkraised{1}{Table \ref{tab:mmd}: Maximum mean discrepancy (MMD)}{tab:mmd}
+		\textbf{Maximum mean discrepancy} (MMD) comparing the distributions of the batches to each other after a given adjustment.
+		Lower MMD indicates that the distributions of the different batches are more similar.
+		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
+		See also \figurename{} \ref{fig:mmd}.
+	}
+	\label{tab:mmd}
+\end{table}
 
 We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable by any machine learning classifier.
-Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see Figure \ref{fig:batch}).
+Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see \figurename{} \ref{fig:batch}).
 With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:batch}}{batch}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:batch}: Batch classification accuracy}{fig:batch}
 	\centering
@@
 	\caption[Batch classification accuracy]{
-		\pdfbookmark[1]{Table \ref{tab:batch}}{batch}
+		\pdfbookmarkraised{1}{Table \ref{tab:batch}: Batch classification accuracy}{tab:batch}
 		\textbf{Batch classification accuracy} for several datasets and adjusters.
 		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
-		See also Figure \ref{fig:batch}.
+		See also \figurename{} \ref{fig:batch}.
 	}
@@
 
-With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see Figure \ref{fig:true_class} and Table \ref{tab:true_class}).
+With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
 For the Bladderbatch dataset, true class accuracy is much lower after adjusting with any algorithm, indicating that cancer status and batch may not be independent.
@@
 With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
-A look at the MNIST digits before and after adjustment (see Figure \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
+A look at the MNIST digits before and after adjustment (see \figurename{} \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
 With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs.
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:true_class}}{true_class}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:true_class}: True class classification accuracy}{fig:true_class}
 	\centering
@@
 	\caption[True class classification accuracy]{
-		\pdfbookmark[1]{Table \ref{tab:true_class}}{true_class}
+		\pdfbookmarkraised{1}{Table \ref{tab:true_class}: True class classification accuracy}{tab:true_class}
 		\textbf{True class classification accuracy} for several datasets and adjusters.
 		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
-		See also Figure \ref{fig:true_class}.
+		See also \figurename{} \ref{fig:true_class}.
 	}
@@
 \begin{figure}
-	\pdfbookmark[1]{Figure \ref{fig:mnist}}{asdf}
+	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mnist}: MNIST handwritten digits}{fig:mnist}
 	\centering
@@
 \begin{enumerate*}[(a)]
-	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in Figure \ref{fig:mnist}).
+	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in \figurename{} \ref{fig:mnist}).
 	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.

### 2019-05-30 thesis.tex
Add acknowledgments

@@
 	\begin{center}
-		ACKNOWLEDGMENTS\\
-		TODO: Fill this out.
+		ACKNOWLEDGMENTS
 	\end{center}
+	\paragraph{}
+	Thanks to Dr. Stephen Piccolo, whose mentorship over the past four years has shaped my education much more than the classes I took.
+	Thanks to Dr. Mark Clement and Dr. David Wingate for participating in my graduate committee and providing helpful feedback throughout my degree.
+	Thanks to my office- and lab-mates, who let me bounce ideas off of them.
+	Thanks to the BYU Department of Biology for funding my education, and specifically to Dr.
+Byron Adams for making the graduate experience more enjoyable, and to Gentri Glaittli for helping me to navigate the university requirements.
+	Thanks to my grandmother, Jennette Hawkes, whose financial help came at a critical time in my education.
+	Thanks to my parents, Minnie and Bryan Dayton, for teaching me the value of an education, and to my sisters Merinda, Louisa, Eliza, and Anna for their support.
+	Thanks to Dr. Paul Horton, whose thorough peer review provided me with the idea for this project.
+	Thanks to the researchers who generated the data for this project and who made it available.
+
+	Most of all, thanks to my wife, Zanna, for her love and support over the course of my graduate education.
 \end{titlingpage}

### 2019-05-31 thesis.tex
Remove formatting specific to the thesis submission

--- 2019-05-29 thesis.tex
@@
-\documentclass[12pt]{article}
-\usepackage{titling}
-
-\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
-\author{Jonathan Bryan Dayton}
-% \author[1]{Jonathan B. Dayton}
-% \author[1]{Stephen R. Piccolo}
-% \affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
-\date{2019}
-
-% Times New Roman for the thesis submission
-% \usepackage{mathptmx}
+\documentclass[11pt]{article}
 \usepackage{graphicx}
@@
 \usepackage[square,numbers]{natbib}
-% \usepackage{natbib}
 \usepackage{floatrow}
 \floatsetup[table]{capposition=top}
-\usepackage[nomarkers,nolists,tabhead,fighead,tablesfirst]{endfloat}
-% \renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
-\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
-\usepackage{csvsimple}
+\usepackage[nomarkers,nolists]{endfloat}
+\usepackage[margin=1in,letterpaper]{geometry}
 \usepackage{multirow}
 \usepackage[table]{xcolor}
-\usepackage{changepage}
-\usepackage{tocloft}
-\usepackage[bookmarksnumbered,bookmarksopen=true,bookmarks=true,hidelinks,pdfpagelabels=true]{hyperref}
-\usepackage{bookmark}
-
-% Make all figures and tables hang towards the top
-\makeatletter
-  \providecommand*\setfloatlocations[2]{\@namedef{fps@#1}{#2}}
-\makeatother
-\setfloatlocations{figure}{htbp}
-\setfloatlocations{table}{htbp}
-
-% Prevent captions from setting stuff 2x and therefore making 2 pdfbookmarks
-\usepackage{caption}
-\captionsetup{singlelinecheck=off}
-
-% Some manual \pdfbookmarks are too low. This should raise them.
-\makeatletter
- \newcommand{\pdfbookmarkraised}[3]{\Hy@raisedlink{\pdfbookmark[#1]{#2}{#3}}}
-\makeatother
-
-% TODO: change `\nolinkurl`s to `\url`s
-% Add dots to sections in TOC
-\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}
-
-% stuff for restyling the table of contents
-\renewcommand\cftsecfont{\mdseries}
-\renewcommand\cftsecpagefont{\mdseries}
-\renewcommand{\contentsname}{\begin{center}\normalsize{\textnormal{TABLE OF CONTENTS}}\end{center}}
-\renewcommand{\listtablename}{\begin{center}\normalsize{\textnormal{LIST OF TABLES}}\end{center}}
-\renewcommand{\listfigurename}{\begin{center}\normalsize{\textnormal{LIST OF FIGURES}}\end{center}}
+\usepackage[hidelinks]{hyperref}
+\usepackage{lineno}
+\linenumbers
 
@@
 
-\pagenumbering{roman}
-\begin{titlingpage}
-	\setcounter{page}{1}
-	\addcontentsline{toc}{section}{Title Page}
-	\begin{adjustwidth}{+0.5in}{+0.5in}
-		\begin{center}
-			\doublespacing \thetitle\\ \singlespacing
-		\end{center}
-	\end{adjustwidth}
-	\centering{
-		\vfill
-		\theauthor\\
-		\vfill
-		A thesis submitted to the faculty of\\
-		Brigham Young University\\
-		in partial fulfillment of the requirements for the degree of\\
-		\vspace{1em}
-		Master of Science\\
-		\vfill
-		Stephen R. Piccolo, Chair\\
-		Mark Clement\\
-		David Wingate\\
-		\vfill
-		Department of Biology\\
-		\vspace{1em}
-		Brigham Young University
-		\vfill
-		Copyright \textcopyright \ \thedate \ \theauthor\\
-		\vspace{1em}
-		All Rights Reserved
-	\par}
+\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
+\author{Jonathan Bryan Dayton}
+\author[1]{Jonathan B. Dayton}
+\author[1]{Stephen R. Piccolo}
+\affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
 
-\end{titlingpage}
+\maketitle
 
-% Title / abstract page
-\begin{titlingpage}
-	\setcounter{page}{2}
-	\phantomsection
-	\addcontentsline{toc}{section}{Abstract}
-	\begin{center}
-		ABSTRACT\\
-		\vspace{1em}
-		\begin{adjustwidth}{+0.5in}{+0.5in}
-			\begin{center}
-				\thetitle\\
-			\end{center}
-		\end{adjustwidth}
-		\vspace{1em}
-		\theauthor\\
-		Department of Biology, BYU\\
-		Master of Science
-	\end{center}
-	% \begin{abstract}
-		\paragraph{} % adds indentation
+\begin{abstract}
 	Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
@@
 	We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
-		Our software is publicly available at \nolinkurl{https://github.com/jdayton3/Confounded}.
-	% \end{abstract}
-	\vfill
-	\noindent
-	Keywords:\@ batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder % the \@ removes the double "\nonfrenchspacing" spaces after the colon, which is required for thesis submission
-\end{titlingpage}
+	Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
+\end{abstract}
 
 \doublespacing
-
-\begin{titlingpage}
-	\setcounter{page}{3}
-	\phantomsection
-	\addcontentsline{toc}{section}{Acknowledgments}
-	\begin{center}
-		ACKNOWLEDGMENTS\\
-		TODO: Fill this out.
-	\end{center}
-\end{titlingpage}
-
-\setcounter{page}{4}
-\phantomsection
-\addcontentsline{toc}{section}{Table of Contents}
-\tableofcontents
-\newpage
-\phantomsection
-\addcontentsline{toc}{section}{List of \tablename{s}}
-{% Change table refs "1", "2" -> "Table 1", "Table 2" in LOT
-\let\oldnumberline\numberline%
-\renewcommand{\numberline}{\tablename~\oldnumberline}%
-\listoftables%
-}
-\newpage
-\phantomsection
-\addcontentsline{toc}{section}{List of \figurename{s}}
-{% See comment on list of tables
-\let\oldnumberline\numberline%
-\renewcommand{\numberline}{\figurename~\oldnumberline}%
-\listoffigures%
-}
-\newpage
-
-% make bookmarks for figure and tables sections
-\renewcommand{\figuresection}{\pdfbookmarkraised{0}{\figurename{s}}{figure}\figurename{s}}
-\renewcommand{\tablesection}{\pdfbookmarkraised{0}{\tablename{s}}{table}\tablename{s}}
-
-\pagenumbering{arabic}
 \section{Background} \label{sec:background}
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:workflow}: Batch adjustment justification and steps}{fig:workflow}
 	\centering
-	% TODO I changed the width so "Figures" would be on the same page as the 1st fig per Gentri's request. Change this back eventually
-	\includegraphics[width=0.85\columnwidth]{figures/final/adjuster_workflow.pdf}
+	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
 	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
@@
 
-All our code has been made publicly available at \nolinkurl{https://github.com/jdayton3/Confounded}, and all our data are available as described below.
+All our code has been made publicly available at \url{https://github.com/jdayton3/Confounded}, and all our data are available as described below.
 
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:network}: Network architecture of Confounded}{fig:network}
 	\centering
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:training_loss}: Autoencoder and discriminator loss}{fig:training_loss}
 	\centering
@@
 	\caption[Dataset information]{
-		\pdfbookmarkraised{1}{Table \ref{tab:datasets}: Dataset information}{tab:datasets}
 		\textbf{Dataset information} for each dataset used.
@@
 
-% \paragraph{Synthetic Batch Effects} % TODO: uncomment this after submitting thesis
+\paragraph{Synthetic Batch Effects}
 
@@
 The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this dataset from \nolinkurl{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+We accessed a version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
 It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
@@
 However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
-Here, we used the same version of the dataset that we tidied in this previous study (available at \nolinkurl{https://osf.io/7xjdn/}).
+Here, we used the same version of the dataset that we tidied in this previous study (available at \url{https://osf.io/7xjdn/}).
 This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
@@
 There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
-% TODO: add like 8 more methods
 Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:pca}: Principal components analysis (PCA)}{fig:pca}
 	\centering
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:tsne}: T-distributed Stochastic Neighbor Embedding (t-SNE)}{fig:tsne}
 	\centering
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mse}: Mean squared error (MSE)}{fig:mse}
 	\centering
@@
 	\caption[Mean squared error (MSE)]{
-		\pdfbookmarkraised{1}{Table \ref{tab:mse}: Mean squared error (MSE)}{tab:mse}
 		\textbf{Mean squared error (MSE)} of the unadjusted input data compared to the data output by the given adjusters.
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mmd}: Maximum mean discrepancy (MMD)}{fig:mmd}
 	\centering
@@
 	\caption[Maximum mean discrepancy (MMD)]{
-		\pdfbookmarkraised{1}{Table \ref{tab:mmd}: Maximum mean discrepancy (MMD)}{tab:mmd}
 		\textbf{Maximum mean discrepancy} (MMD) comparing the distributions of the batches to each other after a given adjustment.
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:batch}: Batch classification accuracy}{fig:batch}
 	\centering
@@
 	\caption[Batch classification accuracy]{
-		\pdfbookmarkraised{1}{Table \ref{tab:batch}: Batch classification accuracy}{tab:batch}
 		\textbf{Batch classification accuracy} for several datasets and adjusters.
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:true_class}: True class classification accuracy}{fig:true_class}
 	\centering
@@
 	\caption[True class classification accuracy]{
-		\pdfbookmarkraised{1}{Table \ref{tab:true_class}: True class classification accuracy}{tab:true_class}
 		\textbf{True class classification accuracy} for several datasets and adjusters.
@@
 \begin{figure}
-	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mnist}: MNIST handwritten digits}{fig:mnist}
 	\centering

### 2019-06-14 supplement.tex
Add supplementary material

@@
-We also tried replacing the ReLU activations with SELUs \cite{klambauer_self-normalizing_2017} but saw little if any improvement and therefore returned to using ReLUs.
+\documentclass{article}
+\usepackage{graphicx}
+\usepackage{setspace}
+% \usepackage{authblk}
+\usepackage{url}
+\usepackage[inline,shortlabels]{enumitem}
+\usepackage[square,numbers]{natbib}
+\usepackage{floatrow}
+\floatsetup[table]{capposition=top}
+\usepackage[nomarkers,nolists]{endfloat}
+\usepackage[margin=1in,letterpaper]{geometry}
+\usepackage{multirow}
+\usepackage[table]{xcolor}
+\usepackage[hidelinks]{hyperref}
+\usepackage{lineno}
+\linenumbers
+
+\bibliographystyle{unsrtnat}
+
+\begin{document}
+
+\title{Supplementary Material}
+\date{\vspace{-5ex}}
+
+\maketitle
+
+\doublespacing
+
+\section{Optimizing default parameters}
+
+We tried a number of parameter optimizations to set reasonable defaults.
+We list them below and further discuss the bolded items.
+
+\paragraph{Things that helped}
+
+\begin{enumerate}
+	\item \textbf{Adjust autoencoder depth}
+	\item \textbf{Adjust discriminator depth}
+	\item \textbf{Scale to [0, 1] with sigmoid}
+	\item Make code layer more narrow
+	\item Optimizing cross-entropy instead of MSE - trains faster
+	\item Adjusting the lambda value - somewhat dataset specific
+	\item Variational instead of normal autoencoder - loss is less erratic
+	\item Dropout in the discriminator - less overfitting
+\end{enumerate}
+
+\paragraph{Things that didn't help}
+
+\begin{enumerate}
+	\item ReLU vs SELU \cite{klambauer_self-normalizing_2017} activation - no difference
+	\item Recurrent discriminator connections - no difference
+	\item Recurrent autoencoder connections (like UNet) - no difference
+	\item Training autoencoder, discriminator, then dual optimizers sequentially - did worse
+	\item Stopping early - did worse
+\end{enumerate}
+
+\paragraph{Adjust autoencoder depth}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/ae_layers.png}
+	\caption{\textbf{Loss vs. Number of Autoencoder Layers}
+	}
+	\label{fig:ae}
+\end{figure}
+
+In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
+We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
+We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
+
+\paragraph{Adjust discriminator depth}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/disc_layers.png}
+	\caption{\textbf{Loss vs. Number of Discriminator Layers}
+	}
+	\label{fig:disc}
+\end{figure}
+
+We noticed that the discriminator was overfitting on the data.
+Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
+
+\paragraph{Scale to [0, 1] with sigmoid}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/scaling.png}
+	\caption{\textbf{Loss vs. Scaling Method}
+	}
+	\label{fig:scaling}
+\end{figure}
+
+\paragraph{Other ideas}
 
@@
 
-Autoencoder
-
-In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
-We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
-We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
-
-Discriminator
-
-We noticed that the discriminator was overfitting on the data.
-Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
-
-Datasets
+\paragraph{Datasets}
 
@@
 
-\subsubsection{Data Format}
+\section{Data Format}
 
@@
 
-GSE37199
+\paragraph{GSE37199}
 
@@
 
-Other methods
+\section{Testing different-sized datasets}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/tcga_mse.pdf}
+	\caption{\textbf{TCGA MSE}
+	}
+	\label{fig:mse}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/tcga_mmd.pdf}
+	\caption{\textbf{TCGA MMD}
+	}
+	\label{fig:mmd}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/tcga_true_class_accuracy.pdf}
+	\caption{\textbf{TCGA True Class Accuracy}
+	}
+	\label{fig:true_class}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/tcga_batch_accuracy.pdf}
+	\caption{\textbf{TCGA Batch Accuracy}
+	}
+	\label{fig:batch}
+\end{figure}
+
+\section{Other methods}
 
@@
 With SVA, the documentation showed how to identify surrogate variables and use them in modeling, but we were unable to find examples of removing the effects that those surrogate variables represent.
-We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code (see \ref{section:mmd} for more information).
+We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code.
 
-MSE
+\section{Metrics used}
+
+\paragraph{MSE}
 
@@
 
-MMD
+\paragraph{MMD}
 
@@
 % TODO: make this formula ^ apply to more than 3 batches
+
+\paragraph{PCA \& t-SNE}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/mnist_pca.pdf}
+	\caption{\textbf{MNIST PCA}
+	}
+	\label{fig:pca}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/supplement/mnist_tsne.pdf}
+	\caption{\textbf{MNIST t-SNE}
+	}
+	\label{fig:tsne}
+\end{figure}
+
+\bibliography{references}
+
+\end{document}

### 2019-06-14 supplement.tex
Add information about dropout and ae depth

@@
 
-We tried a number of parameter optimizations to set reasonable defaults.
-We list them below and further discuss the bolded items.
+We tried a number of parameter optimizations so we could set reasonable defaults for Confounded.
+As expected, some parameter adjustments helped, and some did not.
+We list several of our adjustments below and further discuss the bolded items.
 
@@
 	\item \textbf{Scale to [0, 1] with sigmoid}
+	\item \textbf{Dropout in the discriminator} (less overfitting) % Cite dropout paper
 	\item Make code layer more narrow
-	\item Optimizing cross-entropy instead of MSE - trains faster
-	\item Adjusting the lambda value - somewhat dataset specific
-	\item Variational instead of normal autoencoder - loss is less erratic
-	\item Dropout in the discriminator - less overfitting
+	\item Optimizing cross-entropy instead of MSE (training happened faster) % Cite xentropy paper
+	\item Adjusting the lambda value (somewhat dataset specific)
+	\item Variational instead of normal autoencoder \cite{kingma_auto-encoding_2013} (loss became less erratic)
 \end{enumerate}
@@
 \begin{enumerate}
-	\item ReLU vs SELU \cite{klambauer_self-normalizing_2017} activation - no difference
-	\item Recurrent discriminator connections - no difference
-	\item Recurrent autoencoder connections (like UNet) - no difference
-	\item Training autoencoder, discriminator, then dual optimizers sequentially - did worse
-	\item Stopping early - did worse
+	\item ReLU vs SELU \cite{klambauer_self-normalizing_2017} activation (no difference)
+	\item Recurrent discriminator connections (no difference) % Cite paper about recurrent connections
+	\item Recurrent autoencoder connections, similar to the U-Net architecture \cite{ronneberger_u-net_2015} (no difference)
+	\item Training autoencoder, discriminator, then dual optimizers sequentially (did worse)
+	\item Stopping early (did worse)
 \end{enumerate}
@@
 \paragraph{Adjust autoencoder depth}
+
+A neural network with no hidden layers (i.e. a layer of perceptrons) can only model linear functions;
+it must have one or more hidden layers in order to model nonlinear functions.
+However, in some cases, neural networks work better with large amounts of hidden layers, and in other cases, the opposite is true. % Citation needed
+The number of hidden layers is somewhat problem-specific.
+We varied the number of hidden layers in the autoencoder to figure out how many layers are sufficient, and in our testing, after adding a single hidden layer in both the encoder and the decoder, additional hidden layers had no noticeable effect (see \figurename{} \ref{fig:ae}).
+Therefore, Confounded's default autoencoder depth is 1.
 
@@
 \end{figure}
-
-In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
-We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
-We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
 
@@
 \end{figure}
+
+\paragraph{Dropout in the discriminator}
+
+When training initially with a plain autoencoder and discriminator (meaning subnetworks with only weights, biases, and activation functions), the network trained very poorly.
+For example, the outputs for the MNIST datasets looked only like random noise rather than like handwritten digits.
+We hypothesized that this may be because the discriminator was overfitting;
+rather than learn what the batch effect looked like, it would memorize every sample in each batch, and then when it saw output from the autoencoder that looked like one of the memorized outputs, it would be able to predict the batch perfectly, regardless of whether the batch effect remained in the data.
+In order to prevent this overfitting, we added dropout to all the layers with a probability of 50\%.
+Once we added dropout, the network began to output images that looked like the original MNIST digits.
+Our interpretation of this result is that dropout was able to successfully discourage the discriminator from overfitting.
 

### 2019-06-15 supplement.tex
Add captions to supplementary figures

@@
 	\includegraphics[width=\columnwidth]{figures/supplement/ae_layers.png}
-	\caption{\textbf{Loss vs. Number of Autoencoder Layers}
+	\caption{\textbf{Loss vs. Number of Autoencoder Layers}.
+	After one hidden layer in both the encoder and the decoder, adding additional hidden layers to the autoencoder didn't have a large effect on how well Confounded performed.
 	}
@@
 
+It is crucial that a batch adjuster be able to identify and model batch effects if it is to remove them.
+This is particularly the case for Confounded, which tries to identify and remove even complex nonlinear batch and confounding effects.
+The discriminator must be adequately powerful in order for the adversarial system to remove these effects.
+A deeper discriminator, or one with more hidden layers, could potentially be more powerful in identifying these complex nonlinear relationships, but adding hidden layers to a network comes at the costs of longer training times and potentially less effective training.
+% Find a source for this.
+Confounded's discriminator must then have a ``just right'' number of hidden layers to balance these benefits and costs.
+Over testing the otherwise same architecture with different-depth discriminators, we found the general trend that as the number of hidden discriminator layers increases, the overall dual loss decreases (see \figurename{} \ref{fig:disc}).
+There seemed to be a trough to this effect around 10 hidden layers, so we left 10 as the default.
+
 \begin{figure}
@@
 	\includegraphics[width=\columnwidth]{figures/supplement/disc_layers.png}
-	\caption{\textbf{Loss vs. Number of Discriminator Layers}
+	\caption{\textbf{Loss vs. Number of Discriminator Layers}.
+	As the number of hidden layers increases in the discriminator, the dual loss generally decreases.
+	This pattern happens naturally because the discriminator learns more slowly with more layers, but there seems to be a possible trough at 10 hidden layers.
 	}
@@
 
-We noticed that the discriminator was overfitting on the data.
-Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
+\paragraph{Scale to [0, 1] with sigmoid}
 
-\paragraph{Scale to [0, 1] with sigmoid}
+Neural networks typically perform better if their inputs are in the same range. % Find a source.
+However, gene expression data often have highly variable column ranges. % Get the min and max for the TCGA dataset
+Prior to passing data through Confounded, scaled each column to the same range, [0, 1].
+Initially, we did this linearly by subtracting the minimum from each column and then dividing by the column range.
+We then tried scaling the data using the sigmoid function, which scales values to [0, 1] but squishes very low or very high values more than the average values.
+In both cases, we expand the data to the original range after adjustment by applying the inverse of the scaling function.
+The sigmoid function appeared to give slightly better results (see \figurename{} \ref{fig:scaling}), so sigmoid scaling is Confounded's default.
 
@@
 	\caption{\textbf{Loss vs. Scaling Method}
+	Confounded performed slightly better wen scaling the input data to [0, 1] with the sigmoid function rather than a linear scaling function.
 	}
@@
 
-\paragraph{Other ideas}
+\paragraph{Other unimplemented ideas}
 
@@
 
+The GSE37199 dataset has two levels of batches (``centre'' and ``plate'').
+We identified this as an item of potential interest: can Confounded adjust for multiple batch levels of effects simultaneously?
+However, we left this question for future research and chose instead to focus on the ``plate'' effect for this study.
+
 \paragraph{Datasets}
@@
 Working with each new dataset enlightened us of a new edge case that hadn't yet been addressed in our code.
-By applying to our methods to each of the following datasets, our code became much more robust than if we had only developed and tested on one or two toy datasets.
+By applying to our methods to each of the four described datasets, our code became much more robust than if we had only developed and tested on one or two datasets.
 
@@
 In a future iteration, it would be more robust to allow the user to specify continuous columns that should not be interpreted as gene expression data (e.g.
-ages in years with decimal precision representing fractions of years).
+ages in years with decimal precision representing fractions of years), or to pass expression values in as a different file or dataframe.
 However for the datasets we tested, this didn't prove to be a problem.
 
-\paragraph{GSE37199}
+\section{Testing different-sized datasets}
 
-Additionally, this dataset has two levels of batches (``centre'' and ``plate'').
-We identified this as an item of potential interest: can Confounded adjust for multiple batch effects simultaneously?
-However, we left this question for future research and chose instead to focus on the ``plate'' effect for the time being.
+One big question that remained after testing our four datasets was why Confounded works better on some datasets and worse on others.
+Our main hypothesis was that Confounded, like most tools that use deep learning, would inherently work better on larger datasets than on smaller ones.
+We made two random subsets (sizes X and Y) of the TCGA dataset to assess how well Confounded performs relative to a scaling adjuster and to ComBat. % Put the actual sizes in here
+Our results show that with the TCGA dataset, Confounded continues to outperform ComBat with the medium and smaller subsets (see \figurename{s} \ref{fig:mse}, \ref{fig:mmd}, \ref{fig:true_class}, and \ref{fig:batch}).
+This may indicate that Confounded works better on more complex confounding effects, or perhaps that it works better with a smaller columnset (our TCGA dataset was restricted to 325 columns).
+Future testing may be required to identify general cases when Confounded will outperform classic batch adjusters such as ComBat, but we anticipate that Confounded may be better particularly in removing confounding effects, whereas classic methods may be better on typical batch effects.
 
-\section{Testing different-sized datasets}
+\newcommand{\disclaimer}{Note: the scaling function produced errors when run on the small TCGA dataset and was therefore not used in this chart.}
 
@@
 	\includegraphics[width=\columnwidth]{figures/supplement/tcga_mse.pdf}
-	\caption{\textbf{TCGA MSE}
+	\caption{\textbf{TCGA MSE}.
+	Confounded consistently outperforms other adjusters with different sizes of the TCGA dataset in terms of mean squared error, meaning that Confounded is less disruptive to the original data.
+	\disclaimer{}
 	}
@@
 	\caption{\textbf{TCGA MMD}
+	Confounded outperforms other adjusters on the different sizes of the TCGA dataset in terms of max mean discrepancy, meaning that Confounded better aligns the expression value distributions between different cancer types.
+	\disclaimer{}
 	}
@@
 	\includegraphics[width=\columnwidth]{figures/supplement/tcga_true_class_accuracy.pdf}
-	\caption{\textbf{TCGA True Class Accuracy}
+	\caption{\textbf{TCGA True Class Accuracy}.
+	Confounded typically has a lower true class accuracy than the other adjusters for the TCGA dataset.
+	However, this is likely because the true class (TP53 mutation status) is confounded with the batch (CancerType) that we've tried to remove.
+	\disclaimer{}
 	}
@@
 	\includegraphics[width=\columnwidth]{figures/supplement/tcga_batch_accuracy.pdf}
+	Confounded consistently removes more of the batch effects than the other adjusters as measured by Naive Bayes and Random Forests classification accuracy.
 	\caption{\textbf{TCGA Batch Accuracy}
+	\disclaimer{}
 	}
@@
 
-We wanted to also compare with SVA \cite{leek_capturing_2007} and the Shaham lab's work \cite{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods to work.
+We wanted to also compare with SVA \cite{leek_capturing_2007} and the work of \citet{shaham_removal_2017} \citep{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods running.
 With SVA, the documentation showed how to identify surrogate variables and use them in modeling, but we were unable to find examples of removing the effects that those surrogate variables represent.
-We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code.
+We were unable to get \citet{shaham_removal_2017}'s methods running on our datasets, and we believe that this is due to some underlying assumptions that make their code more applicable to larger datasets.
 
@@
 
-We wanted to compare our network against this network, so we calculated MMD for each adjuster.
-Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).
-
-following formula:
+We calculated MMD using the following formula:
 
@@
 Where $k(x, y)$ is the Gaussian kernel between $x$ and $y$ as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel}, and where $x$ represents the data from one batch and $y$ represents the data from another batch.
-In cases where there were more than two batches (e.g. $x$, $y$, and $z$), we averaged the pairwise MMD values to calculate an overall MMD:
+In cases where there were more than two batches (e.g. $x$, $y$, and $z$), we averaged the pairwise MMD values to calculate an overall MMD.
 
-\begin{equation}
-	\label{mmd-mean}
-	MMD_{total}(x, y, z) = \frac{1}{3}[MMD(x, y) + MMD(x, z) + MMD(y, z)]
-\end{equation}
-% TODO: make this formula ^ apply to more than 3 batches
+\paragraph{MNIST PCA \& t-SNE}
 
-\paragraph{PCA \& t-SNE}
+See \figurename{s} \ref{fig:pca} and \ref{fig:tsne}.
 
@@
 	\caption{\textbf{MNIST PCA}
+	All adjusters seem to remove the strong linear batch effect in the MNIST dataset.
 	}
@@
 	\caption{\textbf{MNIST t-SNE}
+	Both ComBat and Confounded seem to remove batch effect from the MNIST dataset as shown by a t-SNE plot.
 	}

### 2019-07-03 thesis.tex
Revised Introduction section

@@
 	\begin{enumerate*}[(a)]
-		\item When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
+		\item When measurements are collected from a sample ($X$), systemic effects ($H$) may affect the measurements.
 		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
@@
 
-Several methods exist for removing batch effects from gene expression datasets.
+Several methods exist for characterizing and removing batch effects from gene-expression data.
 Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
-ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
-Since both of these methods use linear methods to model confounding effects, they are not designed to account for nonlinear effects, such as the cascading changes in gene pathways in response to environmental stressors (e.g.
-gene X transcripts increase and gene Y transcripts decrease in response to rising temperature, and since both are now past a certain threshold, gene Z transcripts also increase, and so on).
-As machine learning becomes more common in biological research, these nonlinear confounding effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
-For example, advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
-These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
-Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
-However, several limitations complicate their usability on many real-world datasets:
-they require that the input data only contains two batches, that the batches are sufficiently large (we received errors when testing with a subset of MNIST where n=100), and that the batches are perfectly balanced.
-These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
-Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.
+ComBat uses an empirical Bayes approach to estimate batch effect parameters and then uses linear regression to model and remove the effects.
+SVA uses singular value decomposition to model batch effects, which can then be accounted for in statistical analyses.
+Both of these methods use linear methods to model confounding effects and thus may not account for nonlinear effects, such as multi-gene, cascading interactions within signaling pathways.
+As machine learning becomes more common in biological research, these nonlinear confounding effects become more troublesome because many machine-learning algorithms can successfully account for complex interactions among variables.
+%For example, advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
+%These networks have proven effective in removing irrelevant, domain-specific signal in credit-rating, online-review, and image-recognition tasks \citep{louizos_variational_2015}.
 
-To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
-Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
-This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
-Research has shown that neural networks are effective in working with gene expression data; for example, neural networks have been applied
-to detect cancer and identify critical cancer genes \citep{danaee_deep_2016},
-to infer gene-expression values from just the values of a few ``landmark genes'' \citep{chen_gene_2016},
-to extract biologically relevant latent spaces in RNA-Seq data \citep{way_extracting_2017},
-to reduce the dimensions of single-cell RNA-Seq data \citep{lin_using_2017},
-to identify drug-repurposing targets using transcriptomic data \citep{aliper_deep_2016},
-and to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
-Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output to be as similar to the input as possible \citep{hinton_reducing_2006}.
-Neural networks have historically decreased in effectiveness when working with data from multiple research domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on dataset-specific confounding effects (e.g.
-which researcher collected the data) instead of learning based on practically meaningful causal effects (e.g.
-which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
-Recently, researchers have experimented with discouraging neural networks from learning based on domain-specific information by giving them two objective functions:
-1. to learn as much as possible about the input data and
-2. to forget any patterns that help distinguish between domains \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
-\citet{louizos_variational_2015} used this type of dual objective function with a variational autoencoder and successfully removed domain-based variability in credit score, financial savings, and hospital admittance datasets.
+Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
+Neural networks have been applied broadly to gene-expression data.
+For example, they have been used to detect cancer and identify critical cancer genes \citep{danaee_deep_2016}, to impute gene-expression levels from the values of a few ``landmark genes'' \citep{chen_gene_2016}, to extract biologically relevant latent spaces in RNA-Seq data \citep{way_extracting_2017}, to reduce the dimensions of single-cell RNA-Seq data \citep{lin_using_2017}, to identify drug-repurposing targets \citep{aliper_deep_2016}, and to generate synthetic biomedical data as a way to preserve patient privacy \citep{beaulieu-jones_privacy-preserving_2017}.
 
-In this study, we present Confounded, an adversarial autoencoder that identifies and removes confounding effects.
-We test the hypothesis that using an adversarial neural network can correct for confounding effects more completely than previous tools do.
-We also explore the extent to which confounders still remain in different datasets after adjustment with various algorithms, and we present a framework to assess the extent to which confounding effects remain after adjustment using various classification algorithms.
+Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
+However, these methods are designed for a limited range of scenarios:
+1) when the input data contain only two batches, 2) the batches are sufficiently large (> 100 samples in our testing), and 3) the batches are balanced.[@For #2, does this mean the total number of samples?
+Or the number of samples per batch?
+Please clarify the wording.@] These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
+Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
+To address these limitations, we implemented a deep neural network that [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.@] Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.
+
+Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here.@].
+This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.@] Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
+However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.@] Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
+Louizos, et al.
+have used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.
+
+In this study, we present \textit{Confounded}, an adversarial autoencoder that identifies and removes confounding effects from gene-expression data.
+We test the hypothesis that using an adversarial neural network can correct for confounding effects more effectively than alternative tools.
+We present a classification-based framework to assess the extent to which confounding effects remain after adjustment, and we apply these techniques to four datasets.
 

### 2019-07-04 thesis.tex
Added newline characters
Revised methods section

@@
 However, these methods are designed for a limited range of scenarios:
-1) when the input data contain only two batches, 2) the batches are sufficiently large (> 100 samples in our testing), and 3) the batches are balanced.[@For #2, does this mean the total number of samples?
-Or the number of samples per batch?
-Please clarify the wording.@] These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
+1) when the input data contain only two batches, 
+2) the batches are sufficiently large (> 100 samples in our testing), and 
+3) the batches are balanced.[@For #2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording.@] 
+These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
 Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
-To address these limitations, we implemented a deep neural network that [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.@] Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.
+To address these limitations, we implemented a deep neural network that [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.@]
+Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.
 
 Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here.@].
-This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.@] Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
-However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.@] Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
+This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.@]
+Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}. 
+However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.@]
+Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
 Louizos, et al.
@@
 
-All our code has been made publicly available at \url{https://github.com/jdayton3/Confounded}, and all our data are available as described below.
+An open-source implementation of Confounded is publicly available at \url{https://github.com/jdayton3/Confounded}.
+Materials for generating this manuscript, including scripts and data, are available at \url{https://github.com/jdayton3/ms-thesis/tree/bioinformatics} and \url{https://osf.io/b76ch/}, respectively.
 
-\subsection{Network Structure}
+\subsection{Neural network structure}
 
-We used an adversarial autoencoder network to model and remove the confounding effects.
+We used an adversarial autoencoder network to model and remove confounding effects.
 We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining confounding effects in the autoencoder's output.
-By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data with confounding effects minimized.
+By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learns over the course of training to output the expression data with confounding effects minimized.
 We implemented the neural network in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
-All layers in the network were fully connected and all activation functions were Rectified Linear Units (ReLUs) \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.
+All layers in the network are fully connected, and all activation functions are Rectified Linear Units (ReLUs) \cite{agarap_deep_2018}, except the final layers in the autoencoder and the discriminator, which use the sigmoid function.
 
@@
 
-We implemented the variational autoencoder \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
+We implemented the variational autoencoder architecture \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
 This network has 2 hidden encoding layers and 2 decoding layers, each of size 500.
 The code size is 20.
-Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015}.
-It is trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).
+Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015} and trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).
 
@@
 
-We trained the discriminator to determine the original batch of the autoencoder's output.
-The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches.
-In order to combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
-These additions seemed to reduce overfitting in the discriminator.
+We trained the discriminator to predict the original batch of the autoencoder's output.
+The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches[@Please provide a little more detail about how the output layer size is calculated@].
+To combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
+These additions seem to reduce overfitting in the discriminator.
 
@@
 
-The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing confounding effects.
+The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency to more faithfully replicate the input or to more completely remove confounding effects.
 A higher $\lambda$ value indicates that the network should remove confounding effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the input data.
@@
 When optimizing $L_D$ (i.e. training the discriminator), we froze the autoencoder's weights, and vice versa.
-We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ CPU and 16 GB of RAM.
-For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.
+We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ central processing unit and 16 GB of random access memory.
+For each dataset, training took roughly 30 minutes to complete, including time to load the input into memory and to save the output to disk.
 
@@
 	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
-	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded on the MNIST dataset.
+	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded on the MNIST dataset (which we used to validate the network).
 	Over the course of training, the autoencoder more faithfully replicates the input data.
-	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
+	The autoencoder also seems to introduce noise (see the red dashed line near iteration 3100) in response to the discriminator's slight improvements.}
 	\label{fig:training_loss}
@@
 
+\subsection{Comparison to other methods}
+
+We compared Confounded against two other methods that can be used for batch correction. The first was a simple scaling adjuster,
+which we implemented in the R programming language (version 3.6.0) \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
+It adjusts the data by linearly expanding or contracting the values for each variable so that all variables have the same range across the batches.
+We also compared against ComBat \citep{johnson_adjusting_2007} using an implementation from the sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance.
+
+We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
+A number of other methods for batch adjustment are available.
+Some of these use deep neural networks \citep{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}, while others do not\citep{espin-perez_comparison_2018}.
+Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.[@This won't be convincing enough to the reviewers.
+Be more explicit about why we couldn't use these other methods.@]
+
 \subsection{Datasets}
 
-In order to test both theoretical and practical differences between Confounded and previous methods, we compared them for a variety of datasets of varying sizes and type of data measured.
+To test both theoretical and practical differences between Confounded and the alternative methods, we compared them for datasets that varied in size and type of data measured.
 
@@
 	\centering
+	\caption[Dataset information]{
+		\textbf{Information} about each dataset used to compare algorithm performance.
+	}
 	\input{figures/final/datasets.tex}
-	\caption[Dataset information]{
-		\textbf{Dataset information} for each dataset used.
-	}
 	\label{tab:datasets}
@@
 
-The MNIST digits dataset \cite{lecun_mnist_nodate} is a database of images of handwritten digits that are size-normalized and centered.
+The MNIST digits dataset \cite{lecun_mnist_nodate} contains images of handwritten digits that are size-normalized and centered.
 It contains 60,000 training images and 10,000 test images.
-We used MNIST so we could visually assess how well the true signal (in this case, the shape and digit of each handwritten digit image) was preserved after batch adjustment.
-In order to use this dataset, we flattened each 28 by 28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit information.
-We limited our dataset to only the 10,000 test images.
-Although convolutional layers are typically used when working with image data, we only used fully connected layers even for this image dataset.
-In this way, we show that the autoencoder is still able to find and represent spatial relationships without explicitly defining spatial relationships in the model while testing the same network we use on expression data, where no spatial relationships are inherent.
+We used MNIST so we could visually assess how well the batch-correction methods preserve true signal (in this case, the shape and digit of each handwritten digit image) after simulated batch adjustment.
+We flattened each 28-by-28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit (class) information.
+We limited our analysis to the 10,000 test images.
+Although convolutional layers are typically used when working with image data, we used fully connected layers to maintain consistency with what would be used on gene-expression data.
+However, the autoencoder was still able to represent spatial relationships without explicitly defining them in the model.
 
-\paragraph{Synthetic Batch Effects}
+\paragraph{Synthetic batch effects}
 
-Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear confonding effects.
-To do so, we wrote a Python script to take the MNIST data in, apply a nonlinear effect, and output the adjusted data.
+Because there is no batch information in the MNIST digits dataset, we simulated nonlinear confonding effects.
 We applied nonlinear effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
-We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
-We applied the same random vectors to each image in a batch.
-Finally, we added random noise to each image in order to prevent images in a batch from being overly similar to each other.
+We split the images into two balanced batches (5,000 images for each batch) and included the same number of each digit in each batch.
+We applied the same random vectors to each image in a given batch.
+Finally, we added random noise to each image to prevent images in a batch from being overly similar to each other.
 
-\subsubsection{Bladderbatch}
+\subsubsection{Bladderbatch data}
 
-The bladderbatch dataset is a microarray transcriptomic expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
-It has been made available as an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package \cite{leek_sva_2017} to illustrate how to batch-adjust using ComBat.
+The bladderbatch dataset is a microarray gene-expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
+It is available as an R package \cite{leek_bladderbatch_2017}.
 It contains expression values for 57 tissue samples with and without bladder cancer across 5 unbalanced batches.
-The dataset has a cancer status (cancerous vs. normal tissue) column, which we used for ``true class'' classification, and a batch column.
-Because bladderbatch is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.
+The dataset has a cancer status (cancerous vs. normal tissue) variable, which we used as the ``true class'' in our validations; it also has a batch variable.
+Because bladderbatch is small relative to datasets typically used for deep learning, it helped us to evaluate whether our network overfits on gene-expression datasets of the size often used in applied research.
 
@@
 
-The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
-We accessed a version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
+The GSE37199 dataset was profiled using Affymetrix microarrays from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
+We used a curated version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T}\cite{golightly_curated_2018}.
 It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
-We used this cancer status variable as the ``true class'' for classification.
-It has two types of batch variables: ``plate'' and ``centre.''
-We adjusted against the ``plate'' variable because it was more balanced than ``centre'' (with counts of $\{43, 50\}$ compared to $\{27, 66\}$).
-The GSE37199 dataset represents a slightly larger dataset than bladderbatch, with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).
+We used these categories for the ``true class'' labels.
+This dataset has two types of batch variables---``plate'' and ``centre''---which indicate the microarray plate and research lab that processed the data, respectively.
+We adjusted against the ``plate'' variable because it was granular in nature (with counts of $\{43, 50\}$).
 
-\subsubsection{TCGA Pan-cancer Data}
+\subsubsection{TCGA pan-cancer data}
 
-The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
+The Cancer Genome Atlas (TCGA) produced RNA-Sequencing data for tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
 In a previous study \cite{dayton_classifying_2017-1}, we classified this dataset based on the presence or absence of mutations in several known cancer genes.
-We attempted to adjust for the confounding effect of cancer type using ComBat prior to classification.
+In that study, we observed that cancer type had a strong confounding effect on our analysis, in which sought to identify patterns spanning many cancer types. We adjusted for this effect using ComBat.
 However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
-Here, we used the same version of the dataset that we tidied in this previous study (available at \url{https://osf.io/7xjdn/}).
-This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
-We used this dataset as a way to test whether Confounded works on RNA-Seq data and to test whether we could remove confounding effects that ComBat cannot remove.
+Here, we used the same version of TCGA dataset (available at \url{https://osf.io/7xjdn/}),
+which has expression values for 9,365 samples across 25 cancer types.
 
-\subsection{Comparison to other methods}
-
-We compared our method to two other batch adjusters: a scale adjuster and ComBat \citep{johnson_adjusting_2007}.
-We implemented the scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
-It adjusts the data by linearly expanding or contracting each batch so all batches have the same range.
-We used the ComBat implementation from the R sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
-We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
-There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
-Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
-For these reasons, we compared Confounded only to the two methods listed above.
-Future batch adjustment research may benefit from standardization of input formats, user interfaces, and validation datasets.
-
-\subsection{Statistics and Metrics}
+\subsection{Statistics and metrics}
 
@@
 Mean squared error (MSE) is a measure of how much two vectors or matrices deviate from one another.
-It is commonly used as a loss value in autoencoders to make the network minimize the difference between the input and output values.
-We wanted to see how well Confounded and other batch correction software maintain patterns in the input data as measured by MSE.
+It is commonly used as to quantify loss when the objective (as with autoencoders) is to minimize the difference between input and output values.
 
@@
 Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
-We calculated MMD using the same formula as \citeauthor{shaham_removal_2017} to determine whether batches looked like they came from the same distribution after adjustment.
-For the kernel, we used the Gaussian kernel between two batches as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel} \cite{pedregosa_scikit-learn_2011}.
+We calculated MMD using the same formula to determine whether batches appear to come from the same distribution after adjustment.
+We used the Gaussian kernel as implemented in \texttt{sklearn.metrics.pairwise.rbf\_kernel} \cite{pedregosa_scikit-learn_2011}.
 In cases where there were more than two batches, we averaged all pairwise MMD values to calculate an overall MMD.
@@
 
-In order to determine \begin{enumerate*}[(a)]
-	\item whether batch can still be identified post-adjustment and
-	\item how well class-related signal is maintained after adjustment,
-\end{enumerate*}
-we determined classification accuracy based on batch and ``true class'' labels using several machine learning classifiers.
+To determine
+a) whether batch could still be identified post-adjustment and
+b) how well class-related signal was maintained after adjustment,
+we used classification algorithms to predict either batch or the ``true class`` labels based on the expression data and used classification accuracy as a metric.
+Table \ref{tab:datasets} details which columns were used for these analyses.
 
-We used four classifiers from the scikit-learn 0.19.1 Python library \cite{pedregosa_scikit-learn_2011} in order to classify on batch and true class before and after adjustment: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
-Table \ref{tab:datasets} details which columns were used for training.
+We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
+[@Say something about what hyperparameters we used. Defaults?@]
 
-We calculated the average of classification accuracies for four-fold cross-validation repeated three times.
-We interpret lower accuracy for batch classification as meaning that the batch is removed more effectively.
-We also interpret higher true class classification as meaning that the important signal is not lost during the process of adjustment.
-Therefore, given output data from the ideal batch adjuster, batch classification would be no better than random for any classification algorithm, and true class accuracy would be no lower than accuracy for the unadjusted data.
+We calculated the average classification accuracy across four-fold cross-validation repeated three times.
+We interpreted lower accuracy for batch classification to indicate that batch signal was removed more effectively.
+We interpreted higher true-class accuracy to indicate that biologically relevant signal was not lost during the batch-adjustment process.
+Therefore, given output data from a ideal batch adjuster, batch classification accuracy would be no better than random chance, and true-class accuracy would be equal to what would be obtained for the unadjusted data.
 

### 2019-07-04 thesis.tex
Fix error to make latex compile

@@
 2) the batches are sufficiently large (> 100 samples in our testing), and
-3) the batches are balanced.[@For #2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording.@] 
+3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording.@]
 These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.

### 2019-07-06 figures/code/combat.R
Add code for generating figures

@@ -1,0 +1,209 @@
+#Debugging: 
+options(error=function() { traceback(2); if(!interactive()) quit("no", status = 1, runLast = FALSE) })
+
+# Load dependencies --------------------------------
+
+load_stuff <- function() {
+  for (package in c("tidyverse", "docstring", "stringr", "argparse")) {
+    if (!require(package, character.only = TRUE)) { 
+      install.packages(package)
+      library(package, character.only = TRUE)
+    }
+  }
+  install_sva <- function() {
+    ## try http:// if https:// URLs are not supported
+    source("https://bioconductor.org/biocLite.R")
+    biocLite("sva")
+  }
+  if (!require("sva")) install_sva()
+  library(sva)
+}
+
+suppressMessages(load_stuff())
+
+# Parse command line args --------------------------
+
+parser <- ArgumentParser()
+
+parser$add_argument("input_file", help = "path to the input file")
+parser$add_argument("-a", "--adjuster", default = "combat", choices = c("combat", "scale"), help = "method to use for adjustment")
+parser$add_argument("-b", "--batch-col", default = "Batch", help = "title of batch column to adjust for")
+parser$add_argument("-o", "--output-file", help = "path to the output file. default: [input_file]_combat.csv")
+
+args <- parser$parse_args()
+
+if (is.null(args$output_file)) {
+  # /path/to/thing.csv -> /path/to/thing_combat.csv
+  args$output_file <- output_path <- str_replace_all(args$input_file, "(^.*)(\\.csv$)", sprintf("\\1_%s\\2", args$adjuster))
+}
+
+
+# Define functions ---------------------------------
+
+is.whole <- function(a, tol = 1e-7) { 
+  # Snatched from https://stat.ethz.ch/pipermail/r-help/2003-April/032471.html
+  is.eq <- function(x,y) { 
+    r <- all.equal(x,y, tol=tol)
+    is.logical(r) && r 
+  }
+  (is.numeric(a) && is.eq(a, floor(a))) ||
+    (is.complex(a) && {ri <- c(Re(a),Im(a)); is.eq(ri, floor(ri))})
+}
+
+varying_row_mask <- function(matrix_)
+{
+  #' Get a varying row mask
+  #' 
[…]

### 2019-07-06 figures/code/functions.R

@@ -1,0 +1,113 @@
+load_stuff <- function() {
+  for (package in c("tidyverse", "docstring")) {
+    if (!require(package, character.only = TRUE)) { 
+      install.packages(package)
+      library(package, character.only = TRUE)
+    }
+  }
+  install_sva <- function() {
+    ## try http:// if https:// URLs are not supported
+    source("https://bioconductor.org/biocLite.R")
+    biocLite("sva")
+  }
+  if (!require("sva")) install_sva()
+  library(sva)
+}
+suppressMessages(load_stuff())
+
+is.whole <- function(a, tol = 1e-7) { 
+  # Snatched from https://stat.ethz.ch/pipermail/r-help/2003-April/032471.html
+  is.eq <- function(x,y) { 
+    r <- all.equal(x,y, tol=tol)
+    is.logical(r) && r 
+  }
+  (is.numeric(a) && is.eq(a, floor(a))) ||
+    (is.complex(a) && {ri <- c(Re(a),Im(a)); is.eq(ri, floor(ri))})
+}
+
+varying_row_mask <- function(matrix_)
+{
+  #' Get a varying row mask
+  #' 
+  #' Get a boolean vector mask for the rows with & without varying values.
+  #'
+  #' @param matrix_ A numerical matrix.
+  #' 
+  #' @return Boolean vector.
+  #'
+  #' @examples
+  #' x <- matrix(c(2, 1, 3, 3, 1, 4, 4, 1, 5, 5, 1, 6), nrow=3, ncol=4)
+  #' x
+  #' ##      [,1] [,2] [,3] [,4]
+  #' ## [1,]    2    3    4    5
+  #' ## [2,]    1    1    1    1
+  #' ## [3,]    3    4    5    6
+  #' varying_row_mask(x)
+  #' ## [1]  TRUE FALSE  TRUE
+  matrix_ %>%
+    tbl_df() %>%
+    mutate_(min = min(names(.)), max = max(names(.))) %>%
+    mutate(nonvarying = max != min) %>%
+    .$nonvarying
+}
+
+remove_nonvarying_rows <- function(matrix_)
+{
+  #' Remove features that don't vary.
+  #'
+  #' ComBat requires that all inputs have some variance, so we need to remove
+  #' any nonvarying rows before passing a matrix into ComBat.
+  #'
+  #' @param matrix_ A numerical matrix.
+  #'
+  #' @return The original matrix without any nonvarying rows.
+  #' 
[…]

### 2019-07-06 thesis.tex
Add Dr. Piccolo's results comments

@@
 
-In this study, we created Confounded, an adversarial variational autoencoder neural network, to remove nonlinear batch and confounding effects from expression data that may not be accounted for by traditional linear methods.
-We compared Confounded to a scaling method and to ComBat \citep{johnson_adjusting_2007} using various metrics.
-The scaling method performed consistently worse across the qualitative and quantitative evaluations that we performed;
-ComBat and Confounded performed relatively well overall, but each of these algorithms excelled in different types of scenarios, which we illustrate below.
+In this study, we created Confounded, an adversarial, variational, autoencoder neural network, to remove nonlinear confounding effects from gene-expression data that may not be accounted for by linear methods.
+We compared Confounded to a scaling method and to ComBat \citep{johnson_adjusting_2007} and used qualitative and quantitative techniques to compare the performance of these methods.
+Overall, the scaling method performed consistently worse than the other methods;
+ComBat and Confounded each excelled in different scenarios, which we illustrate below.
 
@@
 
-To compare Confounded to other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).
+[@At least in my mind, we justified using MNIST as a way to visually see how our method performs and to modify it accordingly.
+However, the Results don't reflect that (in my mind) because the MNIST images are blurry for Confounded, whereas they are clear for the other methods.
+There is some explanation below about why this might be happening with Confounded.
+But as an outsider reading this, I would question why we used MNIST at all because our method didn't perform particularly well.
+It's hard to justify using MNIST anyway because it's so different from gene-expression data.
+So I think we should remove it from the paper (but it was fine to use it in your thesis).
+Let me know if you disagree.
+If not, please remove it.@]
 
-PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods for the GSE37199 dataset (see \figurename{s} \ref{fig:pca} and \ref{fig:tsne}).
-However, previous research has shown that these plots are not completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
+[@For simplicity, I think we should only include t-SNE graphs, not PCA. t-SNE accounts for nonlinear patterns, which is the point of our method.@]
+[@I also think we should have t-SNE graphs for all three datasets.]
+[@Please put in the full name for t-SNE and cite the paper for this method.@]
+We used principal component analysis (PCA) and t-Stochastic... (t-SNE) plots to visualize the data before and after batch adjustment.
+Before any adjustment, batches (microarray plates) in the GSE37199 dataset showed distinct expression patterns for each batch.
+After adjustment with each of the three methods, the batches were less distinct (see \figurename{s} \ref{fig:pca} and \ref{fig:tsne}).
 In the PCA plot, Confounded appears to maintain a similar distribution to the unadjusted data, indicating that the underlying distribution has been faithfully reproduced by the networks.
@@
 This may indicate an effective removal of nonlinear effects in both cases.
+%However, previous research has shown that these plots are not completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
 
@@
 
+%%%To compare Confounded against the other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).
 Confounded shows mixed success with the MSE and MMD metrics.

### 2019-07-11 figures/code/loss_chart.Rmd
Add all the code that was removed from Confounded

---
title: "Loss Charts"
author: "Jonathan Dayton"
date: "12/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman"); library(pacman)
p_load("tidyverse", "gridExtra", "png", "stringr", "magick", "colorspace", "pracma")
```

# Load data

```{r}
training_log_files <- dir("../../data/metrics/training/", pattern = ".*\\.csv", full.names = TRUE)
log_dfs <- do.call("rbind", lapply(training_log_files, read_csv))
```

# Look at the data

## Number of autoencoder layers doesn't seem to matter after 1

```{r}
final_iter <- log_dfs %>% group_by(start_time) %>% summarize(iteration = max(iteration)) %>% left_join(log_dfs)
final_iter %>% filter(str_detect(output_path, "ae_layers_")) %>%
  ggplot(aes(x = autoencoder_layers, y = dual_loss)) +
  geom_line() +
  labs(x = "Number of Autoencoder Layers", y = "Dual Loss Value")
```

## More layers in the discriminator generally means lower final loss

```{r}
final_iter %>% filter(str_detect(output_path, "/layers_")) %>%
  filter(discriminator_layers > 0) %>%
  ggplot(aes(x = discriminator_layers, y = dual_loss)) +
  geom_line() +
  labs(x = "Number of Discriminator Layers", y = "Dual Loss Value")
```

## The optimal code size seems to be between 1000 and 10,000

```{r}
final_iter %>% filter(str_detect(output_path, "/code_size")) %>%
  filter(discriminator_layers > 0) %>%
  ggplot(aes(x = code_size, y = dual_loss)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Nodes in Code Layer", y = "Dual Loss Value") +
  scale_x_log10()
```

```{r}
final_iter %>% filter(str_detect(output_path, "/scaling")) %>%
  ggplot(aes(x = scaling_method, y = dual_loss)) +
  geom_bar(stat = "identity") +
  labs(x = "Scaling Method", y = "Dual Loss Value")
```

[1] "../../data/metrics/training//ae_layers.csv"   "../../data/metrics/training//code_size.csv"  
[3] "../../data/metrics/training//features.csv"    "../../data/metrics/training//layers.csv"     
[5] "../../data/metrics/training//loss_weight.csv" "../../data/metrics/training//mnist.csv"      
[7] "../../data/metrics/training//scaling.csv"

```{r}

run <- log_dfs %>% filter(discriminator_layers == 9)

run %>%
  # filter(iteration > 4000) %>%
  ggplot(aes(x = iteration, y = dual_loss)) +
  geom_line() +
  geom_smooth() +
  labs(title = "Dual Loss", y = "Loss", x = "Iteration")

run %>%
  ggplot(aes(x = iteration, y = ae_loss)) +
  geom_line() +
  geom_smooth() +
  labs(title = "Autoencoder Loss", y = "Loss", x = "Iteration")

run %>%
  ggplot(aes(x = iteration, y = disc_loss)) +
  geom_line() +
  geom_smooth() +
  labs(title = "Discriminator Loss", y = "Loss", x = "Iteration")
```

### 2019-07-11 code/metrics/calc/mmd.py

@@ -1,0 +1,101 @@
+import pandas as pd
+import numpy as np
+from sklearn.metrics.pairwise import rbf_kernel, linear_kernel
+
+def split_discrete_continuous(df):
+    discrete_types = ["int", "object"]
+    discrete = df.select_dtypes(include=discrete_types)
+    continuous = df.select_dtypes(exclude=discrete_types)
+    return discrete, continuous
+
+def mmd(sample1, sample2, kernel=rbf_kernel):
+    # TODO: update the kernel to reflect the DMResNet paper:
+    """ "the kernel we used is a sum of three Gaussian kernels with
+    different scales...We chose the [sigma_i]s to be m/2,m,2m, where m
+    is the median of the average distance between a point in the
+    target sample to its nearest 25 neighbors, a popular practice in
+    kernel-based methods."
+    - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5870543/
+    """
+    k_x_x = kernel(sample1, sample1)
+    k_x_y = kernel(sample1, sample2)
+    k_y_y = kernel(sample2, sample2)
+    n = len(sample1)
+    m = len(sample2)
+    mean_xx = (1 / n**2) * k_x_x.sum()
+    mean_xy = (1 / (m * n)) * k_x_y.sum()
+    mean_yy = (1 / m**2) * k_y_y.sum()
+    mmd_squared = mean_xx - 2 * mean_xy + mean_yy
+    return mmd_squared ** 0.5
+
+def split_into_batches(df, batch_col):
+    discrete, continuous = split_discrete_continuous(df)
+    batches = set(df[batch_col])
+    return tuple((continuous[df[batch_col] == batch] for batch in batches))
+
+def mmd_multi_batch(batches):
+    mmds = []
+    for i in range(len(batches)):
+        for j in range(i+1, len(batches)):
+            mmds.append(mmd(batches[i], batches[j]))
+    return np.array(mmds).mean()
+
+def calculate_mmd(df, batch_col, log_adjust=False):
+    batches = split_into_batches(df, batch_col)
+    if log_adjust:
+        batches = tuple([log_scale(batch) for batch in batches])
+    return mmd_multi_batch(batches)
+
+def log_scale(df):
+    # Get rid of negative values
+    df = df.where(df.min() < 0, df - df.min())
+    return np.log(df + 1.0)
+
+class Logger(object):
+    def __init__(self, metric):
+        self.metric = metric
+        self.values = {
+            "metric": [],
+            "adjuster": [],
[…]

### 2019-07-11 code/metrics/calc/mse.py

@@ -1,0 +1,26 @@
+import pandas as pd
+import numpy as np
+from .mmd import split_discrete_continuous, Logger, DataFrameCache, COMPARISONS_PATH, log_scale
+
+def calculate_mse(df1, df2, log_adjust=False):
+    _, genes1 = split_discrete_continuous(df1)
+    _, genes2 = split_discrete_continuous(df2)
+    if log_adjust:
+        genes1 = log_scale(genes1)
+        genes2 = log_scale(genes2)
+    squared_error = (np.array(genes1) - np.array(genes2))**2
+    return squared_error.mean()
+
+if __name__ == "__main__":
+    cache = DataFrameCache()
+    comparisons = cache.get_dataframe(COMPARISONS_PATH)
+    logger = Logger("MSE")
+
+    for i, row in comparisons.iterrows():
+        print(row.dataset)
+        df = cache.get_dataframe(row["path"])
+        original = cache.get_dataframe(row["unadjusted"])
+        value = calculate_mse(df, original, log_adjust=("TCGA" in row.dataset))
+        logger.log(row["adjuster"], row["dataset"], value)
+
+    logger.save("./data/metrics/mse.csv")

### 2019-07-11 code/metrics/classifiers/csvread.py

@@ -1,0 +1,40 @@
+"""Read a CSV file in the format scikit-learn requires.
+
+The CSV should follow the tidy data format (i.e. one sample per row, one
+feature per column) and should have the following columns:
+
+- Sample: a unique identifier for each row
+- Batch: the batch to which each sample belongs
+- Other columns specified as `meta_cols`
+
+Every other column should be a quantitative feature.
+"""
+
+
+import pandas as pd
+from sklearn.model_selection import train_test_split
+
+class CSVData(object):
+    def __init__(self, path, meta_cols=None, predict="Batch"):
+        df = pd.read_csv(path)
+        if meta_cols is None:
+            meta_cols = list(df.select_dtypes(include=['object', 'int']).columns)
+        self.labels = df[predict]
+        self.features = df.drop(meta_cols, axis="columns")
+        self.X_train, self.X_test, self.Y_train, self.Y_test = train_test_split(
+            self.features, self.labels, test_size=0.2
+        )
+
+    @property
+    def X(self):
+        return self.features
+
+    @property
+    def Y(self):
+        return self.labels
+
+if __name__ == "__main__":
+    data = CSVData("test_data.csv")
+    print(data.X)
+    print()
+    print(data.Y)

### 2019-07-11 code/metrics/classifiers/random_forests.py

@@ -1,0 +1,35 @@
+"""Random forests classifier from scikit learn
+"""
+
+import time
+from sklearn.ensemble import RandomForestClassifier
+from sklearn.model_selection import cross_val_score
+from .csvread import CSVData
+
+def cross_validate(path, meta_cols=None, predict="Batch", times=100, folds=5, model=RandomForestClassifier, **kwargs):
+    classifier = model(**kwargs)
+    data = CSVData(path, meta_cols, predict)
+    accuracies = []
+    elapsed_times = []
+    for _ in range(times):
+        start = time.time()
+        try:
+            accuracies += list(cross_val_score(
+                classifier, data.X, data.Y, cv=folds, scoring="accuracy"
+            ))
+            elapsed_times.append(time.time() - start)
+        except ValueError as e:
+            print("Something didn't work with file {}, column {}, and classifier {}:".format(
+                path, predict, model.__name__)
+            )
+            print(str(e))
+    return {
+        "accuracies": accuracies,
+        "times": elapsed_times
+    }
+
+if __name__ == "__main__":
+    INPUT_PATH = "../../data/tidy_combat.csv"
+    print(
+        cross_validate(INPUT_PATH, times=5)
+    )

### 2019-07-11 code/metrics/classifiers/true_class.py

@@ -1,0 +1,91 @@
+"""Predict the true class of MNIST digits with a SVC classifier.
+"""
+import pandas as pd
+from sklearn.svm import SVC
+from sklearn.ensemble import RandomForestClassifier, GradientBoostingClassifier
+from sklearn.neural_network import MLPClassifier
+from sklearn.naive_bayes import GaussianNB
+from sklearn.neighbors import KNeighborsClassifier
+from .random_forests import cross_validate
+from ..calc.mmd import COMPARISONS_PATH, DataFrameCache
+
+class Logger(object):
+    def __init__(self, log_file):
+        self.log_file = log_file
+        self.values = {
+            "path": [],
+            "model": [],
+            "predict": [],
+            "dataset": [],
+            "col_type": [],
+            "baseline": [],
+            "iteration": [],
+            "accuracy": [],
+            "time_elapsed": [],
+            "adjuster": [],
+        }
+
+    def log(self, path, model, predict_col, dataset, col_type, baseline, accuracies, times_elapsed, adjuster):
+        for i, (accuracy, time) in enumerate(zip(accuracies, times_elapsed)):
+            self.values["path"].append(path)
+            self.values["model"].append(model.__name__)
+            self.values["predict"].append(predict_col)
+            self.values["dataset"].append(dataset)
+            self.values["col_type"].append(col_type)
+            self.values["baseline"].append(baseline)
+            self.values["iteration"].append(i)
+            self.values["accuracy"].append(accuracy)
+            self.values["time_elapsed"].append(time)
+            self.values["adjuster"].append(adjuster)
+
+    def save(self):
+        pd.DataFrame(self.values).to_csv(self.log_file, index=False)
+
+
+def baseline(path, column, cache=None):
+    try:
+        df = cache.get_dataframe(path)
+    except AttributeError:
+        df = pd.read_csv(path)
+    return df[column].value_counts().max() / len(df)
+
+if __name__ == "__main__":
+    cache = DataFrameCache()
+    comparisons = cache.get_dataframe(COMPARISONS_PATH)
+    LEARNERS = [
+        (RandomForestClassifier, {"n_estimators": 10}),
+        # (MLPClassifier, {"hidden_layer_sizes": tuple([5]*10), "max_iter": 1000}),
+        (GaussianNB, {}),
+        # (KNeighborsClassifier, {}),
+        # (SVC, {"kernel": "rbf"})
+    ]
+
[…]

### 2019-07-11 code/other/autoencoder.py

@@ -1,0 +1,314 @@
+# pylint: disable=E1129,E0611,E1101
+import argparse
+import datetime
+import random
+from tqdm import tqdm
+from . import hide_warnings
+import tensorflow as tf
+import pandas as pd
+from . import reformat
+from .load_data import split_features_labels, list_categorical_columns
+from .adjustments import Scaler, SigmoidScaler, split_discrete_continuous
+from .network import Confounded
+
+
+def parse_arguments():
+    activation = tf.nn.relu
+
+    def positive_int(value):
+        ivalue = int(value)
+        if ivalue < 0:
+            raise argparse.ArgumentTypeError("%s is an invalid positive int value" % value)
+        return ivalue
+
+    parser = argparse.ArgumentParser()
+    parser.add_argument(
+        'file', metavar='source-file', type=str,
+        help='Path to input file.')
+    parser.add_argument(
+        '-o', "--output-file", type=str,
+        help="Path to output file.")
+    parser.add_argument(
+        "-m", "--minibatch-size", type=positive_int, default=100,
+        help="The size of the mini-batch for training. Must be positive integer.")
+    parser.add_argument(
+        "-l", "--layers", type=positive_int, default=10,
+        help="How many layers deep the discriminator should be. Must be positive integer.")
+    parser.add_argument(
+        "-a", "--ae-layers", type=positive_int, default=2,
+        help="How many layers in each of the encoding and decoding portions of the autoencoder.")
+    parser.add_argument(
+        "-b", "--batch-col", type=str, default="Batch",
+        help="Which column contains the batch to adjust for.")
+    parser.add_argument(
+        "-c", "--code-size", type=positive_int, default=20,
+        help="How many nodes in the code layer of the autoencoder.")
+    parser.add_argument(
+        "-e", "--early-stopping", type=positive_int, default=None,
+        help="How many iterations without improvement before stopping early. Default: None")
+    parser.add_argument(
+        "-s", "--scaling", choices=["linear", "sigmoid"], default="linear",
+        help="Type of scaling to perform on the input data.")
+    parser.add_argument(
+        "-w", "--loss-weight", type=float, default=1.0,
+        help="Weight applied to the discriminator loss when training the autoencoder.")
+    parser.add_argument(
+        "-f", "--log-file", type=str, default="./data/metrics/log.csv",
+        help="Path to file to log results.")
+    parser.add_argument(
+        "-i", "--iterations", type=positive_int, default=10000,
+        help="Number of iterations of minibatches to run.")
[…]

### 2019-07-11 code/other/load_data.py

@@ -1,0 +1,75 @@
+import numpy as np
+import pandas as pd
+import gzip
+
+class RNASeq(object):
+    def __init__(self):
+        self.data = self.load_data()
+        self.cur_ix = 0
+
+    def load_data(self):
+        data = []
+        with gzip.open("./data/RNASeq.txt.gz", "r") as infile:
+            header = infile.readline()
+            for line in infile:
+                line = line.strip("\n").split("\t")[1:]
+                line = [float(x) for x in line]
+                data.append(line)
+        return np.array(data)
+
+    def next_batch(self, batch_size):
+        if batch_size > len(self.data):
+            raise ValueError("Argument `batch_size` greater than length of dataset.")
+        if self.cur_ix + batch_size > len(self.data):
+            np.random.shuffle(self.data)
+            self.cur_ix = 0
+        batch = self.data[self.cur_ix:self.cur_ix + batch_size]
+        self.cur_ix += batch_size
+        return batch
+
+def split_features_labels(df, batch_col, meta_cols=None, sample=None):
+    """Split a dataframe into features and labels numpy arrays.
+
+    Arguments:
+        df {pandas.DataFrame} -- A tidy dataframe with meta columns, a
+            Batch column, and quantitative data.
+
+    Keyword Arguments:
+        meta_cols {list of strings} -- Columns that should not be
+            used as features to be batch-adjusted (default:
+            {["Sample", `batch_col`]})
+        sample {int} -- The number of rows to sample. If None, return
+            all rows. (default: {None})
+
+    Returns:
+        [(numpy.array, numpy.array)] -- Tuple of features and labels,
+            where features are the quantitative data from the given
+            dataframe and labels are the one-hot encoded batches for
+            each instance.
+    """
+    if meta_cols is None:
+        meta_cols = ["Sample", batch_col]
+    features = np.array(df.drop(meta_cols, axis=1))
+    labels = pd.get_dummies(df[batch_col])
+    labels = np.array(labels, dtype=float)
+    if sample is not None:
+        rows = pd.DataFrame(df.index).sample(sample, replace=True)[0].values.tolist()
+        features = features[rows]
+        labels = labels[rows]
+    return features, labels
+
+def list_categorical_columns(df):
+    """Get the names of all categorical columns in the dataframe.
+
+    Arguments:
+        df {pandas.DataFrame} -- The dataframe.
+
+    Returns:
[…]

### 2019-07-11 code/other/matrix.py

@@ -1,0 +1,49 @@
+#!/usr/bin/env python3
+
+from . import png
+from .adjustments import Noise
+from . import reformat
+import numpy as np
+import pandas as pd
+import glob
+
+def pngs2matrix(directory):
+    paths = glob.glob(directory + "/**/*.png", recursive=True)
+    n_pixels = len(png.png2array(paths[0]).flatten())
+    n_pngs = len(paths)
+    noiser1 = Noise((n_pixels,), order=2)
+    noiser2 = Noise((n_pixels,), order=2)
+    # ComBat needs n_probes x n_samples
+    nonadjusted = np.ndarray((n_pixels, n_pngs))
+    adjusted = np.ndarray((n_pixels, n_pngs))
+    for i, path in enumerate(paths):
+        pixels = png.png2array(path).flatten()
+        nonadjusted[:, i] = pixels
+        if i % 2 == 0:
+            pixels = noiser1.adjust(pixels)
+        else:
+            pixels = noiser2.adjust(pixels)
+        adjusted[:, i] = pixels
+    return nonadjusted, adjusted, paths
+
+def unflatten(array):
+    return np.reshape(array, (28, 28))
+
+if __name__ == "__main__":
+    directory = '/home/jdayton3/Downloads/mnist_png/testing'
+    nonadjusted, adjusted, paths = pngs2matrix(directory)
+    df = pd.DataFrame(adjusted)
+    df.to_csv("./data/mnist_matrix.csv", index=False, header=False)
+
+    df2 = pd.DataFrame(nonadjusted)
+    reformat.to_csv(df2, "./data/tidy_nonadjusted.csv", meta_cols={
+        "Batch": ['A'] * 10000,
+        "Sample": df.T.index,
+        "Digit": [path.split('/')[-2] for path in paths]
+    })
+
+    reformat.to_csv(df, "./data/tidy_batches_balanced.csv", meta_cols={
+        "Batch": ['A', 'B'] * 5000,
+        "Sample": df.T.index,
+        "Digit": [path.split('/')[-2] for path in paths]
+    })

### 2019-07-11 code/other/network.py

@@ -1,0 +1,219 @@
+"""Definitions for the neural networks in Confounded.
+"""
+
+# pylint: disable=E1129
+
+import functools
+import tensorflow as tf
+from tensorflow.contrib.layers import fully_connected, batch_norm # pylint: disable=E0611
+from math import ceil
+
+LEARNING_RATE = 0.0001
+
+def is_square(n):
+    sqrt = n**0.5
+    return int(sqrt) == sqrt
+
+def var_scope(scope):
+    """Decorator to wrap a function in a tensorflow variable scope
+
+    Arguments:
+        scope {str} -- Name of the variable scope
+
+    Returns:
+        function -- The decorated function wrapped in the variable scope
+    """
+    def decorator_var_scope(func):
+        @functools.wraps(func)
+        def wrapper_var_scope(*args, **kwargs):
+            with tf.variable_scope(scope):
+                return func(*args, **kwargs)
+        return wrapper_var_scope
+    return decorator_var_scope
+
+def show_before_and_after_images(func):
+    @functools.wraps(func)
+    def wrapper_show_images(*args, **kwargs):
+        inputs = args[0]
+        show_image(inputs, name="inputs")
+        outputs = func(*args, **kwargs)
+        if isinstance(outputs, tuple):
+            # The autoencoder functions might return (outputs, loss)
+            show_image(outputs[0], name="outputs")
+        else:
+            show_image(outputs, name="outputs")
+        return outputs
+    return wrapper_show_images
+
+@var_scope("vae")
+@show_before_and_after_images
+def variational_autoencoder(inputs, code_size=20):
+    """Creates a variational autoencoder based on "Hands-On Machine
+    Learning with Scikit-Learn and TensorFlow by Aurélien Géron
+    (O’Reilly). Copyright 2017 Aurélien Géron, 978-1-491-96229-9."
+
+    Arguments:
+        input_size {int} -- Size of the input to the autoencoder
+
+    Returns:
+        input {Tensor} -- The input tensor
+        output {Tensor} -- The output tensor
+        loss {Tensor} -- The loss operation
+    """
+    layer_sizes = [500, 500]
+    activations = [tf.nn.elu for _ in layer_sizes]
+    input_size = layer_size(inputs)
+
+    encoding = make_layers(inputs, layer_sizes, activations)
+    code_mean, code_gamma, code = vae_code_layer(encoding, code_size)
[…]

### 2019-07-11 code/other/reformat.py

@@ -1,0 +1,58 @@
+import pandas as pd
+
+def to_csv(df, path, tidy=False, meta_cols=None):
+    """Save a dataframe as a CSV, and optionally add meta columns and
+        transpose the dataframe.
+
+    Arguments:
+        df {pandas.DataFrame} -- The data to be saved.
+        path {str} -- The path where the csv should be saved.
+
+    Keyword Arguments:
+        tidy {bool} -- True if the rows are instances and the columns
+            are features; False otherwise. (default: {False})
+        meta_cols {dict} -- Dictionary of {column_name: [values, ...]}
+            (default: {None})
+    """
+    if not tidy:
+        df = df.T
+    df = _add_meta_cols(df, meta_cols)
+    df.to_csv(path, index=False)
+
+def _add_meta_cols(df, meta_cols):
+    if meta_cols is None:
+        meta_cols = {}
+    sample = meta_cols.get("Sample")
+    df['Sample'] = _check_sample(df, sample)
+    for col_name, col in meta_cols.items():
+        if col_name in ["Sample"]:
+            continue
+        df[col_name] = col
+    return _reorder_cols(df, list(meta_cols.keys()))
+
+def _check_sample(df, sample):
+    if sample is None:
+        return df.index
+    return sample
+
+def _reorder_cols(df, meta_cols):
+    cols = df.columns.tolist()
+    for col in meta_cols:
+        cols.remove(col)
+    cols = meta_cols + cols
+    return df[cols]
+
+if __name__ == "__main__":
+    import functools
+    import operator
+    in_path = "data/mnist_matrix.csv"
+    out_path = "data/tidy_mnist2.csv"
+    df = pd.read_csv(in_path, header=None)
+    to_csv(df, out_path, tidy=False, meta_cols={
+        "Batch": None,
+        "Sample": None,
+        "Digit": functools.reduce(
+            operator.add,
+            [[x] * 1000 for x in range(10)]
+        )
+    })

### 2019-07-11 code/scripts/code_size.sh

@@ -1,0 +1,23 @@
+input_folder="./data/avery/GSE37199/"
+output_folder="./data/output/code_size/"
+input_path="${input_folder}clinical.csv"
+
+mkdir -p $output_folder
+
+log_folder="./data/metrics/"
+mkdir -p $log_folder
+
+log_file="${log_folder}code_size.csv"
+
+source activate py36
+python --version
+
+for i in 10 50 100 500 1000 5000 10000 20000
+do
+    printf -v padded "%05d" $i
+    python -m src.autoencoder $input_path \
+        --output-file ${output_folder}size_${padded}_plate.csv \
+        --code-size $i \
+        --batch-col "plate" \
+        --log-file $log_file
+done

### 2019-07-11 code/scripts/features.sh

@@ -1,0 +1,21 @@
+data_dir="./data/feature_selection/"
+output_dir="./data/output/features/"
+
+mkdir -p $output_dir
+
+log_folder="./data/metrics/"
+mkdir -p $log_folder
+
+log_file="${log_folder}features.csv"
+
+source activate py36
+python --version
+
+for path in $(ls $data_dir)
+do
+    for column in 'centre' 'plate'
+    do
+        python -m src.autoencoder ${data_dir}${path} -o ${output_dir}${path} -l 10 -b $column \
+            --log-file $log_file
+    done
+done

### 2019-07-11 code/scripts/mnist.py

@@ -1,0 +1,36 @@
+import numpy as np
+import pandas as pd
+from PIL import Image
+
+def array2png(array, path):
+    # Copied from png.py :(
+    img = Image.fromarray((255 * (1.0 - array)).astype(np.uint8), 'L')
+    img.save(path)
+
+if __name__ == "__main__":
+    unadjusted = pd.read_csv("./data/mnist/unadjusted.csv")
+    noisy = pd.read_csv("./data/mnist/noisy.csv")
+    combat = pd.read_csv("./data/mnist/noisy_combat.csv")
+    confounded = pd.read_csv("./data/mnist/noisy_confounded.csv")
+
+    dataframes = [
+        unadjusted,
+        noisy,
+        combat,
+        confounded
+    ]
+
+    names = [
+        "01_unadjusted",
+        "02_noisy",
+        "03_combat",
+        "04_confounded"
+    ]
+
+    indices = [0, 1500, 3000, 4500]
+
+    for df, name in zip(dataframes, names):
+        continuous = df.select_dtypes(exclude=["object", "int"])
+        for i in indices:
+            pixels = np.array(continuous.iloc[i]).reshape((28, 28))
+            array2png(pixels, "./data/output/mnist/{}_{:04d}.png".format(name, i))

### 2019-07-11 code/scripts/mnist.sh

@@ -1,0 +1,26 @@
+input_path="./data/mnist/noisy.csv"
+output_dir="./data/output/mnist/"
+output_path="${output_dir}mnist.csv"
+
+mkdir -p $output_dir
+
+log_folder="./data/metrics/"
+mkdir -p $log_folder
+
+log_file="${log_folder}mnist.csv"
+
+source activate py36
+python --version
+
+
+python -m src.autoencoder $input_path \
+    --output-file $output_path \
+    --layers 10 \
+    --ae-layers 2 \
+    --batch-col "Batch" \
+    --code-size 200 \
+    --scaling linear \
+    --loss-weight 1 \
+    --log-file $log_file
+
+bash "../scripts/mnist.py"

### 2019-07-11 code/scripts/pretrain.sh

@@ -1,0 +1,8 @@
+time python -m src.autoencoder data/avery/GSE37199/clinical.csv -o
+data/avery/GSE37199/clinical_pretrained_10000_confounded.csv -b plate -r data/gpl570/checkpoints/weights.ckpt -i 10000
+time python -m src.autoencoder data/avery/GSE37199/clinical.csv -o
+data/avery/GSE37199/clinical_pretrained_1000_confounded.csv -b plate -r data/gpl570/checkpoints/weights.ckpt -i 1000
+time python -m src.autoencoder data/avery/GSE37199/clinical.csv -o
+data/avery/GSE37199/clinical_pretrained_100_confounded.csv -b plate -r data/gpl570/checkpoints/weights.ckpt -i 100
+time python -m src.autoencoder data/avery/GSE37199/clinical.csv -o
+data/avery/GSE37199/clinical_pretrained_confounded.csv -b plate -r data/gpl570/checkpoints/weights.ckpt -i 0

### 2019-07-11 code/prepdata/all.sh
Add scripts to download and tidy all the data

@@ -1,0 +1,6 @@
+#!/bin/bash
+
+bash mnist.sh
+bash bladderbatch.sh
+bash gse37199.sh
+bash tcga.sh

### 2019-07-11 code/prepdata/bladderbatch.R

@@ -1,24 +1,16 @@
----
-title: "Bladder Batch to tidy CSV"
-author: "Jonathan Dayton"
-date: "1/3/2019"
-output: html_document
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
 if (!require("pacman")) install.packages("pacman"); library(pacman)
-p_load("tidyverse")
+p_load("tidyverse", "argparse")
+
+parser <- ArgumentParser()
+parser$add_argument("outfile", help = "Path to the output file")
+args <- parser$parse_args()
+
 if (!requireNamespace("BiocManager", quietly = TRUE))
     install.packages("BiocManager")
 BiocManager::install("bladderbatch")
 library(bladderbatch)
 data(bladderdata)
-```
 
-# Turn it into a tibble
-
-```{r}
 clin_df <- bladderEset %>%
   pData() %>%
   as.tibble(rownames = "id")
@@ -29,5 +21,4 @@ gene_df <- bladderEset %>%
 
 all <- inner_join(clin_df, gene_df, by = "id")
 
-write_csv(all, "../../data/bladderbatch/bladderbatch.csv")
-```
+write_csv(all, args$outfile)

### 2019-07-11 code/prepdata/bladderbatch.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+dest="../data/bladderbatch/"
+out_csv="${dest}/bladderbatch.csv"
+
+mkdir -p $dest
+
+Rscript bladderbatch.R $out_csv

### 2019-07-11 code/prepdata/gse37199.sh

@@ -1,0 +1,17 @@
+#!/bin/bash
+
+dest="../data/gse37199/"
+raw="${dest}/raw/"
+raw_loc="${raw}/raw.tar.gz"
+expression_tsv="${raw}/GSE37199_Expression_BatchUnadjusted.txt.gz"
+clinical_tsv="${raw}/GSE37199_Clinical.txt"
+out_csv="${dest}/gse37199.csv"
+
+mkdir -p $raw
+
+if [ ! -f $raw_loc ]; then
+    wget https://osf.io/av3yt/download -O $raw_loc
+fi
+tar -zxvf $raw_loc -C $raw
+
+python gse37199.py $expression_tsv $clinical_tsv $out_csv

### 2019-07-11 code/prepdata/mnist.py

@@ -1,0 +1,16 @@
+import pandas as pd
+import argparse
+
+parser = argparse.ArgumentParser()
+parser.add_argument("infile", help="Path to the input file")
+parser.add_argument("outfile", help="Path to the output file")
+args = parser.parse_args()
+
+df = pd.read_csv(args.infile, header=None, names=["Digit"] + list(range(28**2)))
+df.set_index("Digit")\
+    .apply(lambda row: 1 - (row / 255.0))\
+    .sort_values("Digit")\
+    .reset_index()\
+    .reset_index()\
+    .rename(columns={"index": "Sample"})\
+    .to_csv(args.outfile, index=False)

### 2019-07-11 code/prepdata/mnist.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+
+dest="../data/mnist/"
+raw_csv="${dest}/raw.csv"
+tidy_csv="${dest}/unadjusted.csv"
+batched_csv="${dest}/noisy.csv"
+
+mkdir -p $dest
+
+if [ ! -f $raw_csv ]; then
+    wget https://pjreddie.com/media/files/mnist_test.csv -O $raw_csv
+fi
+
+python mnist.py $raw_csv $tidy_csv
+python artificial_batch.py $tidy_csv $batched_csv

### 2019-07-11 code/prepdata/tcga.py

@@ -1,0 +1,33 @@
+import argparse
+import pandas as pd
+
+def maybe_to_int(col):
+    try:
+        return int(col)
+    except:
+        return col
+
+def to_sparse_matrix(df):
+    df["one"] = 1
+    df["MutatedGene"] = df["MutatedGene"].apply(lambda x: str(x) + "_mutation")
+    matrix = df.pivot(index="Sample", columns="MutatedGene", values="one").fillna(0)
+    small_mx = matrix.loc[:, matrix.columns[matrix.sum() > 2000]]
+    return small_mx
+
+parser = argparse.ArgumentParser()
+parser.add_argument("rnaseq", help="Path to RNA Seq data")
+parser.add_argument("mutations", help="Path to somatic mutations data")
+parser.add_argument("cancer_types", help="Path to cancer types data")
+parser.add_argument("outfile", help="Path to outfile")
+args = parser.parse_args()
+
+mutations = pd.read_csv(args.mutations, sep="\t")
+ctypes = pd.read_csv(args.cancer_types, sep="\t")
+expression = pd.read_csv(args.rnaseq, sep="\t")
+
+mutations_matrix = to_sparse_matrix(mutations)
+clinical = ctypes.merge(mutations_matrix, on="Sample", how="left").fillna(0)
+clinical = clinical.applymap(maybe_to_int)
+all_data = clinical.merge(expression, on="Sample", how="inner").drop_duplicates()
+
+all_data.to_csv(args.outfile, index=False)

### 2019-07-11 code/prepdata/tcga.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+
+dest="../data/tcga/"
+rnaseq="${dest}/rnaseq.tsv.gz"
+mutations="${dest}/mutations.tsv.gz"
+labels="${dest}/labels.tsv.gz"
+final="${dest}/tcga.csv"
+
+mkdir -p $dest
+
+if [ ! -f $rnaseq ]; then
+    wget https://osf.io/7xjdn/download -O $rnaseq
+fi
+if [ ! -f $mutations ]; then
+    wget https://osf.io/na3rp/download -O $mutations
+fi
+if [ ! -f $labels ]; then
+    wget https://osf.io/frxv6/download -O $labels
+fi
+
+python tcga.py $rnaseq $mutations $labels $final

### 2019-07-15 code/metrics/classify.py
Get classification metrics working

@@ -1,0 +1,149 @@
+"""Predict the true class of MNIST digits with a SVC classifier.
+"""
+import pandas as pd
+from sklearn.svm import SVC
+from sklearn.ensemble import RandomForestClassifier, GradientBoostingClassifier
+from sklearn.neural_network import MLPClassifier
+from sklearn.naive_bayes import GaussianNB
+from sklearn.neighbors import KNeighborsClassifier
+from sklearn.model_selection import cross_val_score
+from util import DataFrameCache, get_dataset_path_dict, no_extension
+import time
+from sklearn.model_selection import train_test_split
+
+class CSVData(object):
+    def __init__(self, path, meta_cols=None, predict="Batch"):
+        df = pd.read_csv(path)
+        if meta_cols is None:
+            meta_cols = list(df.select_dtypes(include=['object', 'int']).columns)
+        self.labels = df[predict]
+        self.features = df.drop(meta_cols, axis="columns")
+        self.X_train, self.X_test, self.Y_train, self.Y_test = train_test_split(
+            self.features, self.labels, test_size=0.2
+        )
+
+    @property
+    def X(self):
+        return self.features
+
+    @property
+    def Y(self):
+        return self.labels
+
+def cross_validate(path, meta_cols=None, predict="Batch", times=100, folds=5, model=RandomForestClassifier, **kwargs):
+    classifier = model(**kwargs)
+    data = CSVData(path, meta_cols, predict)
+    accuracies = []
+    elapsed_times = []
+    for _ in range(times):
+        start = time.time()
+        try:
+            accuracies += list(cross_val_score(
+                classifier, data.X, data.Y, cv=folds, scoring="accuracy"
+            ))
+            elapsed_times.append(time.time() - start)
+        except ValueError as e:
+            print("Something didn't work with file {}, column {}, and classifier {}:".format(
+                path, predict, model.__name__)
+            )
+            print(str(e))
+    return {
+        "accuracies": accuracies,
+        "times": elapsed_times
+    }
+
+class Logger(object):
+    def __init__(self, log_file):
+        self.log_file = log_file
+        self.values = {
+            "path": [],
+            "model": [],
+            "predict": [],
+            "dataset": [],
+            "col_type": [],
+            "baseline": [],
+            "iteration": [],
+            "accuracy": [],
+            "time_elapsed": [],
+            "adjuster": [],
+        }
+
+    def log(self, path, model, predict_col, dataset, col_type, baseline, accuracies, times_elapsed, adjuster):
[…]

### 2019-07-15 code/other/autoencoder.py
Remove or rename unused code

@@ -1,314 +1,0 @@
-# pylint: disable=E1129,E0611,E1101
-import argparse
-import datetime
-import random
-from tqdm import tqdm
-from . import hide_warnings
-import tensorflow as tf
-import pandas as pd
-from . import reformat
-from .load_data import split_features_labels, list_categorical_columns
-from .adjustments import Scaler, SigmoidScaler, split_discrete_continuous
-from .network import Confounded
-
-
-def parse_arguments():
-    activation = tf.nn.relu
-
-    def positive_int(value):
-        ivalue = int(value)
-        if ivalue < 0:
-            raise argparse.ArgumentTypeError("%s is an invalid positive int value" % value)
-        return ivalue
-
-    parser = argparse.ArgumentParser()
-    parser.add_argument(
-        'file', metavar='source-file', type=str,
-        help='Path to input file.')
-    parser.add_argument(
-        '-o', "--output-file", type=str,
-        help="Path to output file.")
-    parser.add_argument(
-        "-m", "--minibatch-size", type=positive_int, default=100,
-        help="The size of the mini-batch for training. Must be positive integer.")
-    parser.add_argument(
-        "-l", "--layers", type=positive_int, default=10,
-        help="How many layers deep the discriminator should be. Must be positive integer.")
-    parser.add_argument(
-        "-a", "--ae-layers", type=positive_int, default=2,
-        help="How many layers in each of the encoding and decoding portions of the autoencoder.")
-    parser.add_argument(
-        "-b", "--batch-col", type=str, default="Batch",
-        help="Which column contains the batch to adjust for.")
-    parser.add_argument(
-        "-c", "--code-size", type=positive_int, default=20,
-        help="How many nodes in the code layer of the autoencoder.")
-    parser.add_argument(
-        "-e", "--early-stopping", type=positive_int, default=None,
-        help="How many iterations without improvement before stopping early. Default: None")
-    parser.add_argument(
-        "-s", "--scaling", choices=["linear", "sigmoid"], default="linear",
-        help="Type of scaling to perform on the input data.")
-    parser.add_argument(
-        "-w", "--loss-weight", type=float, default=1.0,
-        help="Weight applied to the discriminator loss when training the autoencoder.")
-    parser.add_argument(
-        "-f", "--log-file", type=str, default="./data/metrics/log.csv",
-        help="Path to file to log results.")
-    parser.add_argument(
-        "-i", "--iterations", type=positive_int, default=10000,
-        help="Number of iterations of minibatches to run.")
[…]

### 2019-07-15 code/other/load_data.py

@@ -1,75 +1,0 @@
-import numpy as np
-import pandas as pd
-import gzip
-
-class RNASeq(object):
-    def __init__(self):
-        self.data = self.load_data()
-        self.cur_ix = 0
-
-    def load_data(self):
-        data = []
-        with gzip.open("./data/RNASeq.txt.gz", "r") as infile:
-            header = infile.readline()
-            for line in infile:
-                line = line.strip("\n").split("\t")[1:]
-                line = [float(x) for x in line]
-                data.append(line)
-        return np.array(data)
-
-    def next_batch(self, batch_size):
-        if batch_size > len(self.data):
-            raise ValueError("Argument `batch_size` greater than length of dataset.")
-        if self.cur_ix + batch_size > len(self.data):
-            np.random.shuffle(self.data)
-            self.cur_ix = 0
-        batch = self.data[self.cur_ix:self.cur_ix + batch_size]
-        self.cur_ix += batch_size
-        return batch
-
-def split_features_labels(df, batch_col, meta_cols=None, sample=None):
-    """Split a dataframe into features and labels numpy arrays.
-
-    Arguments:
-        df {pandas.DataFrame} -- A tidy dataframe with meta columns, a
-            Batch column, and quantitative data.
-
-    Keyword Arguments:
-        meta_cols {list of strings} -- Columns that should not be
-            used as features to be batch-adjusted (default:
-            {["Sample", `batch_col`]})
-        sample {int} -- The number of rows to sample. If None, return
-            all rows. (default: {None})
-
-    Returns:
-        [(numpy.array, numpy.array)] -- Tuple of features and labels,
-            where features are the quantitative data from the given
-            dataframe and labels are the one-hot encoded batches for
-            each instance.
-    """
-    if meta_cols is None:
-        meta_cols = ["Sample", batch_col]
-    features = np.array(df.drop(meta_cols, axis=1))
-    labels = pd.get_dummies(df[batch_col])
-    labels = np.array(labels, dtype=float)
-    if sample is not None:
-        rows = pd.DataFrame(df.index).sample(sample, replace=True)[0].values.tolist()
-        features = features[rows]
-        labels = labels[rows]
-    return features, labels
-
-def list_categorical_columns(df):
-    """Get the names of all categorical columns in the dataframe.
-
-    Arguments:
-        df {pandas.DataFrame} -- The dataframe.
-
-    Returns:
[…]

### 2019-07-15 code/other/network.py

@@ -1,219 +1,0 @@
-"""Definitions for the neural networks in Confounded.
-"""
-
-# pylint: disable=E1129
-
-import functools
-import tensorflow as tf
-from tensorflow.contrib.layers import fully_connected, batch_norm # pylint: disable=E0611
-from math import ceil
-
-LEARNING_RATE = 0.0001
-
-def is_square(n):
-    sqrt = n**0.5
-    return int(sqrt) == sqrt
-
-def var_scope(scope):
-    """Decorator to wrap a function in a tensorflow variable scope
-
-    Arguments:
-        scope {str} -- Name of the variable scope
-
-    Returns:
-        function -- The decorated function wrapped in the variable scope
-    """
-    def decorator_var_scope(func):
-        @functools.wraps(func)
-        def wrapper_var_scope(*args, **kwargs):
-            with tf.variable_scope(scope):
-                return func(*args, **kwargs)
-        return wrapper_var_scope
-    return decorator_var_scope
-
-def show_before_and_after_images(func):
-    @functools.wraps(func)
-    def wrapper_show_images(*args, **kwargs):
-        inputs = args[0]
-        show_image(inputs, name="inputs")
-        outputs = func(*args, **kwargs)
-        if isinstance(outputs, tuple):
-            # The autoencoder functions might return (outputs, loss)
-            show_image(outputs[0], name="outputs")
-        else:
-            show_image(outputs, name="outputs")
-        return outputs
-    return wrapper_show_images
-
-@var_scope("vae")
-@show_before_and_after_images
-def variational_autoencoder(inputs, code_size=20):
-    """Creates a variational autoencoder based on "Hands-On Machine
-    Learning with Scikit-Learn and TensorFlow by Aurélien Géron
-    (O’Reilly). Copyright 2017 Aurélien Géron, 978-1-491-96229-9."
-
-    Arguments:
-        input_size {int} -- Size of the input to the autoencoder
-
-    Returns:
-        input {Tensor} -- The input tensor
-        output {Tensor} -- The output tensor
-        loss {Tensor} -- The loss operation
-    """
-    layer_sizes = [500, 500]
-    activations = [tf.nn.elu for _ in layer_sizes]
-    input_size = layer_size(inputs)
-
-    encoding = make_layers(inputs, layer_sizes, activations)
-    code_mean, code_gamma, code = vae_code_layer(encoding, code_size)
[…]

### 2019-07-15 code/other/reformat.py

@@ -1,58 +1,0 @@
-import pandas as pd
-
-def to_csv(df, path, tidy=False, meta_cols=None):
-    """Save a dataframe as a CSV, and optionally add meta columns and
-        transpose the dataframe.
-
-    Arguments:
-        df {pandas.DataFrame} -- The data to be saved.
-        path {str} -- The path where the csv should be saved.
-
-    Keyword Arguments:
-        tidy {bool} -- True if the rows are instances and the columns
-            are features; False otherwise. (default: {False})
-        meta_cols {dict} -- Dictionary of {column_name: [values, ...]}
-            (default: {None})
-    """
-    if not tidy:
-        df = df.T
-    df = _add_meta_cols(df, meta_cols)
-    df.to_csv(path, index=False)
-
-def _add_meta_cols(df, meta_cols):
-    if meta_cols is None:
-        meta_cols = {}
-    sample = meta_cols.get("Sample")
-    df['Sample'] = _check_sample(df, sample)
-    for col_name, col in meta_cols.items():
-        if col_name in ["Sample"]:
-            continue
-        df[col_name] = col
-    return _reorder_cols(df, list(meta_cols.keys()))
-
-def _check_sample(df, sample):
-    if sample is None:
-        return df.index
-    return sample
-
-def _reorder_cols(df, meta_cols):
-    cols = df.columns.tolist()
-    for col in meta_cols:
-        cols.remove(col)
-    cols = meta_cols + cols
-    return df[cols]
-
-if __name__ == "__main__":
-    import functools
-    import operator
-    in_path = "data/mnist_matrix.csv"
-    out_path = "data/tidy_mnist2.csv"
-    df = pd.read_csv(in_path, header=None)
-    to_csv(df, out_path, tidy=False, meta_cols={
-        "Batch": None,
-        "Sample": None,
-        "Digit": functools.reduce(
-            operator.add,
-            [[x] * 1000 for x in range(10)]
-        )
-    })

### 2019-07-15 code/adjust/all.sh
Add status echo statements and quit on interrupt

@@ -1,5 +1,7 @@
 #!/bin/bash
 
+set -e
+
 bash ./scale.sh
 bash ./combat.sh
 bash ./confounded.sh

### 2019-07-15 code/adjust/combat.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mAdjusting the data with ComBat\033[0m\n"
+
 Rscript combat.R "../data/input/mnist/noisy.csv" -b Batch
 Rscript combat.R "../data/input/bladderbatch/bladderbatch.csv" -b batch
 Rscript combat.R "../data/input/gse37199/gse37199.csv" -b plate

### 2019-07-15 code/adjust/confounded.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mAdjusting the data with Confounded\033[0m\n"
+
 docker run -it \
     -v $(pwd)/../data/input/bladderbatch/:/confounded/data/bladderbatch \
     jdayton3/confounded data/bladderbatch/bladderbatch.csv -b batch

### 2019-07-15 code/adjust/scale.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mAdjusting the data with a linear scaler\033[0m\n"
+
 Rscript combat.R "../data/input/mnist/noisy.csv" -a scale -b Batch
 Rscript combat.R "../data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
 Rscript combat.R "../data/input/gse37199/gse37199.csv" -a scale -b plate

### 2019-07-15 code/metrics/all.sh

@@ -1,5 +1,7 @@
 #!/bin/bash
 
+set -e
+
 export data_dir="../data/metrics/"
 
 bash ./mse.sh

### 2019-07-15 code/metrics/classify.sh

@@ -1,3 +1,7 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mCalculating classification accuracy\033[0m\n"
+
 python classify.py

### 2019-07-15 code/metrics/mmd.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mCalculating MMD\033[0m\n"
+
 out_dir="${data_dir}:-../data/metrics"
 out_path="${out_dir}/mmd.csv"
 in_dirs=$(ls -d ../data/input/*)

### 2019-07-15 code/metrics/mse.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mCalculating MSE\033[0m\n"
+
 out_dir="${data_dir}:-../data/metrics"
 out_path="${out_dir}/mse.csv"
 in_dirs=$(ls -d ../data/input/*)

### 2019-07-15 code/prepdata/all.sh

@@ -1,5 +1,7 @@
 #!/bin/bash
 
+set -e
+
 export data_dir="../data/input/"
 bash mnist.sh
 bash bladderbatch.sh

### 2019-07-15 code/prepdata/bladderbatch.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mDownloading and tidying the bladderbatch dataset\033[0m\n"
+
 data_dir=${data_dir:-"../data/"}
 dest="${data_dir}/bladderbatch/"
 out_csv="${dest}/bladderbatch.csv"

### 2019-07-15 code/prepdata/gse37199.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mDownloading the GSE37199 dataset\033[0m\n"
+
 data_dir=${data_dir:-"../data/"}
 dest="${data_dir}/gse37199/"
 raw="${dest}/raw/"
@@ -14,5 +18,7 @@ if [ ! -f $raw_loc ]; then
     wget https://osf.io/av3yt/download -O $raw_loc
 fi
 tar -zxvf $raw_loc -C $raw
+
+printf "\033[0;32mTidying the GSE37199 dataset\033[0m\n"
 
 python gse37199.py $expression_tsv $clinical_tsv $out_csv

### 2019-07-15 code/prepdata/mnist.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mDownloading the MNIST dataset\033[0m\n"
+
 data_dir=${data_dir:-"../data/"}
 dest="${data_dir}/mnist/"
 raw_csv="${dest}/raw.csv"
@@ -11,6 +15,8 @@ mkdir -p $dest
 if [ ! -f $raw_csv ]; then
     wget https://pjreddie.com/media/files/mnist_test.csv -O $raw_csv
 fi
+
+printf "\033[0;32mTidying the MNIST dataset\033[0m\n"
 
 python mnist.py $raw_csv $tidy_csv
 python artificial_batch.py $tidy_csv $batched_csv

### 2019-07-15 code/prepdata/tcga.sh

@@ -1,5 +1,9 @@
 #!/bin/bash
 
+set -e
+
+printf "\033[0;32mDownloading the TCGA dataset\033[0m\n"
+
 data_dir=${data_dir:-"../data/"}
 dest="${data_dir}/tcga/"
 rnaseq="${dest}/rnaseq.tsv.gz"
@@ -18,5 +22,7 @@ fi
 if [ ! -f $labels ]; then
     wget https://osf.io/frxv6/download -O $labels
 fi
+
+printf "\033[0;32mTidying the TCGA dataset\033[0m\n"
 
 python tcga.py $rnaseq $mutations $labels $final

### 2019-07-15 code/run_all.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+
+set -e
+
+cd prepdata
+bash all.sh
+cd ../
+
+cd adjust
+bash all.sh
+cd ../
+
+cd metrics
+bash all.sh
+cd ../

### 2019-07-16 code/adjust/confounded.sh
Start making figure scripts

@@ -12,7 +12,9 @@ docker run -it \
     jdayton3/confounded data/gse37199/gse37199.csv -b plate
 docker run -it \
     -v $(pwd)/../data/input/mnist/:/confounded/data/mnist \
-    jdayton3/confounded data/mnist/noisy.csv -b Batch
+    -v $(pwd)/../data/metrics/:/confounded/data/log \
+    jdayton3/confounded data/mnist/noisy.csv -b Batch \
+    -f data/log/log.csv # save a log for our loss chart
 docker run -it \
     -v $(pwd)/../data/input/tcga/:/confounded/data/tcga \
     jdayton3/confounded data/tcga/tcga.csv -b CancerType

### 2019-07-16 code/figures/loss.R

@@ -1,0 +1,23 @@
+# Load packages -----
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+p_load("tidyverse", "gridExtra", "png", "stringr", "magick", "colorspace", "pracma")
+
+# Load data ---------
+df <- read_csv("../data/metrics/log.csv") %>% 
+  select(ae_loss, disc_loss, iteration) %>% 
+  gather("Function", "Loss Function", -iteration) %>% 
+  mutate(Function = str_replace(Function, "_loss", "")) %>%
+  mutate(Function = str_replace(Function, "ae", "Autoencoder")) %>%
+  mutate(Function = str_replace(Function, "disc", "Discriminator")) %>%
+  rename(Iteration = iteration)
+
+# Make plots --------
+df %>% 
+  filter(Iteration > 9000) %>% 
+  ggplot(aes(x = Iteration, y = `Loss Function`)) + 
+    geom_line() + 
+    geom_smooth() + 
+    geom_vline(xintercept = 9600, color = "Red", linetype = "longdash") +
+    facet_wrap(vars(Function), nrow=2, scales = "free_y") +
+    theme_bw(base_size = 18)
+ggsave("../data/output/loss_chart.pdf")

### 2019-07-16 code/figures/mse_mmd_classification.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+
+set -e
+
+printf "\033[0;32mGenerating MSE, MMD, and classification figures and tables\033[0m\n"

### 2019-07-16 code/figures/pca_and_tsne.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+
+set -e
+
+printf "\033[0;32mGenerating PCA and t-SNE figures\033[0m\n"
+
+Rscript pca_and_tsne.R

### 2019-07-16 code/prepdata/mnist.py

@@ -13,5 +13,4 @@ df = df.set_index("Digit")\
     .reset_index()\
     .reset_index()\
     .rename(columns={"index": "Sample"})
-df["Batch"] = ["A", "B"] * 5000
 df.to_csv(args.outfile, index=False)

### 2019-07-16 code/figures/mnist.R
Make MSE, MMD, classification, and MNIST plots

@@ -1,0 +1,61 @@
+# Load packages -------
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+p_load("tidyverse", "gridExtra", "png", "stringr", "magick", "colorspace", "pracma")
+
+## MNIST Grid ----------
+
+mnist_dir <- "../data/input/mnist/"
+OUT_DIR <- "../data/output/"
+
+dfs <- list(
+  Unadjusted = read_csv(paste(c(mnist_dir, "unadjusted.csv"), collapse="")) %>% select(-Sample, -Digit),
+  `Artificial noise` = read_csv(paste(c(mnist_dir, "noisy.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
+  Scale = read_csv(paste(c(mnist_dir, "noisy_scale.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
+  ComBat = read_csv(paste(c(mnist_dir, "noisy_combat.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
+  Confounded = read_csv(paste(c(mnist_dir, "noisy_confounded.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit)
+)
+
+
+pic_rows <- c(1, 2223, 4445, 1112, 3334, 5556)
+
+image_from_vector <- function(x) {
+  return(image_read(aperm(array(as.numeric(x), c(28, 28, 1)), c(2, 1, 3))))
+}
+
+big_img <- NULL
+imgs <- list()
+for (name in names(dfs)) {
+  img <- NULL
+  df <- dfs[[name]]
+  for (i in pic_rows) {
+    new_img <- image_from_vector(df[i,])
+    if (is.null(img)) {
+      img <- new_img
+    } else {
+      img <- image_append(c(img, new_img))
+    }
+  }
+  imgs[[i]] <- img
+  if (is.null(big_img)) {
+    big_img <- img
+  } else {
+    big_img <- image_append(c(big_img, img), stack = TRUE)
+  }
+}
+
+label_width <- 65
+text_background <- image_scale(image_read(array(rep(1, 28*label_width), c(28, label_width, 1))), "x100")
+labels <- NULL
+for (name in names(dfs)) {
+  new_label <- image_annotate(text_background, name, gravity = "center", size = 32)
+  if (is.null(labels)) {
+    labels <- new_label
+  } else {
[…]

### 2019-07-16 code/figures/mse_mmd_classification.sh

@@ -3,3 +3,5 @@
 set -e
 
 printf "\033[0;32mGenerating MSE, MMD, and classification figures and tables\033[0m\n"
+
+Rscript mse_mmd_classification.R

### 2019-07-16 code/run_all.sh

@@ -13,3 +13,7 @@ cd ../
 cd metrics
 bash all.sh
 cd ../
+
+cd figures
+bash all.sh
+cd ../

### 2019-07-16 code/figures/code/mnist.Rmd

---
title: "MNIST"
author: "Jonathan Dayton"
date: "7/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman"); library(pacman)
p_load("tidyverse", "gridExtra", "png", "stringr", "magick", "colorspace", "pracma")
```

## MNIST Grid

```{r}
mnist_dir <- "~/Documents/research/confounded/data/mnist/"

dfs <- list(
  Unadjusted = read_csv(paste(c(mnist_dir, "unadjusted.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
  `Artificial noise` = read_csv(paste(c(mnist_dir, "noisy.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
  Scale = read_csv(paste(c(mnist_dir, "noisy_scale.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
  ComBat = read_csv(paste(c(mnist_dir, "noisy_combat.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit),
  Confounded = read_csv(paste(c(mnist_dir, "noisy_confounded.csv"), collapse="")) %>% select(-Batch, -Sample, -Digit)
)
```

```{r}
unadj <- dfs$Unadjusted
set.seed(8)
pic_rows <- sort(sample.int(nrow(unadj), 5))
print(pic_rows)
pic_rows <- c(1, 2223, 4445, 1112, 3334, 5556)

image_from_vector <- function(x) {
  return(image_read(aperm(array(as.numeric(x), c(28, 28, 1)), c(2, 1, 3))))
}

big_img <- NULL
imgs <- list()
for (name in names(dfs)) {
  img <- NULL
  df <- dfs[[name]]
  for (i in pic_rows) {
    new_img <- image_from_vector(df[i,])
    if (is.null(img)) {
      img <- new_img
    } else {
      img <- image_append(c(img, new_img))
    }
  }
  imgs[[i]] <- img
  if (is.null(big_img)) {
    big_img <- img
  } else {
    big_img <- image_append(c(big_img, img), stack = TRUE)
  }
}

label_width <- 65
text_background <- image_scale(image_read(array(rep(1, 28*label_width), c(28, label_width, 1))), "x100")
labels <- NULL
for (name in names(dfs)) {
  new_label <- image_annotate(text_background, name, gravity = "center", size = 32)
  if (is.null(labels)) {
    labels <- new_label
  } else {
    labels <- image_append(c(labels, new_label), stack = TRUE)
  }
}

big_img <- image_scale(big_img, "500")
labelled <- image_append(c(labels, big_img))

image_write(labelled, "../../../../research/thesis/figures/final/mnist.png")
```

### 2019-07-16 thesis.tex
Address comments in the manuscript

@@
 1) when the input data contain only two batches,
-2) the batches are sufficiently large (> 100 samples in our testing), and
-3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording.@]
+2) the batches are sufficiently large (batch sizes smaller than $n=100$ failed in our testing), and
+3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording. -Is this more clear?@]
 These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
 Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
-To address these limitations, we implemented a deep neural network that [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.@]
+To address these limitations, we implemented a deep neural network that 1) has a batch output layer that expands to fit
+the number of batches, 2) works on smaller batch sizes by sampling with replacement from the training data, and 3) has
+no formulas that require batches to be equal sizes.
+[@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.
+Is this clear enough?
+@]
 Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.
 
-Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here.@].
-This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.@]
+Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here.
+- This reference looks great to me.@].
+This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.
+- Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise?
+Either way, we might need a reference.@]
 Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
-However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.@]
+However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.
+- The wording and both references look great to me.@]
 Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
@@
 We trained the discriminator to predict the original batch of the autoencoder's output.
-The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches[@Please provide a little more detail about how the output layer size is calculated@].
+The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches (i.e.
+if there are 4 distinct batches in the data, the output layer will be size 4).
+[@Please provide a little more detail about how the output layer size is calculated - How's this?@].
 To combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
@@
 Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.[@This won't be convincing enough to the reviewers.
-Be more explicit about why we couldn't use these other methods.@]
+Be more explicit about why we couldn't use these other methods.
+- Should I just copy the info in the background here (the "limited range of scenarios" part)?@]
 
@@
 We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
-[@Say something about what hyperparameters we used. Defaults?@]
+In most cases, we used the default hyperparameters, with exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
+[@Say something about what hyperparameters we used. Defaults? - How's this?@]
 
@@
 Let me know if you disagree.
-If not, please remove it.@]
+If not, please remove it.
+- We talked about maybe keeping it after training for 10k at LR=0.001 and 10k at LR=0.0001.
+What are your thoughts on this now?@]
 
-[@For simplicity, I think we should only include t-SNE graphs, not PCA. t-SNE accounts for nonlinear patterns, which is the point of our method.@]
-[@I also think we should have t-SNE graphs for all three datasets.]
-[@Please put in the full name for t-SNE and cite the paper for this method.@]
-We used principal component analysis (PCA) and t-Stochastic... (t-SNE) plots to visualize the data before and after batch adjustment.
+[@For simplicity, I think we should only include t-SNE graphs, not PCA.
+t-SNE accounts for nonlinear patterns, which is the point of our method.
+- I like including both because it's an evidence that linear visualizations don't cut it when using ML algorithms, but I could go either way on this.@]
+[@I also think we should have t-SNE graphs for all three datasets. - Do you think this should be in the main text or just in the supplement?@]
+We used principal component analysis (PCA) and t-distributed Stochastic Neighbor Embedding (t-SNE) \citep{maaten_visualizing_2008} plots to visualize the data before and after batch adjustment.
 Before any adjustment, batches (microarray plates) in the GSE37199 dataset showed distinct expression patterns for each batch.

### 2019-07-16 code/all.sh
Started modifying scripts to be executed in Docker.

@@ -4,6 +4,7 @@ set -e
 
 cd prepdata
 bash all.sh
+exit
 cd ../
 
 cd adjust

### 2019-07-16 code/prepdata/bladderbatch.R

@@ -1,5 +1,6 @@
 if (!require("pacman")) install.packages("pacman"); library(pacman)
-p_load("tidyverse", "argparse")
+#p_load("tidyverse", "argparse")
+p_load("dplyr", "readr", "tibble", "argparse")
 
 parser <- ArgumentParser()
 parser$add_argument("outfile", help = "Path to the output file")
@@ -13,11 +14,12 @@ data(bladderdata)
 
 clin_df <- bladderEset %>%
   pData() %>%
-  as.tibble(rownames = "id")
+  as_tibble(rownames = "id")
+
 gene_df <- bladderEset %>%
   exprs() %>%
   t() %>%
-  as.tibble(rownames = "id")
+  as_tibble(rownames = "id")
 
 all <- inner_join(clin_df, gene_df, by = "id")
 

### 2019-07-16 code/prepdata/gse37199.sh

@@ -4,19 +4,14 @@ set -e
 
 printf "\033[0;32mDownloading the GSE37199 dataset\033[0m\n"
 
-data_dir=${data_dir:-"../data/"}
-dest="${data_dir}/gse37199/"
-raw="${dest}/raw/"
+dest="${data_dir}/gse37199"
+raw="/tmp"
 raw_loc="${raw}/raw.tar.gz"
 expression_tsv="${raw}/GSE37199_Expression_BatchUnadjusted.txt.gz"
 clinical_tsv="${raw}/GSE37199_Clinical.txt"
 out_csv="${dest}/gse37199.csv"
 
-mkdir -p $raw
-
-if [ ! -f $raw_loc ]; then
-    wget https://osf.io/av3yt/download -O $raw_loc
-fi
+wget https://osf.io/av3yt/download -O $raw_loc
 tar -zxvf $raw_loc -C $raw
 
 printf "\033[0;32mTidying the GSE37199 dataset\033[0m\n"

### 2019-07-16 code/prepdata/mnist.sh

@@ -4,17 +4,12 @@ set -e
 
 printf "\033[0;32mDownloading the MNIST dataset\033[0m\n"
 
-data_dir=${data_dir:-"../data/"}
-dest="${data_dir}/mnist/"
-raw_csv="${dest}/raw.csv"
+dest="${data_dir}/mnist"
+raw_csv=/tmp/raw.csv
 tidy_csv="${dest}/unadjusted.csv"
 batched_csv="${dest}/noisy.csv"
 
-mkdir -p $dest
-
-if [ ! -f $raw_csv ]; then
-    wget https://pjreddie.com/media/files/mnist_test.csv -O $raw_csv
-fi
+wget https://pjreddie.com/media/files/mnist_test.csv -O $raw_csv
 
 printf "\033[0;32mTidying the MNIST dataset\033[0m\n"
 

### 2019-07-16 code/run_docker.sh

@@ -1,0 +1,11 @@
+#! /bin/bash
+
+image=srp33/confounded-paper:version1
+
+docker build -t $image .
+
+mkdir -p data/input/mnist data/input/bladderbatch data/input/gse37199 data/input/tcga
+
+docker run -i -t --rm \
+  -v $(pwd)/data:/data \
+  $image

### 2019-07-16 code/prepdata/tcga.py
Make to_sparse_matrix more memory-efficient

@@ -25,6 +25,8 @@ mutations = pd.read_csv(args.mutations, sep="\t")
 ctypes = pd.read_csv(args.cancer_types, sep="\t")
 expression = pd.read_csv(args.rnaseq, sep="\t")
 
+most_mutated = mutations['MutatedGene'].value_counts()[:5].index.tolist()
+mutations = mutations[mutations['MutatedGene'].isin(most_mutated)]
 mutations_matrix = to_sparse_matrix(mutations)
 clinical = ctypes.merge(mutations_matrix, on="Sample", how="left").fillna(0)
 clinical = clinical.applymap(maybe_to_int)

### 2019-07-16 code/all.sh
Updated Docker so that dependencies are installed when image is built.

@@ -4,12 +4,13 @@ set -e
 
 cd prepdata
 bash all.sh
-exit
 cd ../
+exit
 
 cd adjust
 bash all.sh
 cd ../
+exit
 
 cd metrics
 bash all.sh

### 2019-07-16 code/install.R

@@ -1,0 +1,6 @@
+install.packages(c("pacman", "dplyr", "readr", "stringr", "tibble", "argparse", "docstring"))
+
+if (!requireNamespace("BiocManager", quietly = TRUE))
+    install.packages("BiocManager")
+BiocManager::install("bladderbatch")
+BiocManager::install("sva")

### 2019-07-16 code/run_docker.sh

@@ -1,5 +1,7 @@
 #! /bin/bash
 
+set -e
+
 image=srp33/confounded-paper:version1
 
 docker build -t $image .

### 2019-07-16 thesis.tex
Revised abstract

@@
 
-\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
+\title{Adversarial deep neural networks remove nonlinear batch effects from gene-expression data}
 \author{Jonathan Bryan Dayton}
@@
 \affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
+%[@Please list my contact information. I don't remember how to do this in LaTeX. But please make sure it looks similar to our previous paper.]
 
@@
 \begin{abstract}
-	Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
-	However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
-	For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
-	known covariates that are not of interest to the study may have strong effects,
-	and there may be large systemic effects when integrating multiple expression datasets.
-	Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
-	We created Confounded to remove these effects from expression data.
+	Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insights into functional mechanisms of diseases and other biological processes.
+	However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analyses.
+	For example, collecting data in multiple runs can cause nontrivial differences in the data known as batch effects.
+	Covariates that are not of interest to the study may have strong effects on gene expression,
+	and there may be systemic platform effects when integrating multiple datasets.
+	These confounding effects may be driven by higher-order interactions that are not removable using existing techniques that identify linear patterns.
+	We created *Confounded* to remove these effects from expression data.
+	[@Please modify the following sentence to better reflect the dual-objective aspect and the fact that it uses neural networks.@]
 	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
-	We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
-	We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
-	Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
+	We tested the model on artificially constructed data and three gene-expression datasets and compared against other batch-adjustment algorithms.
+	Our software is available at \url{https://github.com/jdayton3/Confounded}.
 \end{abstract}

### 2019-07-16 thesis.tex
Updated abstract and part of the Intro

@@
 	For example, collecting data in multiple runs can cause nontrivial differences in the data known as batch effects.
-	Covariates that are not of interest to the study may have strong effects on gene expression,
-	and there may be systemic platform effects when integrating multiple datasets.
+	In addition, covariates that are not of interest to the study may have strong effects on gene expression.
+%	and there may be systemic effects when integrating multiple datasets.
 	These confounding effects may be driven by higher-order interactions that are not removable using existing techniques that identify linear patterns.
@@
 
-Gene expression data can be applied in many ways to advance our understanding of medicine and biology.
-For example, expression data has been applied to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict the clinical outcome in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
-Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
-Because of the sensitive nature of these data (i.e.
-gene product levels in organisms may respond drastically to small environmental changes) and of the equipment used to quantify expression levels, expression data often contains unwanted confounding effects that can skew analysis.
-Three examples of this include
-\begin{enumerate*}[(a)]
-	\item batch effects, \label{it:batch}
-	\item known covariates, and \label{it:cov}
-	\item dataset-level effects. \label{it:int}
-\end{enumerate*}
-\ref{it:batch} Batch effects occur when expression data are generated in multiple runs or multiple batches, and slight systemic differences occur during the different runs, such as different technicians operating the machine or slight temperature differences in the room.
-Batch effects are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
-In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
-however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
-\ref{it:cov} Systemic bias can be even more pronounced when there are known covariates---for example, using data collected from different cancer types when cancer type is not of interest to the study \citep{dayton_classifying_2017-1}.
-\ref{it:int} Finally, though batch effects within a dataset collected all by one lab at one time may be somewhat decreased experimentally by careful replication of experimental conditions, this is no longer possible when integrating different datasets;
-however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
-Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see \figurename{} \ref{fig:workflow}).
+Gene-expression data can be used to advance our understanding of medicine and biology.
+For example, gene-expression data have helped to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict clinical outcomes in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
+Expression datasets are quite ``wide,'' typically containing tens of thousands of columns representing genes in the human transcriptome.
+Because of the sensitive nature of technologies used to generate these data (i.e., expression levels may respond drastically to small environmental changes), expression data often contains unwanted confounding effects that can skew analyses.
+Three examples of these include batch effects \label{it:batch}, known covariates, dataset-integration effects.
+
+Batch effects occur when expression data are generated in multiple runs or batches and systemic biases affect the runs.
+Such biases include different technicians operating the equipment or slight temperature differences in the room.
+Batch effects often have a measurable impact on high-throughput expression data \citep{leek_tackling_2010}.
+For example, in one study, researchers found that, contrary to prior knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
+however, a later rebuttal showed that when accounting for batch effects, the biological samples clustered more closely by tissue type than by species, as initially expected \citep{gilad_reanalysis_2015}.
+
+Systemic biases can be even more pronounced when driven by known covariates.
+For example, Dayton and Piccolo performed a pan-cancer analysis of gene-expression values in The Cancer Genome Atlas, identifying patterns associated with cancer-mutation status\citep{dayton_classifying_2017-1}.
+We found that a tumor's tissue of origin exerted a strong effect on gene-expression levels.
+This signal was so strong that we could predict a given tumor's tissue of origin with near-perfect accuracy using a machine-learning classifier.
+However, we sought to identify patterns that were tissue-agnostic, so tissue of origin represented a confounding effect.
+
+%TODO: Move this to Discussion as a future issue to tackle?
+%Finally, data integration is a key goal for gene-expression analyses \citep{lazar_batch_2013}.
+%As sample sizes increase, so can statistical significance.
+%Although batch effects within a single dataset may be somewhat decreased by careful replication of experimental conditions, this is not possible when integrating different datasets.
+Methods used for batch correction may be used to correct for dataset-integration effects; however, this problem is
+%however, , .
+
+These biases represent cases where data measurements are affected by hidden variables and must be removed for effective analyses (see \figurename{} \ref{fig:workflow}).
 
@@
 
-Several methods exist for characterizing and removing batch effects from gene-expression data.
-Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
-ComBat uses an empirical Bayes approach to estimate batch effect parameters and then uses linear regression to model and remove the effects.
-SVA uses singular value decomposition to model batch effects, which can then be accounted for in statistical analyses.
-Both of these methods use linear methods to model confounding effects and thus may not account for nonlinear effects, such as multi-gene, cascading interactions within signaling pathways.
-As machine learning becomes more common in biological research, these nonlinear confounding effects become more troublesome because many machine-learning algorithms can successfully account for complex interactions among variables.
+Several methods exist for characterizing or removing batch effects from gene-expression data.
+SVA uses singular value decomposition to model batch effects, which can then be accounted for in statistical analyses\citep{leek_capturing_2007}.
+ComBat uses an empirical Bayes approach to estimate batch effect parameters and then uses linear regression to model and remove the effects\citep{johnson_adjusting_2007}.
+Both of these techniques use linear methods to model confounding effects and thus may not account for nonlinear effects, such as multi-gene, cascading interactions within signaling pathways.
+As machine learning becomes more common in biological research, these nonlinear confounding effects become more troublesome because many machine-learning algorithms can successfully identify such patterns, which may bias the algorithms' outputs.
 %For example, advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
@@
 
-Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
+Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions; the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
 Neural networks have been applied broadly to gene-expression data.

### 2019-07-16 manuscript.tex
Renamed file

--- 2019-07-16 thesis.tex
@@
 1) when the input data contain only two batches,
-2) the batches are sufficiently large (batch sizes smaller than $n=100$ failed in our testing), and
-3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording. -Is this more clear?@]
-These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
-Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
-To address these limitations, we implemented a deep neural network that 1) has a batch output layer that expands to fit
-the number of batches, 2) works on smaller batch sizes by sampling with replacement from the training data, and 3) has
+2) the batches are sufficiently large (batch sizes smaller than 100 failed in our testing), and
+3) the batches are balanced.
+These requirements rarely hold in applied research; for example, the \textit{bladderbatch} dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
+Additionally, papers that have described these methods have not evaluated whether nonlinear patterns remain in expression data after batch correction.
+To address these limitations, we implemented a deep neural-network architecture that 1) has a batch output layer that expands to fit
+the number of batches, 2) works on relatively small batch sizes by sampling with replacement from the training data, and 3) has
 no formulas that require batches to be equal sizes.
-[@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.
-Is this clear enough?
-@]
-Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.
+Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remain after adjustment.
 
-Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here.
-- This reference looks great to me.@].
+Autoencoders are a type of neural network that encodes input data in a smaller number of dimensions than the original data and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}.
 This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.
 - Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise?
-Either way, we might need a reference.@]
+Either way, we might need a reference.
+You would know better than I.
+Please clarify this based on what is relevant to the paper and add a relevant reference.@]
 Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
-However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.
-- The wording and both references look great to me.@]
-Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
-Louizos, et al.
-have used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.
+However, neural networks are adept at finding subtle patterns in data;
+thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.
+Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions:
+1) to maximize similarity between in the input and output data and
+2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
+For example, Louizos, et al.
+used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.
 

### 2019-07-17 manuscript.tex
Revised through the Datasets subsection

@@
 An open-source implementation of Confounded is publicly available at \url{https://github.com/jdayton3/Confounded}.
-Materials for generating this manuscript, including scripts and data, are available at \url{https://github.com/jdayton3/ms-thesis/tree/bioinformatics} and \url{https://osf.io/b76ch/}, respectively.
+Materials for generating this manuscript, including scripts and data, are available at \url{https://github.com/jdayton3/confounded-paper} and \url{https://osf.io/b76ch/}, respectively.
+%TODO: We may be able to remove the OSF repo now...but let's wait...
 
@@
 
-We used an adversarial autoencoder network to model and remove confounding effects.
+We developed an adversarial autoencoder network to model and remove confounding effects.
 We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining confounding effects in the autoencoder's output.
@@
 
-We trained the discriminator to predict the original batch of the autoencoder's output.
+We trained the discriminator to predict the original batches based on the autoencoder's output.
 The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches (i.e.
 if there are 4 distinct batches in the data, the output layer will be size 4).
-[@Please provide a little more detail about how the output layer size is calculated - How's this?@].
 To combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
@@
 
-We trained the network using three loss functions.
+To evaluate the network's performance, we used three loss functions.
 First, we calculated the autoencoder's loss ($L_A$) by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler (KL) divergence \cite{kullback_information_1951} of the code layer).
 Second, we calculated the discriminator's loss ($L_D$) as sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
-Finally, we also trained the autoencoder layers on a combination of the two previous losses,
+Finally, we trained the autoencoder layers on a combination of the two previous losses,
 
@@
 We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ central processing unit and 16 GB of random access memory.
-For each dataset, training took roughly 30 minutes to complete, including time to load the input into memory and to save the output to disk.
+For each dataset, training took roughly 30 minutes to complete, including time to load the input into memory and save the output to disk.
 
@@
 Some of these use deep neural networks \citep{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}, while others do not\citep{espin-perez_comparison_2018}.
-Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.[@This won't be convincing enough to the reviewers.
-Be more explicit about why we couldn't use these other methods.
-- Should I just copy the info in the background here (the "limited range of scenarios" part)?@]
+Unfortunately, the datasets we tested did not meet the requirements of these methods (e.g., two equal-sized batches, large sample sizes).
 

### 2019-07-17 manuscript.tex
Finished going through the Methods section.

@@
 
-To test both theoretical and practical differences between Confounded and the alternative methods, we compared them for datasets that varied in size and type of data measured.
+We tested each method on four datasets that varied in size and the ype of data measured.
 
@@
 
-The MNIST digits dataset \cite{lecun_mnist_nodate} contains images of handwritten digits that are size-normalized and centered.
-It contains 60,000 training images and 10,000 test images.
-We used MNIST so we could visually assess how well the batch-correction methods preserve true signal (in this case, the shape and digit of each handwritten digit image) after simulated batch adjustment.
+Initially, we used an image dataset so we could visually assess how well the batch-correction methods preserve true signal (in this case, the shape and definition of each handwritten digit image).
+The MNIST dataset \cite{lecun_mnist_nodate} contains images of handwritten digits that are size-normalized and centered.
+We limited our analysis to the 10,000 test images.
 We flattened each 28-by-28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit (class) information.
-We limited our analysis to the 10,000 test images.
 Although convolutional layers are typically used when working with image data, we used fully connected layers to maintain consistency with what would be used on gene-expression data.
@@
 Because there is no batch information in the MNIST digits dataset, we simulated nonlinear confonding effects.
-We applied nonlinear effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
+We applied these effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.[@Be more specific in the last part.
+How did you apply nonlinearity?@]
 We split the images into two balanced batches (5,000 images for each batch) and included the same number of each digit in each batch.
@@
 The dataset has a cancer status (cancerous vs. normal tissue) variable, which we used as the ``true class'' in our validations; it also has a batch variable.
-Because bladderbatch is small relative to datasets typically used for deep learning, it helped us to evaluate whether our network overfits on gene-expression datasets of the size often used in applied research.
+Because bladderbatch is small relative to datasets typically used for deep learning, it helped us to evaluate whether our network overfits on gene-expression datasets of a size typical of applied research.
 
-\subsubsection{GSE37199}
+\subsubsection{Prostate cancer data}
 
@@
 This dataset has two types of batch variables---``plate'' and ``centre''---which indicate the microarray plate and research lab that processed the data, respectively.
-We adjusted against the ``plate'' variable because it was granular in nature (with counts of $\{43, 50\}$).
+We adjusted against the ``plate'' variable because it is more granular in nature (with counts of $\{43, 50\}$).
 
-\subsubsection{TCGA pan-cancer data}
+\subsubsection{Pan-cancer data}
 
@@
 Mean squared error (MSE) is a measure of how much two vectors or matrices deviate from one another.
-It is commonly used as to quantify loss when the objective (as with autoencoders) is to minimize the difference between input and output values.
+It is commonly used to quantify loss when the objective (as with autoencoders) is to minimize the difference between input and output values.
 
@@
 
-We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
-In most cases, we used the default hyperparameters, with exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
-[@Say something about what hyperparameters we used. Defaults? - How's this?@]
+We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
+In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
+[@Say something about what hyperparameters we used. Defaults? - How's this? The default in version 0.19.1 was 10. Are you saying you used 10 or something different?@]
 
-We calculated the average classification accuracy across four-fold cross-validation repeated three times.
+We calculated the average classification accuracy across four cross-validation folds, repeated three times.
 We interpreted lower accuracy for batch classification to indicate that batch signal was removed more effectively.
-We interpreted higher true-class accuracy to indicate that biologically relevant signal was not lost during the batch-adjustment process.
-Therefore, given output data from a ideal batch adjuster, batch classification accuracy would be no better than random chance, and true-class accuracy would be equal to what would be obtained for the unadjusted data.
+We interpreted higher true-class accuracy to indicate that biologically relevant signal was retained during the batch-adjustment process.
+Therefore, given output data from an ideal batch adjuster, batch classification accuracy would be no better than random chance, and true-class accuracy would be equal to what would be obtained for the unadjusted data.
 

### 2019-07-17 manuscript.tex
Added more comments throughout.

@@
 Either way, we might need a reference.
-You would know better than I.
+-You would know better than I.
 Please clarify this based on what is relevant to the paper and add a relevant reference.@]
@@
 We applied these effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.[@Be more specific in the last part.
-How did you apply nonlinearity?@]
+How did you apply nonlinearity?
+Or maybe this last part of the sentence is unnecessary.@]
 We split the images into two balanced batches (5,000 images for each batch) and included the same number of each digit in each batch.
@@
 In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
-[@Say something about what hyperparameters we used. Defaults? - How's this? The default in version 0.19.1 was 10. Are you saying you used 10 or something different?@]
+[@The defaults in version 0.19.1 of sklearn are n_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]
 
@@
 
-In this study, we created Confounded, an adversarial, variational, autoencoder neural network, to remove nonlinear confounding effects from gene-expression data that may not be accounted for by linear methods.
+In this study, we created Confounded, an adversarial, variational autoencoder, neural network, to remove nonlinear confounding effects from gene-expression data that may not be accounted for by linear methods.
 We compared Confounded to a scaling method and to ComBat \citep{johnson_adjusting_2007} and used qualitative and quantitative techniques to compare the performance of these methods.
 Overall, the scaling method performed consistently worse than the other methods;
-ComBat and Confounded each excelled in different scenarios, which we illustrate below.
+ComBat and Confounded each excelled in different scenarios, which we describe below.
 
@@
 
-[@At least in my mind, we justified using MNIST as a way to visually see how our method performs and to modify it accordingly.
-However, the Results don't reflect that (in my mind) because the MNIST images are blurry for Confounded, whereas they are clear for the other methods.
-There is some explanation below about why this might be happening with Confounded.
-But as an outsider reading this, I would question why we used MNIST at all because our method didn't perform particularly well.
-It's hard to justify using MNIST anyway because it's so different from gene-expression data.
-So I think we should remove it from the paper (but it was fine to use it in your thesis).
-Let me know if you disagree.
-If not, please remove it.
-- We talked about maybe keeping it after training for 10k at LR=0.001 and 10k at LR=0.0001.
-What are your thoughts on this now?@]
+[@I also think we should have t-SNE graphs for all three datasets. - Do you think this should be in the main text or just in the supplement? - The supplement.@]
 
-[@For simplicity, I think we should only include t-SNE graphs, not PCA.
-t-SNE accounts for nonlinear patterns, which is the point of our method.
-- I like including both because it's an evidence that linear visualizations don't cut it when using ML algorithms, but I could go either way on this.@]
-[@I also think we should have t-SNE graphs for all three datasets. - Do you think this should be in the main text or just in the supplement?@]
+[@Please structure this so that the MNIST results are described first.
+I think the bladderbatch results should come next because that is the closest thing we have to a reference set for gene expression.
+Then GSE37199.
+Then pan-cancer.
+Please restructure the results and wording to reflect this order.
+It's OK to mix and match.
+But the first set of results should be in this order.@]
+
 We used principal component analysis (PCA) and t-distributed Stochastic Neighbor Embedding (t-SNE) \citep{maaten_visualizing_2008} plots to visualize the data before and after batch adjustment.

### 2019-07-17 code/adjust/adjust.R
Ran tests for adjustment scripts and renamed some things

@@ -2,24 +2,12 @@
 options(error=function() { traceback(2); if(!interactive()) quit("no", status = 1, runLast = FALSE) })
 
 # Load dependencies --------------------------------
-
-load_stuff <- function() {
-  for (package in c("tidyverse", "docstring", "stringr", "argparse")) {
-    if (!require(package, character.only = TRUE)) { 
-      install.packages(package)
-      library(package, character.only = TRUE)
-    }
-  }
-  install_sva <- function() {
-    ## try http:// if https:// URLs are not supported
-    source("https://bioconductor.org/biocLite.R")
-    biocLite("sva")
-  }
-  if (!require("sva")) install_sva()
-  library(sva)
-}
-
-suppressMessages(load_stuff())
+library(dplyr)
+library(readr)
+library(docstring)
+library(stringr)
+library(argparse)
+library(sva)
 
 # Parse command line args --------------------------
 

### 2019-07-17 code/adjust/all.sh

@@ -2,6 +2,6 @@
 
 set -e
 
-bash ./scale.sh
+#bash ./scale.sh
 #bash ./combat.sh
-#bash ./confounded.sh
+bash ./confounded.sh

### 2019-07-17 code/adjust/combat.sh

@@ -4,7 +4,7 @@ set -e
 
 printf "\033[0;32mAdjusting the data with ComBat\033[0m\n"
 
-Rscript combat.R "../data/input/mnist/noisy.csv" -b Batch
-Rscript combat.R "../data/input/bladderbatch/bladderbatch.csv" -b batch
-Rscript combat.R "../data/input/gse37199/gse37199.csv" -b plate
-Rscript combat.R "../data/input/tcga/tcga.csv" -b CancerType
+Rscript adjust.R "/data/input/mnist/noisy.csv" -b Batch
+Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -b batch
+Rscript adjust.R "/data/input/gse37199/gse37199.csv" -b plate
+Rscript adjust.R "/data/input/tcga/tcga.csv" -b CancerType

### 2019-07-17 code/adjust/confounded.sh

@@ -4,17 +4,8 @@ set -e
 
 printf "\033[0;32mAdjusting the data with Confounded\033[0m\n"
 
-docker run -it \
-    -v $(pwd)/../data/input/bladderbatch/:/confounded/data/bladderbatch \
-    jdayton3/confounded data/bladderbatch/bladderbatch.csv -b batch
-docker run -it \
-    -v $(pwd)/../data/input/gse37199/:/confounded/data/gse37199 \
-    jdayton3/confounded data/gse37199/gse37199.csv -b plate
-docker run -it \
-    -v $(pwd)/../data/input/mnist/:/confounded/data/mnist \
-    -v $(pwd)/../data/metrics/:/confounded/data/log \
-    jdayton3/confounded data/mnist/noisy.csv -b Batch \
-    -f data/log/log.csv # save a log for our loss chart
-docker run -it \
-    -v $(pwd)/../data/input/tcga/:/confounded/data/tcga \
-    jdayton3/confounded data/tcga/tcga.csv -b CancerType
+# We also save a log for our loss chart
+confounded /data/input/mnist/noisy.csv -b Batch -f data/confounded_log.csv
+confounded /data/input/bladderbatch/bladderbatch.csv -b batch
+confounded /data/input/gse37199/gse37199.csv -b plate
+confounded /data/input/tcga/tcga.csv -b CancerType

### 2019-07-17 code/adjust/scale.sh

@@ -4,7 +4,7 @@ set -e
 
 printf "\033[0;32mAdjusting the data with a linear scaler\033[0m\n"
 
-Rscript combat.R "/data/input/mnist/noisy.csv" -a scale -b Batch
-#Rscript combat.R "../data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
-#Rscript combat.R "../data/input/gse37199/gse37199.csv" -a scale -b plate
-#Rscript combat.R "../data/input/tcga/tcga.csv" -a scale -b CancerType
+Rscript adjust.R "/data/input/mnist/noisy.csv" -a scale -b Batch
+Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
+Rscript adjust.R "/data/input/gse37199/gse37199.csv" -a scale -b plate
+Rscript adjust.R "/data/input/tcga/tcga.csv" -a scale -b CancerType

### 2019-07-17 code/all.sh

@@ -2,10 +2,9 @@
 
 set -e
 
-cd prepdata
-bash all.sh
-cd ../
-exit
+#cd prepdata
+#bash all.sh
+#cd ../
 
 cd adjust
 bash all.sh

### 2019-07-17 code/metrics/all.sh

@@ -2,7 +2,7 @@
 
 set -e
 
-export data_dir="../data/metrics/"
+export data_dir="/data/metrics/"
 
 bash ./mse.sh
 bash ./mmd.sh

### 2019-07-17 code/adjust/confounded.sh
Modified comments

@@ -5,7 +5,7 @@ set -e
 printf "\033[0;32mAdjusting the data with Confounded\033[0m\n"
 
 # We also save a log for our loss chart
-confounded /data/input/mnist/noisy.csv -b Batch -f data/confounded_log.csv
+#confounded /data/input/mnist/noisy.csv -b Batch -f data/confounded_log.csv
 confounded /data/input/bladderbatch/bladderbatch.csv -b batch
-confounded /data/input/gse37199/gse37199.csv -b plate
-confounded /data/input/tcga/tcga.csv -b CancerType
+#confounded /data/input/gse37199/gse37199.csv -b plate
+#confounded /data/input/tcga/tcga.csv -b CancerType

### 2019-07-19 code/adjust/confounded.sh
Updated permissions and reworked TCGA data prep.

@@ -5,7 +5,7 @@ set -e
 printf "\033[0;32mAdjusting the data with Confounded\033[0m\n"
 
 # We also save a log for our loss chart
-confounded /data/input/mnist/noisy.csv -b Batch -f data/confounded_log.csv
-confounded /data/input/bladderbatch/bladderbatch.csv -b batch
-confounded /data/input/gse37199/gse37199.csv -b plate
+#confounded /data/input/mnist/noisy.csv -b Batch -f /data/metrics/mnist_confounded_log.csv
+#confounded /data/input/bladderbatch/bladderbatch.csv -b batch
+#confounded /data/input/gse37199/gse37199.csv -b plate
 confounded /data/input/tcga/tcga.csv -b CancerType

### 2019-07-19 code/adjust/scale.sh

@@ -4,7 +4,7 @@ set -e
 
 printf "\033[0;32mAdjusting the data with a linear scaler\033[0m\n"
 
-Rscript adjust.R "/data/input/mnist/noisy.csv" -a scale -b Batch
-Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
-Rscript adjust.R "/data/input/gse37199/gse37199.csv" -a scale -b plate
+#Rscript adjust.R "/data/input/mnist/noisy.csv" -a scale -b Batch
+#Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
+#Rscript adjust.R "/data/input/gse37199/gse37199.csv" -a scale -b plate
 Rscript adjust.R "/data/input/tcga/tcga.csv" -a scale -b CancerType

### 2019-07-19 code/all.sh

@@ -6,9 +6,10 @@ set -e
 #bash all.sh
 #cd ../
 
-#cd adjust
-#bash all.sh
-#cd ../
+cd adjust
+bash all.sh
+cd ../
+exit
 
 #cd metrics
 #bash all.sh

### 2019-07-19 code/figures/loss.R

@@ -3,7 +3,7 @@ if (!require("pacman")) install.packages("pacman"); library(pacman)
 p_load("tidyverse", "gridExtra", "png", "stringr", "magick", "colorspace", "pracma")
 
 # Load data ---------
-df <- read_csv("../data/metrics/log.csv") %>% 
+df <- read_csv("/data/metrics/mnist_confounded_log.csv") %>% 
   select(ae_loss, disc_loss, iteration) %>% 
   gather("Function", "Loss Function", -iteration) %>% 
   mutate(Function = str_replace(Function, "_loss", "")) %>%
@@ -20,4 +20,4 @@ df %>%
     geom_vline(xintercept = 9600, color = "Red", linetype = "longdash") +
     facet_wrap(vars(Function), nrow=2, scales = "free_y") +
     theme_bw(base_size = 18)
-ggsave("../data/output/loss_chart.pdf")
+ggsave("/data/output/loss_chart.pdf")

### 2019-07-19 code/prepdata/tcga.py

@@ -23,6 +23,7 @@ args = parser.parse_args()
 
 mutations = pd.read_csv(args.mutations, sep="\t")
 ctypes = pd.read_csv(args.cancer_types, sep="\t")
+print(args.rnaseq)
 expression = pd.read_csv(args.rnaseq, sep="\t")
 
 most_mutated = mutations['MutatedGene'].value_counts()[:5].index.tolist()

### 2019-07-19 code/prepdata/tcga.sh

@@ -10,7 +10,20 @@ mutations="/tmp/mutations.tsv.gz"
 labels="/tmp/labels.tsv.gz"
 final="${dest}/tcga.csv"
 
-wget https://osf.io/7xjdn/download -O $rnaseq
+# This gets the TCGA RNA-Seq dataset
+wget -O /tmp/GSE62944_RAW.tar "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file"
+cd /tmp
+tar -xvf GSE62944_RAW.tar
+rm GSE62944_RAW.tar
+mv GSM1536837_06_01_15_TCGA_24.tumor_Rsubread_TPM.txt.gz rnaseq.tsv.gz
+python /prepdata/transpose_matrix.py rnaseq.tsv.gz rnaseq2.tsv.gz
+python /prepdata/tcga_expression.py rnaseq2.tsv.gz $rnaseq
+
+### Old version. Only has RNA-Seq data for ~300 genes
+###wget https://osf.io/7xjdn/download -O $rnaseq
+
+cd /prepdata
+
 wget https://osf.io/na3rp/download -O $mutations
 wget https://osf.io/frxv6/download -O $labels
 

### 2019-07-19 code/run_docker.sh

@@ -9,6 +9,10 @@ docker build -t $image .
 mkdir -p data/input/mnist data/input/bladderbatch data/input/gse37199 data/input/tcga
 mkdir -p data/metrics
 
-docker run -i -t --rm \
+#docker run -i -t --rm \
+docker run -i --rm \
   -v $(pwd)/data:/data \
+  -v $(pwd)/../metrics:/data/metrics \
+  -v $(pwd)/../figures:/figures \
   $image
+#  --user $(id -u):$(id -g) \

### 2019-07-19 code/adjust/confounded.sh
Revising TCGA processing

@@ -5,7 +5,9 @@ set -e
 printf "\033[0;32mAdjusting the data with Confounded\033[0m\n"
 
 # We also save a log for our loss chart
-#confounded /data/input/mnist/noisy.csv -b Batch -f /data/metrics/mnist_confounded_log.csv
-#confounded /data/input/bladderbatch/bladderbatch.csv -b batch
+#confounded /data/input/mnist/noisy.csv -b Batch --learning-rate 0.0001 -f /data/metrics/mnist_confounded_log.csv
+##confounded /data/input/mnist/noisy.csv -b Batch --learning-rate 0.001  -f /data/metrics/mnist_confounded_log2.csv
+confounded /data/input/bladderbatch/bladderbatch.csv -b batch
+#/confounded/log
 #confounded /data/input/gse37199/gse37199.csv -b plate
-confounded /data/input/tcga/tcga.csv -b CancerType
+#confounded /data/input/tcga/tcga.csv -b CancerType

### 2019-07-19 code/install.R

@@ -1,4 +1,4 @@
-install.packages(c("pacman", "dplyr", "readr", "stringr", "tibble", "argparse", "docstring"))
+install.packages(c("pacman", "dplyr", "readr", "stringr", "tibble", "ggplot2", "gridExtra", "png", "magick", "colorspace", "pracma", "kableExtra", "Rtsne", "argparse", "docstring"))
 
 if (!requireNamespace("BiocManager", quietly = TRUE))
     install.packages("BiocManager")

### 2019-07-19 code/prepdata/all.sh

@@ -4,7 +4,7 @@ set -e
 
 export data_dir="/data/input"
 
-bash mnist.sh
-bash bladderbatch.sh
-bash gse37199.sh
+#bash mnist.sh
+#bash bladderbatch.sh
+#bash gse37199.sh
 bash tcga.sh

### 2019-07-19 code/prepdata/tcga.R

@@ -1,0 +1,28 @@
+library(readr)
+library(dplyr)
+
+rnaseq_file_path = commandArgs()[7]
+mutation_file_path = commandArgs()[8]
+cancer_types_file_path = commandArgs()[9]
+out_file_path = commandArgs()[10]
+
+rnaseq = read_tsv(rnaseq_file_path)
+mutations = read_tsv(mutation_file_path)
+cancer_types = read_tsv(cancer_types_file_path)
+
+unique_sample_ids = intersect(intersect(pull(rnaseq, Sample), pull(mutations, Sample)), pull(cancer_types, Sample))
+
+rnaseq = filter(rnaseq, Sample %in% unique_sample_ids)
+mutations = filter(mutations, Sample %in% unique_sample_ids)
+cancer_types = filter(cancer_types, Sample %in% unique_sample_ids)
+
+mutated_sample_ids = filter(mutations, MutatedGene == "TP53") %>%
+    pull(Sample)
+
+mutations = tibble(Sample = unique_sample_ids, Mutated = as.integer(unique_sample_ids %in% mutated_sample_ids))
+
+data = inner_join(cancer_types, mutations, by = "Sample") %>%
+    inner_join(rnaseq, by = "Sample") %>%
+    dplyr::rename(TP53_Mutated = Mutated)
+
+write_tsv(data, out_file_path)

### 2019-07-19 code/prepdata/tcga.py

@@ -23,12 +23,12 @@ args = parser.parse_args()
 
 mutations = pd.read_csv(args.mutations, sep="\t")
 ctypes = pd.read_csv(args.cancer_types, sep="\t")
-print(args.rnaseq)
 expression = pd.read_csv(args.rnaseq, sep="\t")
 
 most_mutated = mutations['MutatedGene'].value_counts()[:5].index.tolist()
 mutations = mutations[mutations['MutatedGene'].isin(most_mutated)]
 mutations_matrix = to_sparse_matrix(mutations)
+
 clinical = ctypes.merge(mutations_matrix, on="Sample", how="left").fillna(0)
 clinical = clinical.applymap(maybe_to_int)
 all_data = clinical.merge(expression, on="Sample", how="inner").drop_duplicates()

### 2019-07-19 code/prepdata/tcga.sh

@@ -8,25 +8,27 @@ dest="${data_dir}/tcga"
 rnaseq="/tmp/rnaseq.tsv.gz"
 mutations="/tmp/mutations.tsv.gz"
 labels="/tmp/labels.tsv.gz"
-final="${dest}/tcga.csv"
+#final="${dest}/tcga.csv"
+final="${dest}/tcga2.csv"
 
-# This gets the TCGA RNA-Seq dataset
-wget -O /tmp/GSE62944_RAW.tar "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file"
-cd /tmp
-tar -xvf GSE62944_RAW.tar
-rm GSE62944_RAW.tar
-mv GSM1536837_06_01_15_TCGA_24.tumor_Rsubread_TPM.txt.gz rnaseq.tsv.gz
-python /prepdata/transpose_matrix.py rnaseq.tsv.gz rnaseq2.tsv.gz
-python /prepdata/tcga_expression.py rnaseq2.tsv.gz $rnaseq
+## This gets the TCGA RNA-Seq dataset
+#wget -O /tmp/GSE62944_RAW.tar "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file"
+#cd /tmp
+#tar -xvf GSE62944_RAW.tar
+#rm GSE62944_RAW.tar
+#mv GSM1536837_06_01_15_TCGA_24.tumor_Rsubread_TPM.txt.gz rnaseq.tsv.gz
+#python /prepdata/transpose_matrix.py rnaseq.tsv.gz rnaseq2.tsv.gz
+#python /prepdata/tcga_expression.py rnaseq2.tsv.gz $rnaseq
 
 ### Old version. Only has RNA-Seq data for ~300 genes
 ###wget https://osf.io/7xjdn/download -O $rnaseq
 
 cd /prepdata
 
-wget https://osf.io/na3rp/download -O $mutations
-wget https://osf.io/frxv6/download -O $labels
+#wget https://osf.io/na3rp/download -O $mutations
+#wget https://osf.io/frxv6/download -O $labels
 
 printf "\033[0;32mTidying the TCGA dataset\033[0m\n"
 
-python tcga.py $rnaseq $mutations $labels $final
+##python tcga.py $rnaseq $mutations $labels $final
+Rscript --vanilla tcga.R $rnaseq $mutations $labels $final

### 2019-07-19 code/run_docker.sh

@@ -11,8 +11,9 @@ mkdir -p data/metrics
 
 #docker run -i -t --rm \
 docker run -i --rm \
+  --user $(id -u):$(id -g) \
   -v $(pwd)/data:/data \
   -v $(pwd)/../metrics:/data/metrics \
   -v $(pwd)/../figures:/figures \
+  -v /tmp:/tmp \
   $image
-#  --user $(id -u):$(id -g) \

### 2019-07-19 code/prepdata/tcga.sh
Subsample the tcga dataset

@@ -5,6 +5,8 @@ set -e
 printf "\033[0;32mDownloading the TCGA dataset\033[0m\n"
 
 dest="${data_dir}/tcga"
+medium="${data_dir}/tcga_medium"
+small="${data_dir}/tcga_small"
 rnaseq="/tmp/rnaseq.tsv.gz"
 mutations="/tmp/mutations.tsv.gz"
 labels="/tmp/labels.tsv.gz"
@@ -30,3 +32,8 @@ wget https://osf.io/frxv6/download -O $labels
 printf "\033[0;32mTidying the TCGA dataset\033[0m\n"
 
 python tcga.py $rnaseq $mutations $labels $final
+
+# Do the subsampling stuff
+mkdir -p $medium
+mkdir -p $small
+Rscript tcga_sampling.R

### 2019-07-19 code/prepdata/tcga_sampling.R

@@ -1,0 +1,44 @@
+if (!require("pacman")) install.packages("pacman"); library(pacman)
+p_load(tidyverse)
+
+IN_FILE <- "../data/input/tcga/tcga.csv"
+OUT_DIR <- "../data/input/"
+MEDIUM <- paste0(OUT_DIR, "/tcga_medium/medium.csv")
+SMALL <- paste0(OUT_DIR, "/tcga_small/small.csv")
+
+get_group <- function(df, colname, value) {
+  colname <- enquo(colname)
+  df %>%
+    filter(!! colname == value)
+}
+
+sample_stratified <- function(data, strategy, group_col = "group") {
+  group_col <- sym(group_col)
+  result <- list()
+  groups <- unique(unlist(select(data, !!group_col)))
+  for (g in groups) {
+    chunk <- data %>% filter(!!group_col == g)
+    n <- ceiling(strategy(chunk))
+    result[[g]] <- sample_n(chunk, n)
+  }
+  bind_rows(result)
+}
+
+sample_log <- function(data, group_col = "group") {
+  sample_stratified(data, function(data) {
+    log(nrow(data))
+  }, group_col)
+}
+
+sample_sqrt <- function(data, group_col = "group", scale = 1) {
+  sample_stratified(data, function(data) {
+    sqrt(nrow(data)) * scale
+  }, group_col)
+}
+
+tcga <- read_csv(IN_FILE)
+tcga %>% group_by(CancerType) %>% summarize(count=ceiling(sqrt(n()))) %>% arrange(desc(count)) %>% ungroup()
+
+set.seed(0)
+sample_sqrt(tcga, group_col = "CancerType", scale = 2) %>% write_csv(MEDIUM)
+sample_log(tcga, "CancerType") %>% write_csv(SMALL)

### 2019-07-19 manuscript.tex
Add description of competing methods in comments.
Updated manuscript

@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
+	\includegraphics[width=\columnwidth]{figures/static/adjuster_workflow.pdf}
 	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
+	\includegraphics[width=\columnwidth]{figures/static/network.pdf}
 	\caption[Network architecture of Confounded]{\textbf{Network architecture of Confounded.}
@@
 
+% ----- Brief explanation of the other methods -----
+% 1.
+% The Shaham lab has two method that we reference up here^, one from 2017 , and one from 2018.
+% The 2017 one is the distribution-matching residual net.
+% It arbitrarily designates one batch as the "target" and the other batch as the "source".
+% It calculates MMD with 3 kernels and averages them, and uses them as the loss function.
+% Over training, the source's distribution (according to MMD) should be aligned to the target's distribution.
+% In our testing, this was SUPER memory-intensive.
+% I think this is the practical reason they reduced their dataset to the first 37 principal components, but their explanation in the paper is that the people who generated the data said the first 37 components contain most of the variation.
+% 2.
+% The 2018 Shaham paper is about using an adversarial autoencoder to make a batch-free encoding of expression data.
+% They have a variational autoencoder with a shared encoder, but separate decoders for each of the 2 batches.
+% They say that the decoders output data that has batch info in it but that the encoding layer doesn't have batch info.
+% Their encoding layer is only size 15-20, so this is essentially a dimensionality reduction technique--the "batch free encoding" they're producing doesn't have values for each gene.
+% 3.
+% I don't know how worth-mentioning the Upadhyay paper is, but they basically replicate the DMResNet paper and replace the MMD loss with a discriminator loss like we do.
+% They use the Shaham datasets, and I think their code, which is either not publicly available or is just really hard to find, probably has the same caveats as the Shaham code.
+% Most particularly, they only account for 2 batches.
+% I might have misunderstood their approach though because the English is a little difficult to interpret.
+
 \subsection{Datasets}
@@
 	}
-	\input{figures/final/datasets.tex}
+	\input{tables/static/datasets.tex}
 	\label{tab:datasets}
@@
 which has expression values for 9,365 samples across 25 cancer types.
+%TODO: Change this to 8744 samples (after averaging across replicates and limiting to patients that had all types of data), 23,386 genes and 24 cancer types?
+%TODO: Mention that we used all genes rather than 300?
 
@@
 
+%TODO: Update the algorithms listed here.
 We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
 In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
-[@The defaults in version 0.19.1 of sklearn are n_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]
+[@The defaults in version 0.19.1 of sklearn are n\_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]
 
@@
 \begin{table}
-	\input{figures/final/mse.tex}
+	\input{tables/static/mse.tex}
 	\caption[Mean squared error (MSE)]{
@@
 	\centering
-	\input{figures/final/mmd.tex}
+	\input{tables/static/mmd.tex}
 	\caption[Maximum mean discrepancy (MMD)]{
@@
 	\centering
-	\input{figures/final/batch.tex}
+	\input{tables/static/batch.tex}
 	\caption[Batch classification accuracy]{
@@
 	\centering
-	\input{figures/final/true_class.tex}
+	\input{tables/static/true_class.tex}
 	\caption[True class classification accuracy]{

### 2019-07-19 code/adjust/adjust.R
Updated code for parsing TCGA expression data

@@ -187,7 +187,6 @@ if (!(args$batch_col %in% names(df))) {
 }
 
 message(sprintf("Adjusting using the '%s' adjuster", args$adjuster))
-print(df)
 batch_adjust_tidy(
   df, 
   batch_col = args$batch_col, 

### 2019-07-19 code/adjust/all.sh

@@ -2,6 +2,6 @@
 
 set -e
 
-#bash ./scale.sh
+bash ./scale.sh
 #bash ./combat.sh
-bash ./confounded.sh
+#bash ./confounded.sh

### 2019-07-19 code/all.sh

@@ -2,13 +2,13 @@
 
 set -e
 
-cd prepdata
-bash all.sh
-cd ../
-
-#cd adjust
+#cd prepdata
 #bash all.sh
 #cd ../
+
+cd adjust
+bash all.sh
+cd ../
 
 #cd metrics
 #bash all.sh

### 2019-07-19 code/prepdata/all.sh

@@ -4,7 +4,7 @@ set -e
 
 export data_dir="/data/input"
 
-#bash mnist.sh
-#bash bladderbatch.sh
-#bash gse37199.sh
+bash mnist.sh
+bash bladderbatch.sh
+bash gse37199.sh
 bash tcga.sh

### 2019-07-19 code/prepdata/tcga.R

@@ -25,4 +25,4 @@ data = inner_join(cancer_types, mutations, by = "Sample") %>%
     inner_join(rnaseq, by = "Sample") %>%
     dplyr::rename(TP53_Mutated = Mutated)
 
-write_tsv(data, out_file_path)
+write_csv(data, out_file_path)

### 2019-07-19 code/prepdata/tcga.py

@@ -1,36 +1,0 @@
-import argparse
-import pandas as pd
-
-def maybe_to_int(col):
-    try:
-        return int(col)
-    except:
-        return col
-
-def to_sparse_matrix(df):
-    df["one"] = 1
-    df["MutatedGene"] = df["MutatedGene"].apply(lambda x: str(x) + "_mutation")
-    matrix = df.pivot(index="Sample", columns="MutatedGene", values="one").fillna(0)
-    small_mx = matrix.loc[:, matrix.columns[matrix.sum() > 2000]]
-    return small_mx
-
-parser = argparse.ArgumentParser()
-parser.add_argument("rnaseq", help="Path to RNA Seq data")
-parser.add_argument("mutations", help="Path to somatic mutations data")
-parser.add_argument("cancer_types", help="Path to cancer types data")
-parser.add_argument("outfile", help="Path to outfile")
-args = parser.parse_args()
-
-mutations = pd.read_csv(args.mutations, sep="\t")
-ctypes = pd.read_csv(args.cancer_types, sep="\t")
-expression = pd.read_csv(args.rnaseq, sep="\t")
-
-most_mutated = mutations['MutatedGene'].value_counts()[:5].index.tolist()
-mutations = mutations[mutations['MutatedGene'].isin(most_mutated)]
-mutations_matrix = to_sparse_matrix(mutations)
-
-clinical = ctypes.merge(mutations_matrix, on="Sample", how="left").fillna(0)
-clinical = clinical.applymap(maybe_to_int)
-all_data = clinical.merge(expression, on="Sample", how="inner").drop_duplicates()
-
-all_data.to_csv(args.outfile, index=False)

### 2019-07-19 code/prepdata/tcga.sh

@@ -8,27 +8,25 @@ dest="${data_dir}/tcga"
 rnaseq="/tmp/rnaseq.tsv.gz"
 mutations="/tmp/mutations.tsv.gz"
 labels="/tmp/labels.tsv.gz"
-#final="${dest}/tcga.csv"
-final="${dest}/tcga2.csv"
+final="${dest}/tcga.csv"
 
 ## This gets the TCGA RNA-Seq dataset
-#wget -O /tmp/GSE62944_RAW.tar "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file"
-#cd /tmp
-#tar -xvf GSE62944_RAW.tar
-#rm GSE62944_RAW.tar
-#mv GSM1536837_06_01_15_TCGA_24.tumor_Rsubread_TPM.txt.gz rnaseq.tsv.gz
-#python /prepdata/transpose_matrix.py rnaseq.tsv.gz rnaseq2.tsv.gz
-#python /prepdata/tcga_expression.py rnaseq2.tsv.gz $rnaseq
+wget -O /tmp/GSE62944_RAW.tar "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file"
+cd /tmp
+tar -xvf GSE62944_RAW.tar
+rm GSE62944_RAW.tar
+mv GSM1536837_06_01_15_TCGA_24.tumor_Rsubread_TPM.txt.gz rnaseq.tsv.gz
+python /prepdata/transpose_matrix.py rnaseq.tsv.gz rnaseq2.tsv.gz
+python /prepdata/tcga_expression.py rnaseq2.tsv.gz $rnaseq
 
 ### Old version. Only has RNA-Seq data for ~300 genes
 ###wget https://osf.io/7xjdn/download -O $rnaseq
 
 cd /prepdata
 
-#wget https://osf.io/na3rp/download -O $mutations
-#wget https://osf.io/frxv6/download -O $labels
+wget https://osf.io/na3rp/download -O $mutations
+wget https://osf.io/frxv6/download -O $labels
 
 printf "\033[0;32mTidying the TCGA dataset\033[0m\n"
 
-##python tcga.py $rnaseq $mutations $labels $final
 Rscript --vanilla tcga.R $rnaseq $mutations $labels $final

### 2019-07-19 code/run_docker.sh

@@ -15,5 +15,6 @@ docker run -i --rm \
   -v $(pwd)/data:/data \
   -v $(pwd)/../metrics:/data/metrics \
   -v $(pwd)/../figures:/figures \
-  -v /tmp:/tmp \
   $image
+
+#  -v /tmp:/tmp \

### 2019-07-20 code/adjust/all.sh
Uncommented things and verified functionality

@@ -3,5 +3,5 @@
 set -e
 
 bash ./scale.sh
-#bash ./combat.sh
-#bash ./confounded.sh
+bash ./combat.sh
+bash ./confounded.sh

### 2019-07-20 code/adjust/combat.sh

@@ -4,7 +4,9 @@ set -e
 
 printf "\033[0;32mAdjusting the data with ComBat\033[0m\n"
 
-#Rscript adjust.R "/data/input/mnist/noisy.csv" -b Batch
-#Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -b batch
-#Rscript adjust.R "/data/input/gse37199/gse37199.csv" -b plate
+Rscript adjust.R "/data/input/mnist/noisy.csv" -b Batch
+Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -b batch
+Rscript adjust.R "/data/input/gse37199/gse37199.csv" -b plate
 Rscript adjust.R "/data/input/tcga/tcga.csv" -b CancerType
+Rscript adjust.R "/data/input/tcga_medium/tcga_medium.csv" -b CancerType
+Rscript adjust.R "/data/input/tcga_small/tcga_small.csv" -b CancerType

### 2019-07-20 code/adjust/confounded.sh

@@ -5,9 +5,10 @@ set -e
 printf "\033[0;32mAdjusting the data with Confounded\033[0m\n"
 
 # We also save a log for our loss chart
-#confounded /data/input/mnist/noisy.csv -b Batch --learning-rate 0.0001 -f /data/metrics/mnist_confounded_log.csv
-##confounded /data/input/mnist/noisy.csv -b Batch --learning-rate 0.001  -f /data/metrics/mnist_confounded_log2.csv
+confounded /data/input/mnist/noisy.csv -b Batch --learning-rate 0.0001 -f /data/metrics/mnist_confounded_log.csv
+confounded /data/input/mnist/noisy.csv -b Batch --learning-rate 0.001  -f /data/metrics/mnist_confounded_log2.csv
 confounded /data/input/bladderbatch/bladderbatch.csv -b batch
-#/confounded/log
-#confounded /data/input/gse37199/gse37199.csv -b plate
-#confounded /data/input/tcga/tcga.csv -b CancerType
+confounded /data/input/gse37199/gse37199.csv -b plate
+confounded /data/input/tcga/tcga.csv -b CancerType
+confounded /data/input/tcga_medium/tcga_medium.csv -b CancerType
+confounded /data/input/tcga_small/tcga_small.csv -b CancerType

### 2019-07-20 code/adjust/scale.sh

@@ -4,7 +4,9 @@ set -e
 
 printf "\033[0;32mAdjusting the data with a linear scaler\033[0m\n"
 
-#Rscript adjust.R "/data/input/mnist/noisy.csv" -a scale -b Batch
-#Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
-#Rscript adjust.R "/data/input/gse37199/gse37199.csv" -a scale -b plate
+Rscript adjust.R "/data/input/mnist/noisy.csv" -a scale -b Batch
+Rscript adjust.R "/data/input/bladderbatch/bladderbatch.csv" -a scale -b batch
+Rscript adjust.R "/data/input/gse37199/gse37199.csv" -a scale -b plate
 Rscript adjust.R "/data/input/tcga/tcga.csv" -a scale -b CancerType
+Rscript adjust.R "/data/input/tcga_medium/tcga_medium.csv" -a scale -b CancerType
+Rscript adjust.R "/data/input/tcga_small/tcga_small.csv" -a scale -b CancerType

### 2019-07-20 code/all.sh

@@ -2,17 +2,17 @@
 
 set -e
 
-#cd prepdata
-#bash all.sh
-#cd ../
+cd prepdata
+bash all.sh
+cd ../
 
 cd adjust
 bash all.sh
 cd ../
 
-#cd metrics
-#bash all.sh
-#cd ../
+cd metrics
+bash all.sh
+cd ../
 
 #cd figures
 #bash all.sh

### 2019-07-20 code/metrics/classify.py

@@ -98,7 +98,7 @@ if __name__ == "__main__":
     cache = DataFrameCache()
     # comparisons = cache.get_dataframe(COMPARISONS_PATH)
     LEARNERS = [
-        (RandomForestClassifier, {"n_estimators": 10}),
+        (RandomForestClassifier, {"n_estimators": 100}),
         # (MLPClassifier, {"hidden_layer_sizes": tuple([5]*10), "max_iter": 1000}),
         (GaussianNB, {}),
         # (KNeighborsClassifier, {}),

### 2019-07-20 code/prepdata/tcga_sampling.R

@@ -1,10 +1,10 @@
-if (!require("pacman")) install.packages("pacman"); library(pacman)
-p_load(tidyverse)
+library(readr)
+library(dplyr)
 
-IN_FILE <- "../data/input/tcga/tcga.csv"
-OUT_DIR <- "../data/input/"
-MEDIUM <- paste0(OUT_DIR, "/tcga_medium/medium.csv")
-SMALL <- paste0(OUT_DIR, "/tcga_small/small.csv")
+IN_FILE <- "/data/input/tcga/tcga.csv"
+OUT_DIR <- "/data/input/"
+MEDIUM <- paste0(OUT_DIR, "/tcga_medium/tcga_medium.csv")
+SMALL <- paste0(OUT_DIR, "/tcga_small/tcga_small.csv")
 
 get_group <- function(df, colname, value) {
   colname <- enquo(colname)

### 2021-04-06 manuscript.tex
edits to manuscript including results

@@
 	We created *Confounded* to remove these effects from expression data.
-	[@Please modify the following sentence to better reflect the dual-objective aspect and the fact that it uses neural networks.@]
+	[@Please modify the following sentence to better reflect the dual-objective aspect and the fact that it uses neural networks.@] % duel objective meaning remove the effects and effectively replicate the origional expression levels/data?
 	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
@@
 	Our software is available at \url{https://github.com/jdayton3/Confounded}.
+	%I'm not sure if we need to put my repository on there since it's different than jonathan's now
 \end{abstract}
@@
 Expression datasets are quite ``wide,'' typically containing tens of thousands of columns representing genes in the human transcriptome.
-Because of the sensitive nature of technologies used to generate these data (i.e., expression levels may respond drastically to small environmental changes), expression data often contains unwanted confounding effects that can skew analyses.
-Three examples of these include batch effects \label{it:batch}, known covariates, dataset-integration effects.
+Because of the sensitive nature of technologies used to generate these data (i.e., expression levels may respond drastically to small environmental changes), expression data often contain unwanted confounding effects that can skew analyses.
+% removed an s from contains
+Three examples of these include batch effects \label{it:batch}, known covariates, and dataset-integration effects. %included and
+%kind of seems like we say for example/ and example many times.
 
@@
 
-Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions; the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
+Artificial neural networks, a category of machine-learning algorithms, are inspired by the way human brains function; input values pass through layers of linear and nonlinear functions; the final output values are measured against specific objectives, and the function layers are adjusted to bring the outputs closer to the objectives; this process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.%There are a lot of semi colons here.
+Should we break up the sentence a little bit or reword?
 Neural networks have been applied broadly to gene-expression data.
@@
 
-Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
+Progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
+% Was: Recent progress has been made ....
 However, these methods are designed for a limited range of scenarios:
@@
 Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remain after adjustment.
+% What is the difference between the first list and the second? Should we be consistant?
 
@@
 Either way, we might need a reference.
--You would know better than I.
+-You would know better than me.
 Please clarify this based on what is relevant to the paper and add a relevant reference.@]
@@
 thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.
-Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions:
+Researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: % was: Recently, researchers ....
 1) to maximize similarity between in the input and output data and
@@
 
+%% ToDo: This is where we would have to change the default values in the confounded :
+%% for DR Piccolo to Change : 
+%%
+%%
 We implemented the variational autoencoder architecture \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
-This network has 2 hidden encoding layers and 2 decoding layers, each of size 500.
-The code size is 20.
+This network has 2 hidden encoding layers and 2 decoding layers, each of size 500. % Is this sample size? I'm not sure where the 500 comes from
+The code size is 200. %was: 20.
 Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015} and trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).
@@
 \subsubsection{Training}
-
-In all cases, we trained the network using the Adam Optimizer \citep{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
+%% also another place to change the default values
+In all cases, we trained the network using the Adam Optimizer \citep{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 256. % was: 100.
 In each iteration, we optimized on both $L_D$ and $L_{dual}$.
@@
 For each dataset, training took roughly 30 minutes to complete, including time to load the input into memory and save the output to disk.
+% We should probably re-time it to make sure this is still accurate
 
@@
 
-We tested each method on four datasets that varied in size and the ype of data measured.
+We tested each method on four datasets that varied in size and the type of data measured.
 
@@
 \end{table}
+
+%% I keep forgetting whether or not we are leaving the MNIST dataset and just not doing anything, or if we are deleting it out. 
+%% maybe put in the simulated expression data and information here
+
+%\subsection{Simulated Expression}
+%We simulated a balanced two class two batch dataset that contains 200 samples of 1000 features each that allowed us to simulate a large dataset to optimize the parameters of the confounded method.
+%Initially, we used make_classification from the sklearn classification library \url{}, but combat always outperformed confounded because the clustering made by the make_classification algorithm made it possible for linear models to accurately classify the true classes.
+%As we moved away from clustering and added nonlinear patterns to a randomly generated normal distribution set, we saw confounded begin to perform better than combat. 
+%When the true class and batch signals are nonlinear and small, combat is not as able to acuratly classify the classes or remove the batches, but confounded is able to identify the classes and effectively remove the batches.
 
@@
 
-In a recent paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
+In an earlier paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
+%WAS: In a recent paper, \citet{shaham_removal_2017} used neural networks to remove batch effects.
 Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
@@
 %TODO: Update the algorithms listed here.
-We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
+We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
+%We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
 In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
-[@The defaults in version 0.19.1 of sklearn are n\_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]
+[@The defaults in version 0.19.1 of sklearn are n\_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@] %we used those parameters
 
-We calculated the average classification accuracy across four cross-validation folds, repeated three times.
+We calculated the average classification accuracy across three cross-validation folds, repeated ten times.
+%We calculated the average classification accuracy across four cross-validation folds, repeated three times.
 We interpreted lower accuracy for batch classification to indicate that batch signal was removed more effectively.
@@
 With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
-Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
-This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
-Although Naive Bayes is no longer able to identify confounding effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
+%Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
+%This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
+%Although Naive Bayes is no longer able to identify confounding effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
 In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
@@
 	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
-	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
+	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 3-fold cross-validation repeated 10 times for several classifiers.
 	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
@@
 On the MNIST dataset, Random Forests is able to detect batch with perfect or near-perfect accuracy after adjustment with the scale adjuster and ComBat, but the highest batch classification accuracy after adjustment by Confounded is Naive Bayes, with an accuracy of 68.8\%.
-With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 87.6\% to 96.3\% and 97.1\% respectively, whereas Confounded decreases the accuracy to 8.8\%.
+With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 95.0\% to 98.5\% and 99.9\% respectively, whereas Confounded decreases the accuracy to 77.8\%.
+%With TCGA, both the scale adjuster and ComBat drastically increase Random Forests' batch classification accuracy from 87.6\% to 96.3\% and 97.1\% respectively, whereas Confounded decreases the accuracy to 8.8\%.
 
@@
 
-With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
+%What is the true class information for Bladderbatch?
+With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 68.7\% for ComBat and 68.8\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
+%With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat, (with Random Forests, Bladderbatch: 74.3 for ComBat\% and 72.1\% for Confounded, GSE37199: 60.4\% for ComBat and 69.0\% for Confounded; see \figurename{} \ref{fig:true_class} and Table \ref{tab:true_class}).
 For the Bladderbatch dataset, true class accuracy is much lower after adjusting with any algorithm, indicating that cancer status and batch may not be independent.
@@
 With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
-A look at the MNIST digits before and after adjustment (see \figurename{} \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
-With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs.
+%%This isn't accurate anymore. Our fine tuning made it better. 
+%%This may be a good place to include simulated expression results. 
+The Confounded's accuracy with Random Forest on our simulated dataset is still much higher than the baseline(70.5\% vs. 50.0\%). %Not sure if that's the right wording
+A look at the TCGA data before and after adjustment shows that even with larger datasets, the accuracy of Confounded never decreases below the baseline (72.6\% vs. 69.8\%). 
+% compare TCGA datasets here to show that the size of datasets affects the performance. 
+%To compare the effect that the size of the data has on Confounded in relation to the other methods, we created two sub-datasets from the TCGA data, a medium and a small and compared their performance.
+On Random Forest, the ability for machine learning to identify smaples with TP53_Mutated genes as the data size increased changed in Scaled: 71.2 \% to 71.9\% to 80.1\%, in Combat: 71.0\% to 75.2\% to 81.0\%, and Confounded: 67.9\% to 71.0\% to 72.6\%.
+%The ability for machine learning classification algorithms to detect the CancerType also increased with increase in size of the dataset.
+The Random Forest accuracy increased in each adjuster with Scaled: 7.2\% to 89.0\% to 98.5\%, in Combat: 96.3\% to 99.9\% to 99.9\%, and Confounded: 65.8\% to 66.5\% to 78.0\%.
+%Though the ability for Confounded to replicate the origional expression data is a little lower than that of the other adjusters, the ability for the Confounded to remove the patterns that allow machine learning to detect CancerType is sgnificantly better than that of the other adjusters which have very high, and sometimes higher accuracy after adjustment than before any kind of adjustment is made.
+See table?
+%A look at the MNIST digits before and after adjustment (see \figurename{} \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
+%With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs.
 11.4\%), but with TCGA, the accuracy decreases below baseline (66.5\% vs.
 69.8\%) while the other adjusters' accuracies remain above baseline.
-However, the particular set of parameters that we used in Confounded are likely not optimal for every dataset.
+%However, the particular set of parameters that we used in Confounded are likely not optimal for every dataset.
 Additional tuning may improve the performance metrics.
@@
 Therefore, some form of nonlinear adjustment must be used in order to correct for real-world confounding effects.
+%It may just be me, but this senstence is confusing because of the length and the number of commas
 Rather than individually model each of infinitely many possible nonlinear interactions to see which represent confounders, we can use a neural network to both approximate and remove the confounders, since neural networks are proven to be universal function approximators \citep{csaji_approximation_2001}.
@@
 	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in \figurename{} \ref{fig:mnist}).
-	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
+	However, work in 2016 has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}
+%	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
 	Similar research with variational autoencoders and gene expression data may yield improved reconstruction losses and decrease the blurring effect.
 	\item Confounded takes a long time to run in comparison with ComBat and other linear adjusters.
-	Although we acknowledge this as a downfall of many types of machine learning and of neural network in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
+	Although we acknowledge this as a downfall of many types of machine learning and of neural networks in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
 	\item It can be difficult to identify the optimal network structure and parameter set for a neural network.
 	Though this is the case for many applications of neural networks, we feel that Confounded's default structure worked well in our testing and that it will suffice for most batch correction applications.
+	% I wrote that in simulated expression that we optimized the parameters, so we could say that we optimized parameters that worked well across several different datasets and should be good for most applications, but can further be optimized to perform better on individual datasets.
 	\item Neural networks usually perform better when given large amounts of data, and traditional batch datasets typically have very few samples.
